### 2023-02-28 1_Version_Paper/Networksecurity.tex
added paper to source controll

%%%%%%%%%%%%%

\documentclass[10pt,a4paper,titlepage]{paper}
\setcounter{secnumdepth}{5}
\usepackage[a4paper, left=2cm, right=2cm, top=1.6cm, bottom=2.5cm]{geometry}
\usepackage[utf8]{inputenc}
\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% sub sub sub section :) mit paragraph{text}
\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
            {-2.5ex\@plus -1ex \@minus -.25ex}%
            {1.25ex \@plus .25ex}%
            {\normalfont\normalsize\bfseries}}
\makeatother
\setcounter{secnumdepth}{4} % how many sectioning levels to assign numbers to
\setcounter{tocdepth}{4}    % how many sectioning levels to show in ToC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{pst-all}
\usepackage{pst-circ}
\usepackage{amsmath}
\usepackage[ngerman]{babel}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{color}
\usepackage{framed}
\usepackage{ntheorem}
\usepackage{epsfig}
%\usepackage[pdftex]{graphicx}
\usepackage{pstricks}
\usepackage{placeins}
\usepackage{caption}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{fancyvrb,newverbs,xcolor}
\usepackage{lipsum}% just for this example
\usepackage{pdfpages}
\usepackage{fourier} 
\usepackage{array}
\usepackage{makecell}
%für echte hochkommas statt einfachen anführungszeichen
\usepackage{upquote}
\renewcommand\theadalign{bc}
\renewcommand\theadfont{\bfseries}
\renewcommand\theadgape{\Gape[4pt]}
\renewcommand\cellgape{\Gape[4pt]}

\definecolor{cverbbg}{rgb}{0.9, 0.95, 1.0}

\newenvironment{cverbatim}
 {\SaveVerbatim{cverb}}
 {\endSaveVerbatim
  \flushleft\fboxrule=0pt\fboxsep=.5em
  \colorbox{cverbbg}{\BUseVerbatim{cverb}}%
  \endflushleft
}

\definecolor{pverbbg}{rgb}{0.97, 0.91, 0.81}

\newenvironment{pverbatim}
 {\SaveVerbatim{cverb}}
 {\endSaveVerbatim
  \flushleft\fboxrule=0pt\fboxsep=.5em
  \colorbox{pverbbg}{\BUseVerbatim{cverb}}%
  \endflushleft
}

\pagestyle{fancy}
\lhead{Cyber-security - Passwords: Version 0.1}
\rhead{Team HTL Anichstraße (February 2023)}

% Befehl für farbige Rahmen
\definecolor{rotlich}{rgb}{1,.9,.9}
\definecolor{gruenlich}{rgb}{.9,1,.9}
\definecolor{blaulich}{rgb}{.9,.9,1}

\begin{document}

\pagenumbering{roman}
%\pagestyle{plain}
\begin{center}
		\vspace*{1cm}

		\Huge \bf Passwords\\
		\vspace{5cm}

		%\includegraphics[scale=.35]{./images/title.png}\\

		\vspace{1.5cm}
		\huge A brief summary of solutions to the never dying problems of pins. \\
		\vspace{2cm}
		\Large Basierend auf dem Buch \textit{Bitcoin ohne Vorkenntnisse} von Benjamin Spahic,\\ sowie diversen Online-Artikeln.
		\vspace*{1.5cm}

		\Large Version 0.1\\
		\large Febuary 2023\\
		\large Team HTL Anichstraße\\[30pt]
		\vspace{1cm}

		\vspace*{5cm}

        \large no \copyright\ by Lilo Zobl\\

\end{center}
\newpage    %Ende der Titelseite

% Inhaltsverzeichnis
\addtocontents{toc}{\protect\thispagestyle{empty}} 
\tableofcontents
\newpage
%Abbildungsverzeichnis erstellen
\listoffigures
\listoftables
\clearpage
% ab sofort wieder mit arabischen Seitenzahlen
\pagenumbering{arabic}

\section{Abstract}

\section{Introduction}
Password. Everyone has one or rather multiple. But it's a faulty construction. Nowadays passwords are everything but a good invention.\\ These are the reasons: i) Passwords are not user-friendly. In order to make passwords securer providers make them complexer. User have to write them down or use the same one over and over so they don't forget it. ii) Passwords are not secure because they can be shared and guessed. Of course a lot of passwords are also stolen, which is a lot easier than anyone would think. iii) Passwords are weak and get reused. Here are some examples to the most frequently used ones of 2022: "password", "123456", "123456789", "guest", "qwerty", ect.\\
Luckily there is a websites that lets users check if their account information was breached (that's when your email connected to a password is leaked; always a huge amount of accounts get breached from a branch such as Facebook). The website is called "haveibeenpwned.com". HIBP is open source and secure. If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number). It only ever retrieves the data from storage then returns it in the response. 
%ich würd gern ein bild von der Seite einfügen hier

\section{Haupteil?}
Passwords have always been a security threat ever since they where introduced. Providers have tried to find solutions such as requiring users to make the pin at least eight characters long or use upper case and lower case letters. Sometimes they require you to include digits or symbols. However those "solutions" didn't improve the situation overall and sometimes made it worse.
To eliminate this security threat a few solutions created.

\subsection{Password Save}
\subsubsection{Explanation}
A password save allows you to save a list of your usenames linked to the password. It has one master password that looks the save. Instead of remembering a ton of password at once users only have to recall one single pin. The rest is done by the app or rather the software.
\subsubsection{Providers}
\begin{itemize}
\item \textbf{KeepassXC}\\
This software is developed for users with extremely high demands of secure personal data management. \\
KeePassXC uses Advanced Encryption Standard (AES) encryption algorithm with a 256-bit key to secure the password database.\\
The biggest difference to other password saves is that the data (password, account information and additional data such as URLs, attachments ans notes) is stored in an offline, encrypted file that can be stored locally. This prevents your data from getting leaked when the could gets hacked or breached.\\
The program is customizable. It allows its user to customize literally everything to their needs.\\

\item \textbf{1Passwort}\\
This password save is known for its easy to use interface and high security encryption.\\
1Password uses an uncommon encryption known as dual-key encryption. If the server gets breached it's impossible for the hacker to decrypt the users sensitive information because of its two keys. The first key is the users master-key. The second key is a secret key, which is a 128-bit, machine-generated code. The secret key is generated on every device you log into. It will only be saved on your devices and never saved with your other pins.\\
Depending on what account type you choose your data is stored differently. However in only one version users have the option to save their data locally. In every other option the user's data gets saved in a cloud-based vault.\\
The interface is very user-friendly. It even creates strong passwords for accounts that are newly created.
\item \textbf{Password Safe (MATESO)}\\

\item \textbf{LastPass}\\
\item \textbf{Dashlane}\\
\end{itemize}

\subsection{2-factor authentication}
\subsection{The future of passwords}

\section{conclusion}

\section{Grundlagen der Kryptographie}
Durch bekannte, erprobte Verschlüsselungsalgorithmen werden Transaktionen
verschlüsselt, durchgeführt und bestätigt.
\begin{center}
\textit{„Die Verschlüsselung funktioniert. Richtig implementierte starke
Kryptosysteme sind eines der wenigen Dinge, auf die wir uns verlassen können“. - Edward Snowden}
\end{center}
\subsection{Die drei Ziele der Kryptographie}
\begin{itemize}
\item \textbf{Integrity (Datenintegrität)\\ Wie können wir sicher sein, dass Daten unverfälscht sind?}\\ Mit sogenannten Hashing-Funktionen (für mehr Details siehe Kapitel \ref{hash} auf Seite \pageref{hash})\\ Aktueller Standard: Secure Hash Algorithm 2 (SHA2); Beispiel: Wird auch oft bei Downloads angeboten um zu überprüfen, ob das heruntergeladene File mit dem am Server übereinstimmt. 
\item \textbf{Authentication (Authentifizierung)\\ Wie können wir sicher sein, dass die Daten auch tatsächlich vom vermutlichen Absender stammen?}\\Mit keyed Hashes (siehe Kapitel \ref{hmac} auf Seite \pageref{hmac}) oder asymmetrischer Verschlüsselung (siehe Kapitel \ref{asym} auf Seite \pageref{asym}). Beispiel Algorithmen: HMAC-MD5, HMAC-SHA-1, RSA und DSA;
\item \textbf{Confidentiality (Vertraulichkeit)\\ Wie können wir sicher sein, dass die Daten von niemand anderem gelesen werden können?}\\Mit klassischer Verschlüsselung (symmetrisch oder asymmetrisch). Beispiel Algorithmen: DES, 3DES, AES, SEAL, Elliptic-Curve;
\end{itemize}

\newpage
\subsection{Datenintegrität - Hashing} \label{hash}
\textbf{Hashes} werden verwendet um die \textbf{Integrität} einer Nachricht oder eines Files sicherzustellen. Hashing basiert auf einer \textbf{mathematischen Einweg-Funktion} - einfach durchzuführen aber schwer umzudrehen, wie zum Beispiel das Kaffee mahlen: Es ist einfach aus Kaffeebohnen Kaffeepulver zu mahlen, aber aus dem Pulver wieder die Bohnen zusammensetzen ist eher schwierig ;-)\\
\begin{minipage}{6.7cm}
\textbf{Prinzip}:\\Aus einer beliebig langen Nachricht oder einem File-Inhalt wird durch hashing ein Hash-Value mit einer fixen Länge erzeugt.
Beim Hashing wird der ursprüngliche Inhalt in mehreren Stufen verschoben, vertauscht, komprimiert oder ergänzt. Wie genau ist immer vom verwendeten Hashing-Algorithmus abhängig.\\
Am Ende dieser Prozedur entsteht ein immer gleich langer (bei SHA256 z.B: 256 bit) Hash-Wert.\\ Jede noch so kleine Änderung in der Nachricht oder dem File erzeugt ein unterschiedliches Hash-Value. Hashes werden deshalb auch oft als digitale Fingerabdrücke bezeichnet. 
\end{minipage}
\begin{minipage}{8cm}
\begin{center}

%\includegraphics[scale=0.35]{./images/hash.eps}
\captionof{figure}[Hashing Function - die Abbildung stammt vom Cisco Netacad Network Security Kurs]{Hashing Function}

\end{center}
\end{minipage}
\subsection{SHA}
Hashing Algorithmen werden laufend weiterentwickelt. Der aktuell am weitesten verbreitete Algorithmus ist \textbf{SHA}.\\ \textbf{SHA steht für  Secure Hash Algorithm}, und wurde vom National Institute of Standards and Technology (NIST) entwickelt; Erste Version stammt von 1994, aktuelle Versionen wurden im August 2002 (SHA-2-Familie) vorgestellt und sind SHA-256, SHA-384, SHA-512. Die nächste Generation (SHA-3-Familie) wurde 2015 vorgestellt.\vspace{.3cm}\\
\textbf{SHA-256} hat auch für die Funktionsweise von Bitcoin eine große Bedeutung. Mehr dazu sie Kapitel \ref{mining} auf Seite \pageref{mining}.
\vspace{.5cm}\\
File-Hash berechnen unter Linux - Darstellung in hexadezimaler Schreibweise!
\begin{cverbatim}
sha256sum ./main.jpg 
a7f837e28843fba098c0929037743d28edf6a97ed74ff17127aa329423acf52c  ./main.jpg
\end{cverbatim}
File-Hash berechnen unter Windows in der Power-Shell
\begin{cverbatim}
Get-FileHash C:\Users\user1\Downloads\file.txt -Algorithm SHA256 | Format-List

Algorithm : SHA256
Hash      : 3CBCFDDEC145E3382D592266BE193E5BE53443138EE6AB6CA09FF20DF609E268
Path      : C:\Users\user1\Downloads\file.txt
\end{cverbatim}
Für mehr Informationen zur genauen Funktionsweise von SHA-256 siehe:\\
\href{https://en.wikipedia.org/wiki/SHA-2}{\color{blue}https://en.wikipedia.org/wiki/SHA-2}

\newpage
\subsection{HMAC} \label{hmac}
Hash Message Authentication Code (HMAC) oder auch manchmal Keyed-hash message authentication code (KHMAC) sind Hashes die noch einen zusätzlichen, geheimen Schlüssel verwenden. 
\begin{center}
%\includegraphics[scale=0.4]{./images/hmac.png}
\captionof{figure}[HMAC - die Abbildung stammt vom Cisco Netacad Network Security Kurs]{HMAC - hash bases message authentication}
\end{center}
Nur wer die Daten und den geheimen Schlüssel hat kann den richtigen Hash-Wert berechnen.

\subsection{Verschlüsselung - Vertraulichkeit}
Es gibt verschiedene Verschlüsselungsmethoden. Prinzipiell lässt sich eine Einteilung in symmetrische und asymmetrische Verfahren treffen.

\subsubsection{Symmetrische Verschlüsselung}
Symmetrisch bedeutet, dass sowohl für die Verschlüsselung als auch für die Entschlüsselung \textbf{der gleiche Schlüssel} verwendet wird.\\
\begin{center}
%\includegraphics[scale=0.43]{./images/symmetric-encryption.png}
\captionof{figure}[Symmetric-Encryption - die Abbildung stammt vom https://www.cheapsslshop.com/blog/symmetric-vs-asymmetric-encryption-whats-the-difference]{Symmetrische Verschlüsselung}
\end{center}
Symmetrische Verfahren sind meist \textbf{schnell} (schneller als asymmetrische Verfahren) und verwenden im Vergleich zu asymmetrischen Verfahren eher kurz Schlüssel.\\
Die eigentliche Übersetzung kann blockweise erfolgen (z.B. es werden jeweils 128 Bit der Nachricht übersetzt) - dann spricht man von einem \textbf{Block Cipher}, oder Bit by Bit - dann spricht man von \textbf{Stream Chiper}.\vspace{.3cm}\\
\textbf{AES} (Advanced Encryption Standard) ist zur Zeit (seit dem Jahr 2000) der etablierte Standard für symmetrische Verschlüsselung. Es ist ein Blockcipher mit 128 Bit Länge und die Schlüssellänge kann 128, 192 oder 256 Bit betragen.\\
Für mehr Info zu AES siehe:\\
\href{https://de.wikipedia.org/wiki/Advanced_Encryption_Standard}{\color{blue}https://de.wikipedia.org/wiki/Advanced\_Encryption\_Standard}\vspace{.3cm}\\
Das eigentliche \textbf{Problem} bei symmetrischen Verfahren liegt im \textbf{Schlüsselaustausch}! Wie schafft man es über ein unsicheres Medium einen geheimen Schlüssel auszutauschen?\\
Aber auch dafür gibt es Lösungen, z.B. den Diffie-Hellmann Key Exchange, aber dies sprengt den Rahmen dieses Skripts. Mehr dazu gibt es im Fach Netzwerktechnik!
\subsubsection{Asymmetrische Verschlüsselung} \label{asym}
Wird auch \textbf{Public Key Encryption} genannt.\\
Die asymmetrische Verschlüsselung umgeht das Problem des Key-Exchanges in dem es pro Teilnehmer \textbf{zwei Schlüssel} gibt: einen Public Key und einen Private Key. Die Besonderheit an den Keys ist, dass eine Message, die mit einem der Keys verschlüsselt wurde, nur mit dem anderen Key entschlüsselt werden kann! Der öffentliche Schlüssel und der private Schlüssel hängen über mathematische Funktionen zusammen. Es ist jedoch nicht möglich, von dem öffentlichen Schlüssel auf den privaten Schlüssel zurück zu schließen.
\begin{center}
%\includegraphics[scale=0.43]{./images/asymmetric-encryption.png}
\captionof{figure}[Asymmetric-Encryption - die Abbildung stammt vom https://www.cheapsslshop.com/blog/symmetric-vs-asymmetric-encryption-whats-the-difference]{Asymmetrische Verschlüsselung}
\end{center}
Es gibt mehrere asymmetrische Verschlüsselungsverfahren z.B. RSA oder ECC. Sie sind in der Regel deutlich langsamer als symmetrische Verfahren, umgehen aber das Problem des Schlüsselaustauschs und erlauben zusätzlich zur Confidentiality auch noch Integrity und Authentication!
\begin{itemize}
\item \textbf{Public Key (Encrypt) + Private Key (Decrypt) = Confidentiality}\\Schickst Du mir etwas mit meinem Public Key, kann nur ich es mit meinem Private Key lesen!
\item \textbf{Private Key (Encrypt) + Public Key (Decrypt) = Authentication}\\Schicke ich dir etwas mit meinem Private Key, kann es nur von mir gekommen sein, weil nur ich den Private Key habe - lesen kann es jeder, der meinen Public Key hat. Dies nennt man auch eine \textbf{signierte} Nachricht oder \textbf{digitale Signatur}!
\item \textbf{Confidentiality, Authentication und Integrity?} - geht auch:\\
Ich schicke dir eine Nachricht mit deinem Public Key, aber zusätzlich bilde ich noch einen Hash der Nachricht mit meinem Private Key! Du entschlüsselst die Nachricht mit deinem Private Key und überprüfst den Hash mit meinem Public Key - damit ist sicher gestellt, dass die Nachricht von mir ist und nicht verändert wurde (wenn der Hash übereinstimmt).
\end{itemize}

\paragraph{ECC - Eliptic Curve Cryptography} \label{ECC}
Bitcoin verwendet ECC um aus einem Private Key (256 Bit Länge - meist als Zufallszahl erzeugt) einen Public Key zu erzeugen.\\
Achtung: Es gibt eine technische Einschränkung: Der Private Key muss zwischen \verb|0| und \verb|n-1| liegen wobei \verb|n=1,1578*10^77| ist und mit der elliptischen Kurve zu tun hat (Order der Kurve) die Bitcoin verwendet.\\
Der Public Key wird durch \textbf{Elliptische Kurvenmultiplikation} mit der Formel $K=k*G$ erzeugt. Wobei $k$ der Private Key ist. $G$ der \textit{generator point} - ein Punkt auf der elliptischen Kurve und $K$ der resultierende Public Key. Achtung: Der Operator $*$ ist keine normale Multiplikation, sondern eine elliptische Kurvenmultiplikation!\\
Die Besonderheit dabei ist, das $K$ aus $k$ und $G$ relativ einfach berechnet werden kann, aber die Umkehrfunktion, nämlich $k$ aus $K$ und $G$ zu berechnen, nahezu unmöglich ist.\\
Bitcoin verwendet eine elliptische Kurve aus dem \verb|secp256k1| Standard, der vom amerikanischen NIST (National Institute of Standards and Technology) vorgestellt wurde.\\
\subparagraph{Elliptische Kurven Multiplikation}
Im folgenden wird eine vereinfachte Kurve verwendet, um das Prinzip der elliptischen Kurven Multiplikation zu erklären - sie entspricht nicht der von Bitcoin verwendeten Kurve!\\
\textbf{ECC-Multiplikation} bedeutet, dass man die Tangente (differenzieren - erste Ableitung) zum Ausgangspunkt $G$ nimmt und schaut, wo sich diese mit der Kurve kreuzt ($-2G$). Dann springt man auf die andere Seite der Kurve und erhält $2G$. Weitere Multiplikation ergibt $4G$, $8G$ usw.
\begin{center}
%\includegraphics[scale=0.3]{./images/curve.png}
\captionof{figure}[ECC-Multiplication - die Abbildung stammt vom https://www.oreilly.com/library/view/mastering-bitcoin-2nd/9781491954379/ch04.html]{ECC-Multiplikation veranschaulicht}
\end{center}
Der Private Key gibt also an, wie oft ECC-Multiplikationen durchgeführt werden müssen, um den Public Key zu erhalten.
z.B.: Private Key = 2 führt zu Public Key = 4G, Private Key = 3 führt zu Public Key 8G, usw.
%\newpage
\section{Die Entstehung von Bitcoin}
Nach der Finanzkrise 2008 hat eine Person oder ein Kollektiv, unter dem Pseudonym \textbf{Satoshi Nakamoto}, eine digitale, dezentrale und begrenzte Währung erschaffen: den Bitcoin. Es ist bis heute ungeklärt, wer Satoshi Nakamoto wirklich ist.\vspace{.3cm}\\
Link zum ursprünglichen White-Paper:\\
\href{https://bitcoin.org/bitcoin.pdf}{\color{blue}https://bitcoin.org/bitcoin.pdf}\vspace{.3cm}\\
Link zum Bitcoin-Sourcecode auf Github:\\
\href{https://github.com/bitcoin/}{\color{blue}https://github.com/bitcoin/} \vspace{.3cm}\\
Die Grundlage der Digitalen Währung sind:
\begin{itemize}
\item \textbf{Transparenz}\\alle Transaktionen werden in einer Blockchain aufgezeichnet und sind seit der ersten Transaktion einsehbar.
\item \textbf{Sicherheit}\\die Blockchain ist im Nachhinein nicht mehr (nur sehr schwer) änderbar.
\item \textbf{Dezentralisierung}\\die Kopien der Blockchain liegen verteilt im Netz (Full-Nodes) und werden regelmäßig aktualisiert.
\item \textbf{Inflationssicherheit}\\es gibt nur eine begrenzte Anzahl an Bitcoins (21 Millionen)  
\end{itemize}

\subsection{Bitcoin, mBTC und Satoshi - die Währungseinheiten}
Ähnlich wie es beim Euro noch Euro-Cents gibt, gibt es auch bei Bitcoin (\textbf{BTC}) kleinere Einheiten: Den Milli-Bitcoin (\textbf{mBTC}) und den \textbf{Satoshii}. Es gilt:
\begin{center}
\verb|1 BTC = 1000 mBTC = 100.000.000 Satoshii|
\end{center}

%\newpage
\section{Blockchain}
Im folgenden wird der Aufbau einer Blockchain am Beispiel von Bitcoin erklärt, das Prinzip kann aber auf beliebige Daten (nicht nur Bitcon-Transaktionen) angewandt werden.
\subsection{Aufbau der Blockchain}
Die Blockchain besteht aus Blöcken. Der \textbf{erste Eintrag} eines Blocks ist immer ein \textbf{Verweis} auf den \textbf{vorherigen Block}. Dadurch ist die Verbindung zum vorherigen Block festgelegt und die Blöcke verknüpfen sich zu einer Kette.\\
Bei Bitcoin enthält ein Block eine Reihe von Transaktionen die in einem bestimmten Zeitraum angefallen sind.\\
Die Blockgröße ist bei Bitcoin ursprünglich auf 1 MByte limitiert (ist inzwischen auf 4MByte angehoben, aber die durchschnittliche Blockgröße ist bei ca. 1,13 MByte). Dadurch können pro Block ca. 2000 Transaktionen durchgeführt werden.
Über die gesamten Daten eines Blocks wird mittels SHA-256 ein Hash gebildet.
\begin{center}
%\includegraphics[scale=0.4]{./images/block.png}
\captionof{figure}[Abbildung 22 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Hash-Generierung aus den Daten eines Blocks}
\end{center}
Ändern wir nur ein Zeichen des gesamten Blocks, ändert sich der Hash
vollständig und unvorhersehbar. Eine einzige Änderung in nur einer Transaktion würde den kompletten Hash des Blocks und damit alle weiteren Blöcke verändern.\\
Somit ist der Hash des aktuellen Blocks gleichzeitig ein eindeutiger Verweis auf diesen Block. Der Hash des aktuellen Blocks wird als erster Eintrag des Folgeblocks eingesetzt. So entsteht eine Reihenfolge von Blöcken - die Blockchain.
\begin{center}
%\includegraphics[scale=0.4]{./images/blocks.png}
\captionof{figure}[Abbildung 23 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Aneinanderreihung von Blöcken}
\end{center}
Der erste Block in der Blockchain ist vorgegeben und wird \textbf{Genesisblock} genannt. 
\subsection{Dezentralisierung, Full-Nodes und Größe}
\textbf{Dezentralität} bedeutet, dass es keine zentrale Institution wie  z.B. eine Bank oder einen Staat gibt, der über ein System (die Blockchain) wacht. Stattdessen liegen viele Kopien der Blockchain verteilt in einem großen Netzwerk. Jeder Teilnehmer des Netzwerks, welcher die \textbf{volle Transaktionshistorie} (also die gesamte Blockchain) hostet, wird \textbf{Full-Node} genannt.\\
Damit jeder Knotenpunkt des Netzwerks (Full-Nodes) die aktuelle Blockchain besitzt, wird die Blockchain in regelmäßigen Abschnitten aktualisiert und an jeden Teilnehmer übermittelt.
Nach einem gewissen zeitlichen Abschnitt, bei Bitcoin alle zehn Minuten, wird ein Paket mit neuen Informationen, ein sogenannter Block, an die bisherige Transaktionshistorie angehängt. Das hat den Vorteil, dass nicht jede einzelne Transaktion aktualisiert werden muss, sondern ein Bündel von Transaktionen.\\
Die zehn Minuten werden durch den Bitcoin Algorithmus gesteuert, indem die Mining-Difficulty angepasst wird. Mehr dazu siehe im Kaptiel \ref{mining} auf Seite \pageref{mining}.
Die \textbf{momentane Größe} (Oktober 2022) der Bitcoin-Blockchain liegt bei ca. 432 GByte. Zum Startzeitpunkt von Bitcoin (03.01.09) war sie 285 Byte groß. 
\begin{center}
%\includegraphics[scale=0.35]{./images/bc-size.png}
\captionof{figure}[https://www.blockchain.com/explorer/charts/blocks-size]{Wachstum der Bitcoin Blockchain seit 2009}
\end{center}

\subsection{Konsensbildung in einem dezentralisiertem System}
Wenn jeder Full-Node eine Kopie der Blockchain hat, und sie theoretisch verändern kann, wer hat dann die \textit{richtige} Blockchain?\\
Gültige Blöcke werden nur durch das rechenintensive \textbf{Mining} (siehe Kapitel \ref{mining} auf Seite \pageref{mining}) erschaffen. So vertraut jeder Bitcoin-Node der \textbf{längsten gültigen Blockchain}, da hinter dieser die meiste Rechenleistung steht und deswegen auch die Mehrheit der Teilnehmer vermutet wird. Man kann also vereinfacht sagen: \textit{Die Mehrheit bestimmt was wahr ist.} \\
Dies kann unter bestimmten Umständen problematisch werden - siehe \textit{51\% Angriff} im Kapitel \ref{51} auf Seite \pageref{51}.\\
\subsubsection{Konsensmechanismen}
Es gibt mehrere Konsensverfahren. Die zwei bekanntesten sind:
\paragraph{Proof-of-Work}
Der Zweck eines \textbf{Arbeitsnachweises} (Proof-of-work) ist es sicherzustellen, dass das Erzeugen gültiger Blöcke mit einem gewissen Aufwand verbunden ist, so dass eine nachträgliche Modifikation der Blockkette praktisch ausgeschlossen werden kann (weil es für einen einzelnen Angreifer zu rechenintensiv wäre).\\
Der Proof-of-Work besteht bei Bitcoin darin, einen SHA256 Hashwert für einen Block zu finden, der unterhalb eines bestimmten Schwellwerts (muss mit einer bestimmten Anzahl von Nullen beginnen) liegt. Der Schwellwert ist variabel um bei steigender Hash-Leitung im Netzwerk trotzdem den gleichen Zeitaufwand (ca. 10 Minuten) für die Validierung zu benötigen. Mehr Details dazu im Kapitel \ref{mining} auf Seite \pageref{mining}.
\paragraph{Proof-of-Stake}
Beim Proof-of-Stake wird eine \textbf{gewichtete Zufallsauswahl} eingesetzt. Die Wahrscheinlichkeit für die Validierung des nächsten Blocks ausgewählt zu werden, hängt vom Coin-Vermögen (dem „Stake“) und/oder der Teilnahmedauer der einzelnen Teilnehmer ab.  Es ist ein wenig mit einer Aktiengesellschaft vergleichbar - wer mehr Aktien besitzt hat mehr Stimmrecht bei Entscheidungen.\\Im Gegensatz zum, bei Bitcoin eingesetzten Proof of Work, kommt Proof of Stake ohne zeit- und energieintensives Mining aus. Zudem ist nicht möglich, das Netzwerk allein durch Besitz von Rechenleistung zu übernehmen.

\subsubsection{Fork}
Was passiert, wenn es zu Unstimmigkeiten oder Regeländerung kommt, und manche Nodes einen unterschiedlichen Block an ihre lokale Chain anhängen? In diesem Fall spricht man von einem \textbf{Fork}. Es bedeutet, dass die Blockchain im verteilten Netz nicht mehr einheitlich ist!
\begin{center}
%\includegraphics[scale=0.5]{./images/fork-01.png}
\captionof{figure}[https://www.researchgate.net/figure/The-fork-forms-and-disappears-in-the-PoW-Blockchain]{Ein Fork - die Blockchain ist nicht mehr einheitlich}
\end{center}

 Dies kann zwei Ursachen haben:
 \begin{enumerate}
 \item \textbf{Zufällig:} Wenn zwei Miner fast zeitgleich einen passend Hash für einen Block finden und diesen im Netzwerk verteilen. Aufgrund der endlichen Verteilungsgeschwindigkeit bekommen Full-Nodes unterschiedliche Blöcke, die sie aber erfolgreich verifizieren und an ihre lokale Blockchain anhängen.\\ Es gibt also gleich lange, aber unterschiedliche Chains im Netzwerk.\vspace{.4cm}\\
Diese Forks lösen sich aber meist innerhalb von wenigen (meist 5-6) weiteren Blöcken wieder auf - weil bis dahin wieder eine längste Chain existiert, von der man annimmt, dass sie die richtige Chain ist. Sie wird wieder auf alle Full-Nodes verteilt.
\begin{center}
%\includegraphics[scale=0.45]{./images/fork-02.png}
\captionof{figure}[https://www.researchgate.net/figure/The-fork-forms-and-disappears-in-the-PoW-Blockchain]{Ein Fork löst sich wieder auf - die Blockchain ist wieder einheitlich}
\end{center}

\item \textbf{Absichtlich:} Wenn es zu Regeländerungen im Sourecode kommt, z.B. wenn es Unstimmigkeit in der Community gibt. In diesem Fall spricht man von einem \textbf{Hard-Fork}. Ab diesem Zeitpunkt trennt sich das Netzwerk auf, manche Teilnehmer übernehmen die Regeländerung und manche nicht. Es entsteht dadurch eine neue Blockchain-Community wie z.B. beim Split von Ethereum und Ethereum-Classic im Juli 2016.
\end{enumerate}

\subsection{Mining} \label{mining}
Das Mining hat zwei wichtige Aufgaben. Zum einen die Verifizierung eines Blocks und zum anderen erzeugt es neue Coins.
\subsubsection{Verifizierung - Hashing}
Die Miner berechnen den Hash des aktuellen Blocks. Allerdings gibt der Bitcoin Algorithmus vor, dass der Hash mit einer bestimmten Anzahl von Nullen beginnen muss. Da der Hash nur geändert werden kann, wenn die Daten des Blocks geändert werden, wird dem Block ein weiteres Element hinzugefügt, das als \textbf{Nonce} bezeichnet wird.\\
Der Miner startet nun z.B: mit einer Nonce von 1 und berechnet die SHA256 Summe des Blocks. Wenn der resultierende Hash nicht die geforderte Anzahl von Anfangs-Nullen aufweist, erhöht er die Nonce auf 2 und berechnet wieder den Hash, usw. Es kann viele Billionen Berechnungen (Proof-of-work) brauchen, bis eine akzeptable Hashsumme gefunden wird.
\begin{center}
%\includegraphics[scale=0.3]{./images/mining.png}
\captionof{figure}[Abbildung 24 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Hashgenerierung aus den Daten eines Blocks inkl. der zusätzliche Variable \textbf{Nonce}}
\end{center}
Seit dem Bekanntwerden der Kryptowährungen gibt es Firmen die spezielle Mining (Hashing-) Geräte herstellen. Die wahrscheinlich bekannteste Firma ist \href{https://www.bitmain.com/}{\color{blue}www.bitmain.com}. Selbst kleine Geräte (Antminer) um wenige Tausend Dollar haben schon dreistellige \textbf{TH/s} (Terahashes pro Secunde) Werte.
\vspace{.3cm}\\
Wenn ein Miner nun den passenden Hash gefunden hat, schickt er den gelösten Block, inklusive seiner gefundenen \textbf{Nonce} an alle \textbf{Full-Nodes}. Diese kontrollieren, ob ein Hash mit der geforderten Anzahl an Nullen entsteht (eine einmalige SHA256 Berechnung). Ist das der Fall, wird der Block an die Blockchain angehängt und der Miner erhält seine \textbf{Belohnung}.
\subsubsection{Miner Belohnung und Transaction Fees}
Was ist die Motivation, das sehr rechen- und damit leider sehr energieintensive Mining durchzuführen? Warum macht das ein Miner, was hat er oder sie davon?\\
Es gibt zwei Verdienstmöglichkeiten:
\paragraph{Transaction Fees}
Im Bitcoin Netzwerk können nur ca. 3,3 Transaktionen pro Sekunde durchgeführt werden (Blocksize = 1 MByte -> ca. 2000 Transaktionen pro Block -> alle 10 Minuten ein neuer Block -> 2000/600 = 3,3). Wenn man nun eine dringende Transaktion machen möchte, kann man eine Transaktionsgebühr anbieten, damit die Miner die Transaktion in den aktuellen Block aufnehmen.\\
Ein Miner der einen Block löst (den richtigen Hash findet), bekommt alle Transaktionsgebühren der im Block enthaltenen Transaktionen.\\
Die anderen Miner gehen leer aus. Danach stürzen sich alle Miner sofort auf den nächsten Block und machen sich wieder auf die Suche nach der richtigen Nonce.\\
Die Transaktionsgebühren waren in der Vergangenheit starken Schwankungen unterworfen und lagen zeitweise über 60 USD pro Transaktion.
\begin{center}
%\includegraphics[scale=0.4]{./images/fees.png}
\captionof{figure}[https://www.blockchain.com/explorer/charts/fees-usd-per-transaction]{Transaktionsgebühren (Blau) und Bitcoin Marktwert (Schwarz)}
\end{center}

\paragraph{Belohnung - Block Subsidy}
Zusätzlich zu den Transaktionsgebühren
erhält der Miner, der den Block löst, eine Belohnung (neue Coins) die im Programm code verankert ist und je nach Kryptowährung unterschiedlich ist. Diese Belohnung nennt man Block Subsidy (Block-Subvention).\\
Bei Bitcoin war der Anfangswert der Block Subsidy 50 Bitcoins pro gefundenem Block. Da das Lösen eines Blocks im Mittel 10 Minuten dauert, wurden also alle 10 Minuten 50 neue Bitcoins in Umlauf gebracht.\\
Der Subsidy Wert halbiert sich allerdings alle 210 000 Blöcke (\textbf{Halving}), was einem Zeitraum von ca. 4 Jahren entspricht. Momentan (November 2022) gibt es \textit{"nur"} noch 6,25 Bitcoins pro gelöstem Block.
\begin{table}[h]
\centering
\begin{tabular}{|p{2cm}|p{2.5cm}|p{3.8cm}|p{3.7cm}|}
\hline 
\textbf{Zeitraum} &	\textbf{Block Subsidy} & \textbf{Neue Coins im Zeitraum} & \textbf{Alle Coins im Umlauf} \\
\hline 
2008-2012 & 50 & 10.500.000 BTC & 10.500.000 BTC\\
\hline
2012-2016 & 25 & 5.250.000 BTC & 15.750.000 BTC\\
\hline
2016-2020 & 12,5 & 2.625.000 BTC & 18.375.000 BTC\\
\hline
2020-2024 & 6,25 & 1.312.500 BTC & 19.687.500 BTC\\
\hline
2024-2028 & 3,125 & 656.250 BTC & 20.343.750 BTC\\
\hline
2028-2032 & 1.5625 & 328.125 BTC & 20.671.875 BTC\\
\hline 
. & . & . & .\\
\hline 
\end{tabular}
\caption{\textbf{Halving} - die Block Subsidy halbiert sich alle 4 Jahre}
\label{tabelle_1} 
\end{table}\\ 
Wie man aus Tabelle \ref{tabelle_1} auf Seite \pageref{tabelle_1} erkennen kann, nimmt die Anzahl der neu ausgeschütteten Coins ab und wird sich einen Wert von 21 Millionen BTC 
annähern.

\section{Teil der Blockchain werden}
Welche Schritte sind nun notwendig um selbst Teil der Blockchain zu werden?
\subsection{Meine Adresse auf der Blockchain}
Eine \textbf{Adresse} auf der Blockchain ist vergleichbar mit einer Konto Nummer, mit dem Unterschied, dass die Adresse öffentlich und für jeden Teilnehmer einsehbar ist! Jeder Teilnehmer kann durch Rückschau auf vergangene Transaktionen berechnen welche Adresse wie viele Bitcoins enthält!\\
Man kann z.B. auf \href{https://www.blockchain.com/btc/address/}{\color{blue}https://www.blockchain.com/btc/address/} nachsehen wie der "Kontostand" von Satoshi Nakamotos erster Adresse (hat mehrere) zur Zeit (November 2022) aussieht:
\begin{center}
%\includegraphics[scale=0.5]{./images/adress.png}
\captionof{figure}[https://www.blockchain.com/btc/address/1A1zP1eP5QGefi2DMPTfTL5SLmv7DivfNa]{"Kontostand" der Genesis Adresse - erste Adresse von Satoshi Nakamoto}
\end{center}
Die \textbf{Adresse} auf der Bitcoin Blockchain ist in Wirklichkeit der leicht abgewandelte \textbf{öffentliche Schlüssel (public key)} eines asymmetrischen Schlüsselpaares. Siehe Kapitel \ref{asym} auf Seite \pageref{asym} zur Erinnerung wie Puplic Key Encryption funktioniert.\\
Der public key wird aus einem zufällig gewählten private key (bei Bitcoin 256 Bit Länge) generiert.   
Aus diesem wird dann (mittels Hashfunktionen - zuerst SHA256 und dann RIPEMD160) die öffentliche Adresse (160 Bit = 20 Byte Länge) erstellt.
\subsection{Wallets}
Wallets sind kleine Programme die bei der Key-Erstellung und Verwaltung (eine Wallet kann mehrere Keys für mehreren Crypto-Währungen verwalten) helfen.
\begin{center}
\fcolorbox{green}{gruenlich}{\begin{minipage}{15cm}{
ACHTUNG: Die Wallets sind nicht Vergleichbar mit einer Geldtasche oder einem Konto, weil sie selber keine Coins verwalten, sondern nur die Adressen! Die Blockchain selbst ist mit ihren transparenten Transaktionen das "Konto".
}\end{minipage}}
\end{center}
\subsubsection{Arten von Wallets}
\paragraph{Hot Wallet}
Von einer Hot Wallet spricht man, wenn die Wallet online bei einer Handelsplattform (Exchange) liegt. Zum Beispiel bei \href{https://www.bitpanda.com}{\color{blue}www.bitpanda.com}.\\
Dies ist zwar bequem (Online Zugriff via Browser) aber sehr unsicher, da schon einige Exchanges gehackt worden sind oder zahlungsunfähig wurden oder der Betreiber alle Keys gestohlen hat und sich auf die Bahamas abgesetzt hat :(
\paragraph{Cold Wallet}
Die Wallet ist offline. Man unterscheidet noch zwischen:
\begin{itemize}
\item \textbf{Software Wallet} z.B. auf dem privaten PC Zuhause oder auf einem verschlüsselten USB-Stick. Ist sicher, aber Achtung wenn z.B. der USB Stick verloren geht oder der PC versehentlich entsorgt wird - siehe \href{https://futurezone.at/digital-life/festplatte-7500-bitcoin-muellhalde-finden-james-howells/402088129}{\color{blue}James Howell}, der im Jahr 2013 eine Festplatte mit Keys zu ca. 7500 BTC auf den Müll geworfen hat ;-)
\item \textbf{Hardware Wallets} sind spezielle Hardware und Software zur Key-Sicherung. Sie z.B. Firma \href{https://www.ledger.com/}{\color{blue}www.ledger.com}
\item \textbf{Paper Wallet} die Keys werden mit open source Software einmalig erzeugt und nur analog (z.B: auf einem Blatt Papier) gespeichert. Sehr sicher gegen online Angriffe aber sorgfältige Verwahrung ist ein Muss! 
\end{itemize}
\subsubsection{Key Erstellung}
Wie schon erwähnt hilft die Wallet bei der Key-Erstellung.\\
Als erstes wird der \textbf{private Key} (256 Bit) erstellt, entweder durch einen halb automatischen Algorithmus mit User Interaktion (z.B: Mouse Bewegungen) oder durch einen Zufallszahlengenerator. \\
Aus diesem wird mit Hilfe einer \textbf{elliptischen Kurven-Funktion} (siehe Kapitel \ref{ECC} auf Seite \pageref{ECC}) der zugehörige (unkomprimierte) \textbf{public Key} generiert.\\
Aus diesem wird dann (mittels Hashfunktionen - zuerst SHA256 und dann RIPEMD160) die öffentliche Adresse (160 Bit = 20 Byte Länge) erstellt.
\begin{center}
%\includegraphics[scale=0.4]{./images/key_generation.png}
\captionof{figure}[Abbildung 17 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Generierung der Bitcoin-Adresse}
\end{center}

\subsubsection{Eine Transaktion senden}
Eine weitere Aufgabe die von den Wallets erledigt wird, ist das Senden von Transaktionen.\\
Nehmen wir an, Alice möchte Bob einen Bitcoin (1 BTC) senden.\\
Die gewünschte Transaktion lautet dann: Von der öffentlichen Bitcoin-Adresse von Alice soll ein Bitcoin an die öffentliche Adresse von Bob geschickt werden.\\
Damit sichergestellt wird, dass diese Transaktion wirklich von Alice ausgeht, \textbf{signiert} sie sie mit ihrem \textbf{Private Key}. Siehe dazu Kapitel \ref{asym} auf Seite \pageref{asym}.\\
Diese signierte Transaktion (plus eine Transaktionsgebühr - siehe Kaptiel \ref{mining} Mining auf Seite \pageref{mining}) wird von der Wallet an alle umliegenden Miner und Nodes gesendet, welche sie wiederum weitersenden, bis jeder im Netzwerk diese Transaktion sehen kann.\\
Die Nodes überprüfen, ob der Absender genug Coins im Besitz hat (History der Blockchain).\\
Wenn die Transaktionsgebühr hoch genug ist, wird die Transaktion in den nächsten Block mit aufgenommen und wenn er voll ist (es passen ca. 4200 Transaktionen in einen Block) wird mit dem Mining begonnen.
\begin{center}
%\includegraphics[scale=0.4]{./images/alice.png}
\captionof{figure}[Abbildung 20 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Abfolge beim Senden einer Transaktion}
\end{center}

\section{Stablecoins - Coins zu Dollar}
Stablecoins werden verwendet um Crypto-Coins in eine reale Fiat-Währung wie den US-Dollar umzuwandeln.\\
Die Krypto-Tauschbörse \href{https://www.bitfinex.com/}{\color{blue}Bitfinex}
hat eine Kryptowährung erschaffen, die den Wert von einem US-Dollar abbildet.
Dafür wurden eine Milliarde US-Dollar auf einem Bankkonto hinterlegt und eine
Milliarde sogenannte \textbf{Tether} (USDT) zum Kauf angeboten.\\
Die Firma hinter Tether garantiert, dass wir für den Tether-Coin bei Bedarf auch einen US-Dollar ausbezahlt bekommen.\\
Bitfinex verdient nun, wie eine Bank, beim Tausch zwischen verschiedenen Coins und der Auszahlung (Rückkauf von Tether).\\
Kursschwankungen werden versucht auszugleichen:\\ 
Wenn die Coin-Nachfrage zu niedrig ist, gleicht Tether die Kursschwankungen
durch Kaufen des Coins aus. Ist die Coin-Nachfrage hingegen zu hoch, wirkt Te
ther dem entgegen, indem neue Coins ausgegeben werden, sodass der Nominalwert von 1 USD eingehalten werden kann.\vspace{.3cm}\\
Wichtig bei Stablecoins ist, wie auch im Papiergeldsystem, dass die Währung gedeckt ist. Gedeckt bedeutet, dass zu jedem Tether-Coin auch ein entsprechender US-Dollar auf einem Bankkonto eingefroren ist. Nur so kann der Wert von Tether sichergestellt werden. Stellen wir uns vor, es gäbe eine Milliarde Tether-Coins, aber es sind nur eine Million Dollar auf einem Bankkonto hinterlegt.\\
Aktuelles Beispiel (November 2022) wo dies gerade nicht mehr der Fall ist: \href{https://ftx.com/}{\color{blue}FTX}. Siehe \href{https://www.n-tv.de/wirtschaft/Bei-FTX-soll-mindestens-eine-Milliarde-Dollar-fehlen-article23713547.html}{\color{blue}FTX Zahlungsunfähigkeit}

\section{Probleme}
Bitcoin und die Crypto-Währungen werden häufig als das Zahlungsmittel der Zukunft verkauft. Es soll anonym, frei von Mittelsmännern (Banken) sein und keine Inflation aufweisen.\\
Anstelle einer stabilen Währung haben sie sich aber bis jetzt (November 2022) eher als hochspekulativ und instabil erwiesen.
\subsection{Hohe Volatilität}
Zu den grundlegenden hohen Kurschwankungen aufgrund des völlig unregulierten 24 Stunden Handels kommen noch Euphorie oder Panik hinzu. Zum Teil sind Tweets von Promis ausreichend, um massive Kursschwankungen auszulösen.
\subsubsection{Mittelsmänner}
Es braucht nach wie vor Mittelsmänner, statt Banken sind es nun \textbf{Crypto-Exchanges}. Auf Grund von fehlenden Regulierungen sind sie viel anfälliger gegen Missbrauch, gezielte Angriffe und Zahlungsausfall als traditionelle Banken.\\
Beispiel vom November 2022: \href{https://www.derstandard.at/story/2000140794202/kryptoboerse-kraken-will-mutmasslichen-ftx-hacker-identifiziert-haben}{\color{blue}FTX - Zahlungsausfall und Diebstahl}
\subsubsection{Fairness}
Auch von Fairness kann keine Rede sein, weil in beiden Konsensmechanismen (Proof-of-Work und Proof-of-Stake) immer die \textit{'starken'} oder \textit{'die schon viel besitzen'} belohnt werden.
\subsubsection{51\% Angriff} \label{51}
Bei Coins, die Proof-of-Work als Konsensmechanismus nutzen, kann eine Gruppe die über 51\% der gesamten Hashing-Power des Netzwerks besitzt, das Netzwerk komplett übernehmen. Sie sind schneller im Minen als alle anderen, und erzeugen so die längste Blockchain, welche von allen anderen Nodes als wahre Blockchain wahrgenommen wird.\\
Dies ist bei großen Coins wie Bitcoin zwar schwierig, aber nicht unmöglich. So erzielte der Mining-Pool GHash.IO im Juni 2014 über einen Zeitraum von 24 Stunden einen Anteil von etwa 55\% der Bitcoin-Hashrate.\\
Auf kleinere Coins werden aber immer wieder mit 51\% Angriffe durchgeführt. 
\subsubsection{Transaktionslimit}
ca. 3,3 Transaktionen pro Sekunde weil 1 MByte Blockgröße mit ca. 2000 Transaktionen und 10 min. Zeit für einen Block!\\
Führt im August 2017 zum Hard-Fork und der Abspaltung Bitcoin-Cash (hat nun 8 MB Blockgröße).\\
Zum Vergleich: Visa wickelt ca 65000 Transaktionen pro Sekunde ab, das sind fast 150 Millionen Transaktionen pro Tag!
\subsubsection{Finality}
Finality beschreibt, ab wann ein Teilnehmer der Blockchain sicher sein kann, dass seine Transaktion auch Teil der Blockchain - und somit wahr - ist. Da es ja immer wieder zu Forks kommen kann und sich die längste Blockchain nach meist 5-6 Blöcken (jeweils 10 Minuten pro Block) durchsetzt. Kann ich als Teilnehmer erst nach 6x10 Minuten sicher sein, ob meine Transaktion auch Teil der Blockchain ist!
\subsubsection{Energieaufwand bei Proof-of-Work}
Der Cambridge \href{https://ccaf.io/cbeci/index}{\color{blue}Bitcoin Electricity Consumption Index} hat das Mining des Bitcoin auf seinen Energieverbrauch hin untersucht. Zur Zeit (November 2022) liegt der Energieverbrauch bei ca.100 Terawattstunden. Im Vergleich dazu verbraucht Österreich pro Jahr ca. 70 Terawattstunden Energie.

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Vorlagen
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{cverbatim}
\end{cverbatim}

\begin{pverbatim}
\end{pverbatim}

\begin{center}
\fcolorbox{red}{rotlich}{\begin{minipage}{5cm}{
\fcolorbox{green}{gruenlich}{\begin{minipage}{15cm}{

}\end{minipage}}
\end{center}

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.4]{./images/leistung05.eps}
\end{center}
\end{figure}
\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A}
\subsection{B}
\subsubsection{C}
\paragraph{D}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-02-28 1_Version_Paper/Passwords.tex
added paper to source controll

--- 2023-02-28 1_Version_Paper/Networksecurity.tex
@@
 
-\documentclass[10pt,a4paper,titlepage]{paper}
+\documentclass[10pt,a4paper,titlepage]{article}
 \setcounter{secnumdepth}{5}
@@
 \pagestyle{fancy}
-\lhead{Cyber-security - Passwords: Version 0.1}
-\rhead{Team HTL Anichstraße (February 2023)}
+\lhead{Blockchain \& Crypto: Version 0.3}
+\rhead{Markus Signitzer (Jän. 2023)}
 
@@
 \begin{center}
-		\vspace*{1cm}
+		\vspace*{.5cm}
 
-		\Huge \bf Passwords\\
-		\vspace{5cm}
+		\Huge \bf Blockchain und Cryptowährungen\\
+		\vspace{.6cm}
 
-		%\includegraphics[scale=.35]{./images/title.png}\\
+		\includegraphics[scale=.35]{./images/title.png}\\
 
 		\vspace{1.5cm}
-		\huge A brief summary of solutions to the never dying problems of pins. \\
+		\huge Eine kompakte Zusammenfassung der grundlegenden Ideen.\\
 		\vspace{2cm}
@@
 
-		\Large Version 0.1\\
-		\large Febuary 2023\\
-		\large Team HTL Anichstraße\\[30pt]
+		\Large Version 0.3\\
+		\large Jänner 2023\\[30pt]
 		\vspace{1cm}
 
-		\vspace*{5cm}
+	\vspace*{1cm}
 
-        \large no \copyright\ by Lilo Zobl\\
+        \large no \copyright\ by Markus Signitzer\\
+        Jänner 2023\\
 
+        \vspace{5mm}
+        Typesetting by \LaTeX
 \end{center}
@@
 
-\section{Abstract}
+\section{Danksagung}
+Ich möchte mich an dieser Stelle herzlich bei Benjamin Spahic bedanken.\\ Er hat mir großzügigerweise erlaubt sein Buch \textit{Bitcoin ohne Vorkenntnisse} ISBN: 979-8732047134\\ \href{https://www.amazon.de/Bitcoin-ohne-Vorkenntnisse-Innerhalb-Blockchain/dp/B093B7T7YT}{\color{blue}https://www.amazon.de/Bitcoin-ohne-Vorkenntnisse-Innerhalb-Blockchain/dp/B093B7T7YT} \\als Grundlage für dieses Skript zu verwenden.
+\vspace{.3cm}\\
+Die Abbildung auf der Titelseite (Crypto Coins) stammt von \href{https://freepngimg.com/thumb/bitcoin/63423-binary-option-exchange-bitcoin-trade-cryptocurrency-investment.png}{\color{blue}freepngimg.com} und wurde unter der Creative Common Lizenz veröffentlicht.
 
-\section{Introduction}
-Password.
-Everyone has one or rather multiple.
-But it's a faulty construction.
-Nowadays passwords are everything but a good invention.\\ These are the reasons: i) Passwords are not user-friendly.
-In order to make passwords securer providers make them complexer.
-User have to write them down or use the same one over and over so they don't forget it.
-ii) Passwords are not secure because they can be shared and guessed.
-Of course a lot of passwords are also stolen, which is a lot easier than anyone would think.
-iii) Passwords are weak and get reused.
-Here are some examples to the most frequently used ones of 2022: "password", "123456", "123456789", "guest", "qwerty", ect.\\
-Luckily there is a websites that lets users check if their account information was breached (that's when your email connected to a password is leaked; always a huge amount of accounts get breached from a branch such as Facebook).
-The website is called "haveibeenpwned.com".
-HIBP is open source and secure.
-If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number).
-It only ever retrieves the data from storage then returns it in the response.
-%ich würd gern ein bild von der Seite einfügen hier
+\section{Überlegungen zum Einstieg in die Thematik}
+\subsection{Was ist Geld?}
+Bis circa 2000 Jahre vor Christus war es üblich, dass Güter und Waren direkt getauscht wurden.\\
+Dann etablierte sich die Idee, mit einem Ersatzprodukt zu handeln, das den Wert der verschiedenen Güter vereinheitlicht.
+Gold hat sich bis ins 16.
+Jahrhundert als primäre, stabile Währung durchgesetzt.
+\subsubsection{Welche Aufgaben hat Geld und welche Probleme löst es?}
+\begin{enumerate}
+\item Es muss \textbf{rar} sein!\\ 
+Rar bedeutet, dass der Rohstoff nur in begrenzter Menge zur Verfügung steht. Wenn die Währung (Rohstoff) einfach vermehrt werden könnte, würde ihr Wert sinken.
+\item Es muss \textbf{schwer herzustellen sein}!\\
+Dies dient primär der Fälschungssicherheit. Auch das ist bei Gold der
+Fall. Es ist sehr aufwendig, Gold aus der Erdkruste zu bergen und Münzen zu prägen.
+\item Es muss \textbf{ersetzbar} und \textbf{teilbar} sein!\\
+Ersetzbar bedeutet, dass mehrere Objekte innerhalb einer Währungsklasse den gleichen Wert besitzen.
+Z.B.
+kann Gold in gleich große und gleich schwere Münzen geprägt werden.
+Oder zwei kleine Münzen haben den Wert einer großen Münze usw.
+\item Es muss \textbf{transportabel} und \textbf{fälschungssicher} sein!
+\item \textbf{Vertrauen} in die Währung!\\
+Vertrauen ist der zentrale Baustein des Geldsystems, und dieses Vertrauen hat sich Gold über Tausende Jahre als sicheres Zahlungsmittel erarbeitet.
+Oftmals wird Gold daher auch als \textit{sicherer Hafen} der Vermögensanlage bezeichnet.\\
+Gold an sich hat keinen inneren Wert. Man kann es nicht essen oder
+Kleider daraus herstellen. Allein das \textbf{Vertrauen}, dass wir für ein Stück Gold von einem Dritten Ware erhalten werden, macht es möglich, Gold als Zahlungsmittel zu verwenden. 
+\end{enumerate}
+\subsubsection{Der Goldstandard}
+Gold und Goldmünzen haben jedoch auch Nachteile.
+So ist z.B.
+die Teilbarkeit begrenzt.
+Heutzutage liegt eine \\Feinunze Gold (1 oz.tr.
+= 31,1 g) bei ungefähr 1500-2000 USD.
+Möchten wir an der Kasse einen
+Eurocent Wechselgeld bekommen, würden wir umgerechnet also ca. 0,0002 g Gold zurückbekommen. Dies ist nicht praktikabel!\\
+Deshalb ist im 16.
+Jahrhundert die Idee aufgekommen, das Gold als Tauschbasis zu behalten, jedoch nicht mehr direkt mit dem Gold zu tauschen, sondern eine Art Schuldschein (Papiergeld) zu erstellen, mit dem einfacher gehandelt werden konnte.
+Wichtig ist in diesem Zusammenhang, dass hinter jedem dieser Schuld- bzw.
+Papierscheine ein Stück physisches Gold stand, das jederzeit gegen den Schuldschein eingetauscht werden konnte.
+Zentrale Banken
+gaben die Geldscheine aus und lagerten das hinterlegte Gold ein.
 
-\section{Haupteil?}
-Passwords have always been a security threat ever since they where introduced.
-Providers have tried to find solutions such as requiring users to make the pin at least eight characters long or use upper case and lower case letters.
-Sometimes they require you to include digits or symbols.
-However those "solutions" didn't improve the situation overall and sometimes made it worse.
-To eliminate this security threat a few solutions created.
+\subsubsection{Fiat- oder ungedeckte Währungen}
+Vom lateinischen Wort \textbf{fiat} (Es geschehe! Es werde!)\\
+Der Goldstandard wurde allmählich aufgelöst.
+Weil beispielsweise die Goldreserven im Ersten Weltkrieg nicht ausgereicht haben, um die Kriegskosten zu finanzieren, wurde kurzerhand das Papiergeld vom Goldpreis entkoppelt.\\
+Alle Währungen heutzutage sind \textbf{ungedeckte} Währungen.
+Der Wert wird über die Macht der Regierung gesichert, welche die Währung als gesetzliches Zahlungsmittel vorschreibt, sowie über die Akzeptanz und das Vertrauen der Handelspartner.
+\vspace{.3cm}\\
+Es gibt eine direkte \textbf{Korrelation} zwischen der \textbf{Stabilität einer Regierung} und dem \textbf{Wert der Währung}.\\
+Negativbeispiele: Die Griechenlandkrise oder die Wirtschaftskrise in Venezuela, bei der das Vertrauen in die Regierung und die Währung beinahe vollkommen verloren ging, woraus im Jahr 2018 laut dem Internationalen Währungsfonds eine Inflationsrate von über 130000\% resultierte.
+\vspace{.3cm}\\
+Dies zeigt uns auch das große Problem aller Fiat-Währungen auf: Dadurch, dass sie nicht mehr gedeckt sind und durch eine zentrale Verwaltungsstelle in beliebigen Massen gedruckt werden können, ist die Währung nicht vor Inflation geschützt.
 
-\subsection{Password Save}
-\subsubsection{Explanation}
-A password save allows you to save a list of your usenames linked to the password.
-It has one master password that looks the save.
-Instead of remembering a ton of password at once users only have to recall one single pin.
-The rest is done by the app or rather the software.
-\subsubsection{Providers}
-\begin{itemize}
-\item \textbf{KeepassXC}\\
-This software is developed for users with extremely high demands of secure personal data management. \\
-KeePassXC uses Advanced Encryption Standard (AES) encryption algorithm with a 256-bit key to secure the password database.\\
-The biggest difference to other password saves is that the data (password, account information and additional data such as URLs, attachments ans notes) is stored in an offline, encrypted file that can be stored locally.
-This prevents your data from getting leaked when the could gets hacked or breached.\\
-The program is customizable. It allows its user to customize literally everything to their needs.\\
+\subsubsection{Digitalisierung des Geldes}
+Das klassische Geldsystem ist verbunden mit langsamen und nicht zeitgemäßen Strukturen.
+Mann muss über einen Mittelsmann (Bank) gehen um Überweisungen zu tätigen,  die dann meist noch ein bis zwei Tage dauern.\\Technisch gesehen sind schnellere Transaktionen jedoch durchaus leicht möglich - siehe z.B.
+PayPal.\\
+Die Digitalisierung des Geldes bringt zunächst viele Vorteile mit sich. So sparen wir beispielsweise enorme Lagerkosten und können das Geld schneller und leichter verschieben.\\
+Auch ist die Vermehrung des Geldes einfacher geworden.
+Möchte eine Zentralbank neues Geld drucken, müssen keine großen Gelddruckereien angeworfen werden.
+Es müssen keine Geldtransporter anfahren, um das Geld zu verteilen.
+Mit wenigen Mausklicks am PC kann innerhalb von Sekunden Geld in Milliardenhöhe aus dem Nichts (\textit{fiat} - Es geschehe!
+Es werde!) erschaffen werden.
 
-\item \textbf{1Passwort}\\
-This password save is known for its easy to use interface and high security encryption.\\
-1Password uses an uncommon encryption known as dual-key encryption.
-If the server gets breached it's impossible for the hacker to decrypt the users sensitive information because of its two keys.
-The first key is the users master-key.
-The second key is a secret key, which is a 128-bit, machine-generated code.
-The secret key is generated on every device you log into.
-It will only be saved on your devices and never saved with your other pins.\\
-Depending on what account type you choose your data is stored differently.
-However in only one version users have the option to save their data locally.
-In every other option the user's data gets saved in a cloud-based vault.\\
-The interface is very user-friendly. It even creates strong passwords for accounts that are newly created.
-\item \textbf{Password Safe (MATESO)}\\
-
-\item \textbf{LastPass}\\
-\item \textbf{Dashlane}\\
-\end{itemize}
-
-\subsection{2-factor authentication}
-\subsection{The future of passwords}
-
-\section{conclusion}
+\subsection{Kryptowährungen}
+Die Verbindung aus zentralen Strukturen, die Geldmonopole bilden, und der Digitalisierung, hat die Vermehrung und Entwertung von Geld vereinfacht.
+In all den Punkten wie Vertrauen, Zentralisierung und veralteten Strukturen setzen die Kryptowährungen an.
+Sie versuchen diese Probleme zu lösen z.B.
+durch Dezentralisierung durch \textit{neue} Werte -> \textbf{Kryptowährungen beruhen} auf \textbf{Mathematik} und \textbf{Kryptographie}.
 
@@
 
-%\includegraphics[scale=0.35]{./images/hash.eps}
+\includegraphics[scale=0.35]{./images/hash.eps}
 \captionof{figure}[Hashing Function - die Abbildung stammt vom Cisco Netacad Network Security Kurs]{Hashing Function}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/hmac.png}
+\includegraphics[scale=0.4]{./images/hmac.png}
 \captionof{figure}[HMAC - die Abbildung stammt vom Cisco Netacad Network Security Kurs]{HMAC - hash bases message authentication}
@@
 \begin{center}
-%\includegraphics[scale=0.43]{./images/symmetric-encryption.png}
+\includegraphics[scale=0.43]{./images/symmetric-encryption.png}
 \captionof{figure}[Symmetric-Encryption - die Abbildung stammt vom https://www.cheapsslshop.com/blog/symmetric-vs-asymmetric-encryption-whats-the-difference]{Symmetrische Verschlüsselung}
@@
 \begin{center}
-%\includegraphics[scale=0.43]{./images/asymmetric-encryption.png}
+\includegraphics[scale=0.43]{./images/asymmetric-encryption.png}
 \captionof{figure}[Asymmetric-Encryption - die Abbildung stammt vom https://www.cheapsslshop.com/blog/symmetric-vs-asymmetric-encryption-whats-the-difference]{Asymmetrische Verschlüsselung}
@@
 \begin{center}
-%\includegraphics[scale=0.3]{./images/curve.png}
+\includegraphics[scale=0.3]{./images/curve.png}
 \captionof{figure}[ECC-Multiplication - die Abbildung stammt vom https://www.oreilly.com/library/view/mastering-bitcoin-2nd/9781491954379/ch04.html]{ECC-Multiplikation veranschaulicht}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/block.png}
+\includegraphics[scale=0.4]{./images/block.png}
 \captionof{figure}[Abbildung 22 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Hash-Generierung aus den Daten eines Blocks}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/blocks.png}
+\includegraphics[scale=0.4]{./images/blocks.png}
 \captionof{figure}[Abbildung 23 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Aneinanderreihung von Blöcken}
@@
 \begin{center}
-%\includegraphics[scale=0.35]{./images/bc-size.png}
+\includegraphics[scale=0.35]{./images/bc-size.png}
 \captionof{figure}[https://www.blockchain.com/explorer/charts/blocks-size]{Wachstum der Bitcoin Blockchain seit 2009}
@@
 \begin{center}
-%\includegraphics[scale=0.5]{./images/fork-01.png}
+\includegraphics[scale=0.5]{./images/fork-01.png}
 \captionof{figure}[https://www.researchgate.net/figure/The-fork-forms-and-disappears-in-the-PoW-Blockchain]{Ein Fork - die Blockchain ist nicht mehr einheitlich}
@@
 \begin{center}
-%\includegraphics[scale=0.45]{./images/fork-02.png}
+\includegraphics[scale=0.45]{./images/fork-02.png}
 \captionof{figure}[https://www.researchgate.net/figure/The-fork-forms-and-disappears-in-the-PoW-Blockchain]{Ein Fork löst sich wieder auf - die Blockchain ist wieder einheitlich}
@@
 \begin{center}
-%\includegraphics[scale=0.3]{./images/mining.png}
+\includegraphics[scale=0.3]{./images/mining.png}
 \captionof{figure}[Abbildung 24 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Hashgenerierung aus den Daten eines Blocks inkl. der zusätzliche Variable \textbf{Nonce}}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/fees.png}
+\includegraphics[scale=0.4]{./images/fees.png}
 \captionof{figure}[https://www.blockchain.com/explorer/charts/fees-usd-per-transaction]{Transaktionsgebühren (Blau) und Bitcoin Marktwert (Schwarz)}
@@
 \begin{center}
-%\includegraphics[scale=0.5]{./images/adress.png}
+\includegraphics[scale=0.5]{./images/adress.png}
 \captionof{figure}[https://www.blockchain.com/btc/address/1A1zP1eP5QGefi2DMPTfTL5SLmv7DivfNa]{"Kontostand" der Genesis Adresse - erste Adresse von Satoshi Nakamoto}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/key_generation.png}
+\includegraphics[scale=0.4]{./images/key_generation.png}
 \captionof{figure}[Abbildung 17 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Generierung der Bitcoin-Adresse}
@@
 \begin{center}
-%\includegraphics[scale=0.4]{./images/alice.png}
+\includegraphics[scale=0.4]{./images/alice.png}
 \captionof{figure}[Abbildung 20 aus dem Buch: Bitcoin ohne Vorkenntnisse]{Abfolge beim Senden einer Transaktion}

### 2023-02-28 Aktuelle_Version_Paper/Conference-LaTeX-template_10-17-19/cyber_security_passwords.tex
added paper to source controll

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{The role of passwords in cybersecurity}

\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
lzobl@tsn.at}}

\maketitle

\begin{abstract}
This document is part of the ECER 2023 confrence on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
\end{abstract}
\vspace{2cm}
\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This paper will focus on the wide use of passwords in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
Additionally it will try to highlight alternatives for effective password management and introduce a choice of selected password managing software. \\
Finally the paper includes a best practice guide to:
\begin{itemize}
\item check if user's passwords have already been breached 
\item create an easy to remember but secure master password for a password managing software
\end{itemize} 
\section{Current and future state of the art}
%alles was im intro aufgezählt wurde

\section{Concept and implementation for best practise}
%Schritt für Schritt Anleitung wie man sein system auf neuesten stand bringt

\end{document}

### 2023-02-28 Aktuelle_Version_Paper/cyber_security_passwords.tex
added paper to source controll

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{The role of passwords in cybersecurity}

\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
lzobl@tsn.at}}

\maketitle

\begin{abstract}
This document is part of the ECER 2023 confrence on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
\end{abstract}
\vspace{2cm}
\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This paper will focus on the wide use of passwords in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
Additionally it will try to highlight alternatives for effective password management and introduce a choice of selected password managing software. \\
Finally the paper includes a best practice guide to:
\begin{itemize}
\item check if user's passwords have already been breached 
\item create an easy to remember but secure master password for a password managing software
\end{itemize}

\section{State of the art}
Passwords are used to protect sensitive information. For example users need to key in a user-name and password before they can access any network device over SSH. This is also the case if a client wants to access the raspberry pi inside the wombat controller remotely.

\subsection{The problem with passwords}
As mentioned earlier a password protects sensitive information that users don't want to be leaked. That's why it's so important to have a good, secure password. However users often are too lazy to create a strong password for every account they own. But that's not the only reason. A study from NordPass conducts that the average user has around a 100 accounts that require a unique password. \\
Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
Every year long record gets published by different providers reviling the most used passwords. Here are the top ten used passwords of last years:
\vspace{0.5cm}
\begin{enumerate}
\item 123456
\item 123456789
\item qwerty
\item password
\item 12345
\item 12345678
\item 111111
\item 1234567
\item 123123
\item 1234567890
\end{enumerate}

Also common password are locations, names, birthdays, ect.

\subsection{Current state of art}
In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
2-factor-authentication adds an extra layer of security to your password. First the user has to key in the username and password, which is the same as always. Then the user has to answer an additional question to verify it's really him/her.\\
The second factor has different types of categories:
\begin{itemize}
\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), ect.
\item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
\item \textbf{Something the user is.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
\end{itemize}
A common type of the 2-factor-authentication is the SMS Text-Message. After entering the password and username the user receives a unique one time passcode (OTP) via text message. They need to key in the passcode in a limited time (for example a minute). After both the first factor and second factor are correct the user gains access.

\subsection{Future state of art}
The future of passwords is passkeys, which eliminates the need for username and password all together.\\

Passkeys are easier to use, securer, faster and work on most of the user's devices.
\subsection{Solutions to password managing}

\subsection{Password Save}
\subsubsection{Explanation}
A password save allows you to save a list of your usenames linked to the password. It has one master password that looks the save. Instead of remembering a ton of password at once users only have to recall one single pin. The rest is done by the app or rather the software.
\subsubsection{Providers}
\begin{itemize}
\item \textbf{KeepassXC}\\
This software is developed for users with extremely high demands of secure personal data management. \\
KeePassXC uses Advanced Encryption Standard (AES) encryption algorithm with a 256-bit key to secure the password database.\\
The biggest difference to other password saves is that the data (password, account information and additional data such as URLs, attachments ans notes) is stored in an offline, encrypted file that can be stored locally. This prevents your data from getting leaked when the could gets hacked or breached.\\
The program is customizable. It allows its user to customize literally everything to their needs.\\

\item \textbf{1Passwort}\\
This password save is known for its easy to use interface and high security encryption.\\
1Password uses an uncommon encryption known as dual-key encryption. If the server gets breached it's impossible for the hacker to decrypt the users sensitive information because of its two keys. The first key is the users master-key. The second key is a secret key, which is a 128-bit, machine-generated code. The secret key is generated on every device you log into. It will only be saved on your devices and never saved with your other pins.\\
Depending on what account type you choose your data is stored differently. However in only one version users have the option to save their data locally. In every other option the user's data gets saved in a cloud-based vault.\\
The interface is very user-friendly. It even creates strong passwords for accounts that are newly created.
\item \textbf{Password Safe (MATESO)}\\

\item \textbf{LastPass}\\
\item \textbf{Dashlane}\\
\end{itemize}

\section{Concept and implementation for best practise}
%Schritt für Schritt Anleitung wie man sein system auf neuesten stand bringt

\section{Conclusion}

\end{document}

### 2023-03-05 Aktuelle_Version_Paper/passwords_cyber_security.tex
Finished 1st and 2nd Version of paper

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{The role of passwords in cybersecurity}
\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
lzobl@tsn.at}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Matteo Reiter}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
mareiter@tsn.at}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Niclas Prantl}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
nprantl@tsn.at }
\and
\IEEEauthorblockN{4\textsuperscript{th} Lukas Krahbichler}
\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
\textit{HTL Anichstraße}\\
Innsbruck, Austria \\
lkrahbichler@tsn.at}
}
\maketitle

\begin{abstract}
This document is part of the ECER 2023 confrence on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
\end{abstract}
\vspace{1cm}

\section{Introduction}
This paper will focus on the wide use of passwords in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
Additionally it will try to highlight alternatives for effective password management and introduce a choice of selected password managing software. \\
Finally the paper includes a best practice guide to:
\begin{itemize}
\item check if user's passwords have already been breached 
\item create an easy to remember but secure master password for a password managing software
\end{itemize}

\section{State of the art}
Passwords are used to protect sensitive information or access to hardware.\\
 For example users need to key in a user-name and password before they can access any network device over SSH. This is also the case if a client wants to access the raspberry pi inside the wombat controller remotely.

\subsection{The problem with passwords} \cite{b1} \cite{b2}
As mentioned above a password protects sensitive information or access to hardware like a robot done. That's why it's so important to have a good, secure password. However users often are too lazy to create a strong password for every account they own. This is understandable, because as a study from NordPass conducts, the average user has around a 100 accounts that require a unique password.Remembering a hundred unique Passwords or more is really beyond a users capability.\\
Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last years: \cite{b3}
\begin{enumerate}
\item 123456
\item 123456789
\item qwerty
\item password
\item 12345
\item 12345678
\item 111111
\item 1234567
\item 123123
\item 1234567890
\end{enumerate}
Also common password are locations, names, birthdays, ect.

\subsection{Current state of art}
\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
2-factor-authentication adds an extra layer of security to your password. First the user has to key in the username and password, which is the same as always. Then the user has to answer an additional question to verify it's really him/her.\\
The second factor has different types of categories:
\begin{itemize}
\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), ect.
\item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
\item \textbf{Something unique to the user.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
\end{itemize}
A common type of the 2-factor-authentication is the SMS Text-Message. After entering the password and username the user receives a unique one time passcode (OTP) via text message. They need to key in the passcode in a limited time (for example a minute). After both the first factor and second factor are correct the user gains access.

\subsection{Future state of art}
\cite{b5} In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the users device's security method such as a pin or a biometric sensor (fingerprint, face scan). \\
The passkey is hardware specific, which means user always need to carry the device (that the passkey is installed on) with them. If user want to use passkey on their laptop they need to verify their identity with their smartphone. This eliminates phishing, credential stuffing and other remote attacks. \\
In conclusion when a user is asked to sign-in to an app or website, the user approves the sign-in with the same biometric or PIN that the user has to unlock the device 
(phone, computer or security key). The app or website can use this mechanism instead of the traditional (and insecure) username and password.

\subsection{Solutions to password managing}
If user don't want or can't to use passkeys there are different options how they could protect their sensitive information. User can write their passwords with account info into a booklet. The passwords would be unique and strong. It's a really simple idea but also has a lot of drawbacks. 
For example:
\begin{itemize}
\item User need to carry the booklet with them everywhere. If not they can't always access their accounts.
\item User can lose the booklet. If that happens they can't access any of their accounts any more.
\item It's not a really clear way of organizing passwords. Once something is written down user can't change the position in the booklet. So users would have to search for each password for indefinite time before they find it.
\end{itemize}
The other and better option is a password safe.

\subsection{Password Safe}
\subsubsection{Explanation}
A password safe allows you to save a list of your usernames and passwords for all your accounts. It has one master password or passphrase that unlocks the safe. Now users only have to recall one single master passphrase instead of a huge amount of passwords.
\subsubsection{Examples}
\begin{itemize}
\item \textbf{KeepassXC}\cite{b6}\\ 
This software is developed for users with extremely high demands of secure personal data management. \\
KeePassXC uses Advanced Encryption Standard (AES) encryption algorithm with a 256-bit key to secure the password database.\\
The biggest difference to other password safes is that the data (password, account information and additional data such as URLs, attachments and notes) is stored in an offline, encrypted file that can be stored locally. This prevents user's data from getting leaked when a server gets hacked or breached.\\
The XC stands for Cross-Plattform. It is possible to use on it on Linux, Windows, macOS and Android.
The program is customizable. It allows its user to customize literally everything to their needs.\\
\includegraphics[scale=0.77]{./images/KeePass.png}\\

\item \textbf{1Password}\cite{b7}\\
This password safe is known for its easy to use interface and high security encryption.\\ 
1Password uses an uncommon encryption known as dual-key encryption. If the server gets breached it's impossible for the hacker to decrypt the users sensitive information because of its two keys. The first key is the users master-key. The second key is a secret key, which is a 128-bit, machine-generated code. The secret key is generated on every device the user log into. It will only be saved on user's devices and never saved with the other pins.\\
Depending on what account type user choose their data is stored differently. However in only one version users have the option to save their data locally. In every other option the user's data gets saved in a cloud-based vault.\\
The interface is very user-friendly. It even generates strong passwords for accounts that are newly created.
\includegraphics[scale=0.077]{./images/1Password-Logo.wine.png}\\
\end{itemize}
\subsubsection{Benefits and drawbacks of password safe software}\cite{b8}
\textbf{Benefits}:
\begin{itemize}
\item Security: Password safes use advanced encryption algorithms to protect user's passwords. This makes it extremely difficult for hackers to steal the passwords and gain access to user's online accounts.
\item Convenience: Password safes store all of a user's passwords in one place, which makes it easy for them to access their accounts without having to remember multiple passwords.
\item Auto-fill feature: Many password safes have an auto-fill feature that automatically enters the username and password for user, saving them time and reducing the risk of typing errors.
\item Multi-device access: Many password safes allow user to access their passwords from multiple devices, including smartphones, tablets, and computers.
\item Password strength: Password safes often include a password generator feature, which creates strong passwords for user. This helps to ensure that their passwords are not easily guessable.
\end{itemize}
\textbf{Drawbacks}:
\begin{itemize}
\item Single point of failure: If a user's password safe is hacked or compromised, all of their passwords are at risk. It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
\item Dependency: Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if a user is not able to access their password safe for some reason.
\item Learning curve: Some password safes have a steep learning curve, and it can take some time to get used to using them effectively.
\item Cost: Some password safes charge a fee for their services, which can be a disadvantage if user are looking for a free solution.
\item Compatibility: Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
\end{itemize}

\section{Concept and implementation for best practise}
It can be very overwhelming to start with a secure option to manage passwords. So here is a best practise guide everyone can follow.
\begin{enumerate}
\item First you should check if your password has already been breached. There is a website that let's you easily check just that. It's called \textbf{"haveibeenpwned.com"}.\\
Haveibeenpwned is open source and secure. If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number). It only ever retrieves the data from storage then returns it. \vspace{0.5cm}
\includegraphics[scale=0.32]{./images/haveibeenpwned.png}\\
If you have already been pawned change that password. If not you can rest assured. 
\item Pick a password safe that fits best to your needs. You can choose one of the ones mentioned above or do your own research.
\item Create a good master password or better a master passphrase. Here is a easy way to do it: 
\begin{itemize}
\item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
\item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
\item Thirdly you can change letters into numbers. A "s" gets changes into a "5", an "e" gets changed into a "3" and an "o" gets changed into a "0". For example "Iluvappl5andbanana5" or "Ik0mmau5da5tadtWi3n".
\item The fourth step is to add symbols to your password. Change characters that look like a symbol into one such as "a" to "@" or "s" to "?". For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
\item The fifth step is to make all words start with a capital letter. This is only needed if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example). 
\end{itemize}
In the end you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n". This is a really secure password because it has numbers, upper case and lower case letters, symbols and more that 8 characters. This would make a strong master password that is nearly impossible to crack.
\end{enumerate}

\section{Conclusion}
Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable to attacks. While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and especially creative approaches as well as authentication and access control, that are both secure and user-friendly.

\begin{thebibliography}{00}
\bibitem{b1} Proof for Research, URL:https://tech.co/password-managers/how-many-passwords-average-person

\bibitem{b2} Information for "The problem with passwords", URL:https://www.nomios.com/news-blog/password-problem/

\bibitem{b3} List for most used passwords, URL:https://www.passwordmanager.com/most-common-passwords-latest-2022-statistics/

\bibitem{b4} Information for "Current state of art", URL:\\ https://authy.com/what-is-2fa/

\bibitem{b5} Information for "Future state of art", URL:\\https://fidoalliance.org/passkeys/\\ https://developer.apple.com/passkeys/\\ https://www.passkeys.io/\\ https://developers.google.com/identity/passkeys?hl=de

\bibitem{b6} KeepassXC, Open source password manager, URL: https://keepassXC.org

\bibitem{b7} 1Password, Paid password manager, URL: https://1password.com/de/

\bibitem{b8} Information for "Benefits and drawbacks of password safe software", URL: \\
https://expert.services/blog/managing-your-website/security/password-managers;\\ https://www.orangecountyscu.org/stories/pros-and-cons-of-using-a-password-manager/; \\ https://www.passwordboss.com/pros-and-cons-of-using-a-password-manager/; \\https://www.trustworthy.com/blog/pros-and-cons-password-manager/

\bibitem{b9} HaveIbeenpawnd Website, Screenshot form Website, URL: https://haveibeenpwned.com/

\bibitem{b10} 1Password Logo form Website, URL: https://www.logo.wine/logo/1Password

\bibitem{b11} KeePass Logo form Website, URL: https://fosstorrents.com/softwares/keepassxc/

\end{thebibliography}

\end{document}

### 2023-03-05 Aktuelle_Version_Paper/passwords_cyber_security.tex
Gramma and spelling errors corrected

@@
 \usepackage{xcolor}
+\usepackage{caption}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 \begin{abstract}
-This document is part of the ECER 2023 confrence on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
+This document is part of the ECER 2023 conference on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
 \end{abstract}
@@
 \section{Introduction}
-This paper will focus on the wide use of passwords in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
+This paper will focus on the wide use of passwords or passphrases in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
 It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
@@
 For example users need to key in a user-name and password before they can access any network device over SSH.
-This is also the case if a client wants to access the raspberry pi inside the wombat controller remotely.
+This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely.
 
 \subsection{The problem with passwords} \cite{b1} \cite{b2}
-As mentioned above a password protects sensitive information or access to hardware like a robot done.
+As mentioned above a password protects sensitive information or access to hardware like a robot.
 That's why it's so important to have a good, secure password.
 However users often are too lazy to create a strong password for every account they own.
-This is understandable, because as a study from NordPass conducts, the average user has around a 100 accounts that require a unique password.Remembering a hundred unique Passwords or more is really beyond a users capability.\\
-Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
-Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last years: \cite{b3}
+This is understandable, because as a study from NordPass shows, the average user has around a 100 accounts that require a unique password.
+Remembering a hundred unique passwords or more is really beyond a users capability.\\
+Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases.\\
+Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year: \cite{b3}
 \begin{enumerate}
@@
 \end{enumerate}
-Also common password are locations, names, birthdays, ect.
+Other common password are locations, names, birthdays, etc.
 
 \subsection{Current state of art}
-\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
-2-factor-authentication adds an extra layer of security to your password.
+\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\\
+2-factor-authentication adds an extra layer of security to passwords.
 First the user has to key in the username and password, which is the same as always.
@@
 \item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
-\item \textbf{Something unique to the user.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
+\item \textbf{Something unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print. 
 \end{itemize}
 A common type of the 2-factor-authentication is the SMS Text-Message.
-After entering the password and username the user receives a unique one time passcode (OTP) via text message.
-They need to key in the passcode in a limited time (for example a minute).
+After entering the password and user-name the user receives a unique one time passcode (OTP) via text message.
+They need to key in the passcode during a limited time (for example one minute).
 After both the first factor and second factor are correct the user gains access.
@@
 \\
-The passkey is hardware specific, which means user always need to carry the device (that the passkey is installed on) with them.
-If user want to use passkey on their laptop they need to verify their identity with their smartphone.
+The passkey is hardware specific, which means the user always needs to carry the device (that the passkey is installed on) with them.
+If the user wants to use passkey on their laptop they need to verify their identity with their smartphone.
 This eliminates phishing, credential stuffing and other remote attacks.
 \\
-In conclusion when a user is asked to sign-in to an app or website, the user approves the sign-in with the same biometric or PIN that the user has to unlock the device 
-(phone, computer or security key). The app or website can use this mechanism instead of the traditional (and insecure) username and password.
+In conclusion when a user is asked to sign-in to an app or website, he/she approves the sign-in with the same biometric or PIN that he/she uses to unlock the device (phone, computer or security key).
+The app or website can use this mechanism instead of the traditional (and insecure) username and password.
 
 \subsection{Solutions to password managing}
-If user don't want or can't to use passkeys there are different options how they could protect their sensitive information.
-User can write their passwords with account info into a booklet.
+If users don't want to or can't use passkeys there are different options how they could protect their sensitive information.
+Users can write their passwords with account info into a booklet.
 The passwords would be unique and strong.
@@
 \begin{itemize}
-\item User need to carry the booklet with them everywhere. If not they can't always access their accounts.
-\item User can lose the booklet. If that happens they can't access any of their accounts any more.
+\item Users need to carry the booklet with them everywhere. If not they can't always access their accounts.
+\item Users can lose the booklet. If that happens they can't access any of their accounts any more.
 \item It's not a really clear way of organizing passwords.
@@
 \subsubsection{Explanation}
-A password safe allows you to save a list of your usernames and passwords for all your accounts.
+A password safe allows you to save a list of your user-names and passwords for all your accounts.
 It has one master password or passphrase that unlocks the safe.
@@
 \item \textbf{KeepassXC}\cite{b6}\\ 
+\begin{center}
+\includegraphics[scale=0.6]{./images/KeePass.png}
+\captionof{figure}[KeePassXC - The figure is taken from the homepage of the KeePassXC Project (https://keepassxc.org/)]{KeePassXC Logo}
+\end{center}
+
 This software is developed for users with extremely high demands of secure personal data management. \\
@@
 This prevents user's data from getting leaked when a server gets hacked or breached.\\
-The XC stands for Cross-Plattform. It is possible to use on it on Linux, Windows, macOS and Android.
-The program is customizable. It allows its user to customize literally everything to their needs.\\
-\includegraphics[scale=0.77]{./images/KeePass.png}\\
+The XC stands for Cross-Plattform. It is possible to use it on Linux, Windows, macOS and Android.\\
+The program is customizable. It allows the user to customize literally everything to their needs.\\
 
 \item \textbf{1Password}\cite{b7}\\
+\begin{center}
+\includegraphics[scale=0.06]{./images/1Password-Logo.wine.png}
+\captionof{figure}[1Password - The figure is taken from the homepage of the 1Password Project (https://1password.com/)]{1Password Logo}
+\end{center}
 This password safe is known for its easy to use interface and high security encryption.\\ 
 1Password uses an uncommon encryption known as dual-key encryption.
-If the server gets breached it's impossible for the hacker to decrypt the users sensitive information because of its two keys.
-The first key is the users master-key.
+If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of its two keys.
+The first key is the user's master-key.
 The second key is a secret key, which is a 128-bit, machine-generated code.
-The secret key is generated on every device the user log into.
-It will only be saved on user's devices and never saved with the other pins.\\
-Depending on what account type user choose their data is stored differently.
+The secret key is generated on every device the user logs into.
+It will only be saved on the user's devices and never saved together with the other pins.\\
+Depending on what account type users choose their data is stored differently.
 However in only one version users have the option to save their data locally.
@@
 The interface is very user-friendly. It even generates strong passwords for accounts that are newly created.
-\includegraphics[scale=0.077]{./images/1Password-Logo.wine.png}\\
+
 \end{itemize}
@@
 \begin{itemize}
-\item Security: Password safes use advanced encryption algorithms to protect user's passwords.
-This makes it extremely difficult for hackers to steal the passwords and gain access to user's online accounts.
-\item Convenience: Password safes store all of a user's passwords in one place, which makes it easy for them to access their accounts without having to remember multiple passwords.
-\item Auto-fill feature: Many password safes have an auto-fill feature that automatically enters the username and password for user, saving them time and reducing the risk of typing errors.
-\item Multi-device access: Many password safes allow user to access their passwords from multiple devices, including smartphones, tablets, and computers.
-\item Password strength: Password safes often include a password generator feature, which creates strong passwords for user. This helps to ensure that their passwords are not easily guessable.
+\item \textbf{Security:} Password safes use advanced encryption algorithms to protect the user's passwords.
+This makes it extremely difficult for hackers to steal the passwords and gain access to the user's online accounts.
+\item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without having to remember multiple passwords.
+\item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the user-name and password for the user, saving them time and reducing the risk of typing errors.
+\item \textbf{Multi-device access:} Many password safes allow user to access their passwords from multiple devices, including smartphones, tablets, and computers.
+\item \textbf{Password strength:} Password safes often include a password generator feature, which creates strong passwords. This helps to ensure that the passwords are not easily guessable.
 \end{itemize}
@@
 \begin{itemize}
-\item Single point of failure: If a user's password safe is hacked or compromised, all of their passwords are at risk.
+\item \textbf{Single point of failure:} If a user's password safe is hacked or compromised, all of their passwords are at risk.
 It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
-\item Dependency: Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if a user is not able to access their password safe for some reason.
-\item Learning curve: Some password safes have a steep learning curve, and it can take some time to get used to using them effectively.
-\item Cost: Some password safes charge a fee for their services, which can be a disadvantage if user are looking for a free solution.
-\item Compatibility: Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
+\item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if a user is not able to access their password safe for some reason.
+\item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to using them effectively.
+\item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if user are looking for a free solution.
+\item \textbf{Compatibility:} Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
 \end{itemize}
@@
 \vspace{0.5cm}
-\includegraphics[scale=0.32]{./images/haveibeenpwned.png}\\
-If you have already been pawned change that password. If not you can rest assured. 
-\item Pick a password safe that fits best to your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password or better a master passphrase. Here is a easy way to do it: 
+\begin{center}
+\includegraphics[scale=0.29]{./images/haveibeenpwned.png}
+\captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{Have I been pwned? Homepage}
+\end{center}
+If you have already been pawned, change that password. If not you can rest assured. 
+\item Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
+\item Create a good master password or even better a master passphrase. Here is a easy way to do it: 
 \begin{itemize}
@@
 \item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
-\item Thirdly you can change letters into numbers.
+\item Third, you can change letters into numbers and remove spaces.
 A "s" gets changes into a "5", an "e" gets changed into a "3" and an "o" gets changed into a "0".
@@
 In the end you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-This is a really secure password because it has numbers, upper case and lower case letters, symbols and more that 8 characters.
+This is a really secure passphrase because it has numbers, upper case and lower case letters, symbols and more that 8 characters.
 This would make a strong master password that is nearly impossible to crack.
@@
 While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
-As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and especially creative approaches as well as authentication and access control, that are both secure and user-friendly.
+As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and especially creative approaches to authentication and access control, that are both secure and user-friendly.
+
+%Abbildungsverzeichnis erstellen
+\listoffigures
 
@@
 
-\bibitem{b9} HaveIbeenpawnd Website, Screenshot form Website, URL: https://haveibeenpwned.com/
-
-\bibitem{b10} 1Password Logo form Website, URL: https://www.logo.wine/logo/1Password
-
-\bibitem{b11} KeePass Logo form Website, URL: https://fosstorrents.com/softwares/keepassxc/
-
 \end{thebibliography}

### 2023-03-21 4te_Version_Paper/passwords_cyber_security.tex
Changes from Signitzer were implemented andchanged

--- 2023-03-05 Aktuelle_Version_Paper/passwords_cyber_security.tex
@@
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Matteo Reiter}
+\IEEEauthorblockN{2\textsuperscript{th} Lukas Krahbichler}
+\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
+\textit{HTL Anichstraße}\\
+Innsbruck, Austria \\
+lkrahbichler@tsn.at}
+\and
+\IEEEauthorblockN{3\textsuperscript{nd} Matteo Reiter}
 \IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
@@
 \and
-\IEEEauthorblockN{3\textsuperscript{rd} Niclas Prantl}
+\IEEEauthorblockN{4\textsuperscript{rd} Niclas Prantl}
 \IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
@@
 nprantl@tsn.at }
-\and
-\IEEEauthorblockN{4\textsuperscript{th} Lukas Krahbichler}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-lkrahbichler@tsn.at}
 }
@@
 \begin{abstract}
-This document is part of the ECER 2023 conference on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
+This document follows the topic of cybersecurity and will have a look on the indispensable role of passwords now and in the future, especially in terms of robots.
+
 \end{abstract}
-\vspace{1cm}
 
 \section{Introduction}
-This paper will focus on the wide use of passwords or passphrases in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
+This paper will concentrate on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
+It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems.
+\\
 It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
+Also it is going to show why passwords are so important in robotics. \\
 Additionally it will try to highlight alternatives for effective password management and introduce a choice of selected password managing software. \\
@@
 
+\section{Implications of weak passwords}
+Many robots in today's world are controlled remotely.
+To regulate access, users will need a high quality password.
+Especially in robotics using a weak one can cause major problems.
+Users don't want anybody to have access to their robots, because could keep private information about them and their house.
+
+\subsection{How weak passwords can cause problems}
+
+\begin{itemize}
+\item \textbf{Robot vacuum cleaner:} A harmless example is the case of Mister Greg.
+The computer scientist took part in a beta program where he was gifted a robot vacuum cleaner with an additional camera for orientation.
+What he didn't know was that a picture of him, sitting on the toilet, would find it's way onto FaceBook.
+\cite{b10}\\
+These days many people already got themselves a robot vacuum cleaner with a camera.
+In most cases it communicates with the docking station or a device of its owner (e.g.: smart phone) remotely.
+Because it's accessed remotely the connections have to be secured with a strong password.
+If not the user could undergo a similar experience as Mister Greg.
+\item \textbf{Military drones:} A extreme case of weak passwords is in the military, because spying and attack drones are already an important part there and will become even more significant in the future.
+The major problems is that these drones have to communicate with some kind of wireless radio signal.
+These signals are very vulnerable for hacking attacks.
+If somebody gains access to enemy robots, they can do a lot of damage.
+For example destroying military bases with the hacked drones.
+This can be especially dangerous because people normally don't know that they are dealing with a hacked drone.\\
+All of this could be prevented by a strong and secure password.
+\item \textbf{In the future:} Boston Dynamics is a company which focus on developing robots for the world of tomorrow.
+Robots are developing so fast that soon everyone will have their own personal assistant robot.
+\cite{b12}\\
+In a futuristic world it could become common to own a robotic dog that looks after a users house when he isn't home or asleep.
+If the password to this robot gets hacked or leaked, it could do lethal damage.
+For example robbing a place could get a lot easier with a simple brute force attack.
+\end{itemize}
+
 \section{State of the art}
-Passwords are used to protect sensitive information or access to hardware.\\
+Sensitive information or access to hardware must be protected in some way.
+Nowadays this is most likely to happen with a user name, which is normally easy to guess, and a password, which should be complex and secure.
+Also the data traffic should be encrypted, so nobody can read along and possibly gain access.\\
 For example users need to key in a user-name and password before they can access any network device over SSH.
 This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely.
+Over SSH the data traffic is encrypted and protected with a password.
 
-\subsection{The problem with passwords} \cite{b1} \cite{b2}
+\subsection{The problem with passwords} 
 As mentioned above a password protects sensitive information or access to hardware like a robot.
@@
 This is understandable, because as a study from NordPass shows, the average user has around a 100 accounts that require a unique password.
-Remembering a hundred unique passwords or more is really beyond a users capability.\\
-Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases.\\
-Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year: \cite{b3}
+Remembering a hundred unique passwords or more is really beyond a users capability.
+In a matter of that, users often use the same password for all programs and devices.
+For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots.
+If a user is using 150 different password, the hacker will only control one.\cite{b1} \cite{b2}\\
+Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
+Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
 \begin{enumerate}
@@
 
+But how quickly can a password really be hacked?
+The following graph shows that it is important to choose long passwords with different types of characters (such as numbers, upper-case and lower-case letters and symbols).
+But of course users should remember that names, common words etc.
+are still easily hackable with rainbow list attack.
+Users also shouldn't use personal information as a password.
+
+\begin{center}
+\includegraphics[scale=0.335]{./images/brute_force.png}
+\captionof{figure}[Graph for secure passwords - The figure is taken from the homepage TechRepublic (https://www.techrepublic.com/article/how-an-8-character-password-could-be-cracked-in-less-than-an-hour)]{How secure is a password?}
+\end{center}
+
+Now if a user is not really sure if his or her password is secure, he or she can use a website like PasswordMonster to check how long it would take a hacker to crack his or her password.\cite{b9}
+
 \subsection{Current state of art}
-\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\\
+In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
 2-factor-authentication adds an extra layer of security to passwords.
-First the user has to key in the username and password, which is the same as always.
-Then the user has to answer an additional question to verify it's really him/her.\\
-The second factor has different types of categories:
+First the user has to key in the user name and password, which is the same as always.
+Then the user has to pass an additional login-method to verify it's really him/her.\\
+The second factor is usually divided into these different types of categories:
 \begin{itemize}
-\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), ect.
+\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
 \item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
@@
 \end{itemize}
-A common type of the 2-factor-authentication is the SMS Text-Message.
+The most common type of the 2-factor-authentication is a hardware token (Something the user has.).
+A leading company in this market is yubico.
+They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC.
+Users that have a YubiKey should be aware that they have to carry it with them at all time.
+If they lose it is unfortunate.\cite{11}\\
+An other widely used method is the SMS Text-Message (Something the user knows.).
 After entering the password and user-name the user receives a unique one time passcode (OTP) via text message.
@@
 \subsection{Future state of art}
-\cite{b5} In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
+In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
 A passkey is similar to the 2-factor-authentication but without the first factor.
@@
 The app or website can use this mechanism instead of the traditional (and insecure) username and password.
+\cite{b5}
 
@@
 
+    % Anwendungen (Methods): Verwendung, Welche System, Wie Gut, Ausmaß, https://www.usenix.org/system/files/sec22-mayer.
 \end{itemize}
@@
 \begin{thebibliography}{00}
-\bibitem{b1} Proof for Research, URL:https://tech.co/password-managers/how-many-passwords-average-person
+\bibitem{b1} Proof for NordPass study, URL:https://tech.co/password-managers/how-many-passwords-average-person
 
@@
 
+\bibitem{b9} Link to PasswordMonster: https://www.passwordmonster.com/
+
+\bibitem{b10} Resource for harmless example: https://www.heise.de/hintergrund/Roomba-Wie-private-Fotos-eines-Staubsauger-Roboters-auf-Facebook-landen-koennen-7457283.html
+
+\bibitem{b11} Information for YubiKey: https://www.yubico.com/der-yubikey/?lang=de
+
+\bibitem{b12} Resourece for Future examples: https://www.bostondynamics.com/
+
 \end{thebibliography}

### 2023-03-21 Aktuelle_Version_Paper/passwords_cyber_security.tex
Changes from Signitzer were implemented andchanged

@@
 
-\title{The role of passwords in cybersecurity}
+\title{The role of passwords in cybersecurity and robotics}
 \author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
@@
 \begin{abstract}
-This document is part of the ECER 2023 conference on educational robotics. It follows the topic of cybersecurity and will focus on the role of passwords.
+This document will focus on an important topic of cybersecurity: Passwords!
+It will have a look on the indispensable role of passwords now and in the future.
+Additionally it will cover the role of passwords in robotics and will provide an best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
+
 \end{abstract}
-\vspace{1cm}
 
 \section{Introduction}
-This paper will focus on the wide use of passwords or passphrases in IT-security. It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness. \\
+This paper will concentrate on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
+It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems.
+\\
 It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
@@
 
+\section{Implications of weak passwords}
+In today's digital world many aspects of our daily live relies on remote services.
+Online banking, social media and smart home devices to name just a few of them.
+The majority of those services require the user to authenticate via username and password.
+A weak or re-used password could become a major security issue and if an account gets breached it can have devastating affects on the live of the person owning this account.\\
+But not only humans can be negatively affected.
+Many robots in today's world are controlled or maintained remotely.
+To regulate access, users will need a high quality password.
+Especially in robotics using a weak one can cause major problems.
+Users don't want anybody to have access to their robots, because they could for example, keep private information about them and their house.
+
+\subsection{How weak passwords can cause problems in robotics}
+
+\begin{itemize}
+\item \textbf{Robot vacuum cleaner:} A harmless example is the case of Mister Greg.
+The computer scientist took part in a beta program where he was gifted a robot vacuum cleaner with an additional camera for orientation.
+What he didn't know was that a picture of him, sitting on the toilet, would find it's way onto Facebook.
+\cite{b10}\\
+These days many people already got themselves a robot vacuum cleaner with a camera.
+In most cases it communicates with the docking station or a device of its owner (e.g.: smart phone) remotely.
+Because it's accessed remotely the connections have to be secured with a strong password.
+If not the user could undergo a similar experience as Mister Greg.
+\item \textbf{Military drones:} A extreme case of the danger of weak passwords is the military.
+Remote controlled spying and attack drones are already an important part there and will become even more significant in the future.
+The major problem is that these drones have to communicate with some kind of wireless radio signal.
+These signals are very vulnerable for hacking attacks.
+If somebody gains access to enemy robots, they can do a lot of damage.
+For example destroying military bases with the hacked drones.
+So even if it's not known for sure, there is a high probability that remote access to those drones will require a password at some point, and it should definitely be a strong and secure one.
+\item \textbf{In the future:} Boston Dynamics is a company which focuses on developing robots for the world of tomorrow.
+Robots are developing so fast that it is possible that in the near future everyone will have their own personal assistant robot.
+\cite{b12}\\
+In a futuristic world it could become common to own a robotic dog that looks after a users house when he isn't home or asleep.
+Or an robotic nanny which helps in the house and becomes part of the family.
+If the password to such a robot gets hacked or leaked, it could do all sorts of damage.
+For example spying on the family or robbing their home could get a lot easier, if one would have access to such a household assistant.
+\end{itemize}
+
 \section{State of the art}
-Passwords are used to protect sensitive information or access to hardware.\\
+Sensitive information or access to hardware must be protected in some way.
+Nowadays this is most likely to happen with a user name, which is normally easy to guess, and a password, which should be complex and secure.
+Also the data traffic should be encrypted, so nobody can read along and possibly gain access.\\
 For example users need to key in a user-name and password before they can access any network device over SSH.
 This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely.
+Over SSH the data traffic is encrypted and protected with a password.
 
-\subsection{The problem with passwords} \cite{b1} \cite{b2}
-As mentioned above a password protects sensitive information or access to hardware like a robot.
+\subsection{The problem with passwords} 
+As mentioned above, a password protects sensitive information or access to hardware like a robot.
 That's why it's so important to have a good, secure password.
 However users often are too lazy to create a strong password for every account they own.
-This is understandable, because as a study from NordPass shows, the average user has around a 100 accounts that require a unique password.
-Remembering a hundred unique passwords or more is really beyond a users capability.\\
-Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases.\\
-Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year: \cite{b3}
+This is understandable, because as often cited study from NordPass shows, the average user has around a 100 accounts that require a unique password.
+Remembering a hundred unique passwords or more is really beyond a users capability.
+In a matter of that, users often use the same password for all programs and devices.
+For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots.
+If a user is using 150 different password, the hacker will only control one.\cite{b1} \cite{b2}\\
+Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
+Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
 \begin{enumerate}
@@
 
+But how quickly can a password really be hacked?
+The following graph shows that it is important to choose long passphrases with different types of characters (such as numbers, upper-case and lower-case letters and symbols).
+But of course users should remember that names, common words etc.
+are still easily hackable with rainbow list attacks (attackers try brute force attacks not with generated passwords, but with lists of often used passwords).
+Users also shouldn't use personal information as a password.\\
+Figure 1 shows the time it takes to crack a password in a brute force attack depending on its length and the use of different character sets.
+
+\begin{center}
+\includegraphics[scale=0.335]{./images/brute_force.png}
+\captionof{figure}[Graph for secure passwords - The figure is taken from the homepage TechRepublic (https://www.techrepublic.com/article/how-an-8-character-password-could-be-cracked-in-less-than-an-hour)]{How secure is a password depending on its length?}
+\end{center}
+
+Now if a user is not really sure if his or her password is secure, he or she can use a website like PasswordMonster to check how long it would take a hacker to crack his or her password.\cite{b9}
+
+\begin{center}
+\includegraphics[scale=0.37]{./images/PasswordMonster.png}
+\captionof{figure}[Screen-shot of PasswordMonster example  - The screenshot is taken from the homepage PasswordMonster (https://www.passwordmonster.com/)]{Example in PasswordMonster}
+\end{center}
+
 \subsection{Current state of art}
-\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\\
+In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
 2-factor-authentication adds an extra layer of security to passwords.
-First the user has to key in the username and password, which is the same as always.
-Then the user has to answer an additional question to verify it's really him/her.\\
-The second factor has different types of categories:
+First the user has to key in the user name and password, which is the same as always.
+Then the user has to pass an additional login-method to verify it's really him or her.\\
+The second factor is usually divided into these different types of categories:
 \begin{itemize}
-\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), ect.
+\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
 \item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
@@
 \end{itemize}
-A common type of the 2-factor-authentication is the SMS Text-Message.
-After entering the password and user-name the user receives a unique one time passcode (OTP) via text message.
-They need to key in the passcode during a limited time (for example one minute).
+The most common type of the 2-factor-authentication is a hardware token.
+(Something the user has.) A leading company in this market is yubico.
+They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC.
+Users that have a YubiKey should be aware that they have to carry it with them at all time and be careful not to lose it.\cite{b11}\\
+An other widely used method is the SMS Text-Message.
+(Something the user knows.) After entering the password and user-name the user receives a unique one time passcode (OTP) via text message.
+They need to key in the passcode during a limited time (for example one minute).\\
 After both the first factor and second factor are correct the user gains access.
@@
 The passkey is hardware specific, which means the user always needs to carry the device (that the passkey is installed on) with them.
-If the user wants to use passkey on their laptop they need to verify their identity with their smartphone.
+If the user wants to use a passkey on their laptop they need to verify their identity with their smartphone.
 This eliminates phishing, credential stuffing and other remote attacks.
 \\
-In conclusion when a user is asked to sign-in to an app or website, he/she approves the sign-in with the same biometric or PIN that he/she uses to unlock the device (phone, computer or security key).
+In conclusion when a user is asked to sign-in to an app or website, he or she approves the sign-in with the same biometric or PIN that he or she uses to unlock the device (phone, computer or security key).
 The app or website can use this mechanism instead of the traditional (and insecure) username and password.
+\cite{b5}
 
@@
 \begin{itemize}
-\item \textbf{KeepassXC}\cite{b6}\\ 
+\item \textbf{KeepassXC}\\ 
 \begin{center}
@@
 The XC stands for Cross-Plattform. It is possible to use it on Linux, Windows, macOS and Android.\\
-The program is customizable. It allows the user to customize literally everything to their needs.\\
+The program is also customizable. It allows the user to customize literally everything to their needs.\cite{b6}\\
 
-\item \textbf{1Password}\cite{b7}\\
+\item \textbf{1Password}\\
 \begin{center}
@@
 The interface is very user-friendly. It even generates strong passwords for accounts that are newly created.
+\cite{b7}
 
 \end{itemize}
-\subsubsection{Benefits and drawbacks of password safe software}\cite{b8}
+\subsubsection{Benefits and drawbacks of password safe software}
 \textbf{Benefits}:
@@
 \end{itemize}
+Sources used to find the benefits and drawbacks: \cite{b8}
+\subsubsection{Scientific Studies regarding password managers}
+A very interesting scientific study by Peter Mayer et al.
+shows, that even for highly educated people, which definitely know about the risk of bad or reused passwords, the adoption of password manager software depends largely on how easy it is to use.
+Mayer and his colleagues show, that browser based password managers are more widely used than stand alone password managers, mainly because they are easier to use.\cite{b13}
 
@@
 
+\bibitem{b9} Link to PasswordMonster: https://www.passwordmonster.com/
+
+\bibitem{b10} Resource for harmless example: https://www.heise.de/hintergrund/Roomba-Wie-private-Fotos-eines-Staubsauger-Roboters-auf-Facebook-landen-koennen-7457283.html
+
+\bibitem{b11} Information for YubiKey: https://www.yubico.com/der-yubikey/?lang=de
+
+\bibitem{b12} Resource for Future examples: https://www.bostondynamics.com/
+
+\bibitem{b13} Why Users (Don’t) Use Password Managers at a Large Educational Institution: https://www.usenix.org/system/files/sec22-mayer.pdf
+
 \end{thebibliography}

### 2024-03-02 2023_paper/2te_Version_Paper/cyber_security_passwords.tex
moved 2023 files to subfolder and prepared for 2024

--- 2023-03-05 Aktuelle_Version_Paper/passwords_cyber_security.tex
@@
 \title{The role of passwords in cybersecurity}
+
+%\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
+%\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
+%\textit{HTL Anichstraße}\\
+%Innsbruck, Austria \\
+%lzobl@tsn.at}}
 \author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
@@
 Innsbruck, Austria \\
-nprantl@tsn.at }
+email address or ORCID}
 \and
-\IEEEauthorblockN{4\textsuperscript{th} Lukas Krahbichler}
+\IEEEauthorblockN{4\textsuperscript{th} Lukas}
 \IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
@@
 Innsbruck, Austria \\
-lkrahbichler@tsn.at}
+email address or ORCID}
+
 }
@@
 \section{State of the art}
-Passwords are used to protect sensitive information or access to hardware.\\
+Passwords are used to protect sensitive information or access to hardware.
 For example users need to key in a user-name and password before they can access any network device over SSH.
@@
 
-\subsection{The problem with passwords} \cite{b1} \cite{b2}
-As mentioned above a password protects sensitive information or access to hardware like a robot done.
+\subsection{The problem with passwords}
+As mentioned above a password protects sensitive information or access to hardware like a robot.
 That's why it's so important to have a good, secure password.
 However users often are too lazy to create a strong password for every account they own.
-This is understandable, because as a study from NordPass conducts, the average user has around a 100 accounts that require a unique password.Remembering a hundred unique Passwords or more is really beyond a users capability.\\
+But that's not the only reason.
+A study from NordPass conducts that the average user has around a 100 accounts that require a unique password.
+\\
 Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
-Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last years: \cite{b3}
+Every year long record gets published by different providers reviling the most used passwords. Here are the top ten used passwords of last years:
+\vspace{0.5cm}
 \begin{enumerate}
@@
 \end{enumerate}
+
 Also common password are locations, names, birthdays, ect.
@@
 \subsection{Current state of art}
-\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
+In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
 2-factor-authentication adds an extra layer of security to your password.
@@
 \item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
-\item \textbf{Something unique to the user.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
+\item \textbf{Something the user is.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
 \end{itemize}
@@
 \subsection{Future state of art}
-\cite{b5} In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
-A passkey is similar to the 2-factor-authentication but without the first factor.
-The second and only factor is the users device's security method such as a pin or a biometric sensor (fingerprint, face scan).
-\\
+In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
+A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the users device's security method such as a pin or a biometric
+sensor (fingerprint, face scan). \\
 The passkey is hardware specific, which means user always need to carry the device (that the passkey is installed on) with them.
@@
 \subsubsection{Explanation}
-A password safe allows you to save a list of your usernames and passwords for all your accounts.
-It has one master password or passphrase that unlocks the safe.
-Now users only have to recall one single master passphrase instead of a huge amount of passwords.
+A password safe allows you to save a list of your usenames linked to the associated password.
+It has one master password that locks the safe.
+Now user only have to recall one single password instead of a huge amount.
 \subsubsection{Examples}
 \begin{itemize}
-\item \textbf{KeepassXC}\cite{b6}\\ 
+\item \textbf{KeepassXC}\\
 This software is developed for users with extremely high demands of secure personal data management. \\
@@
 This prevents user's data from getting leaked when a server gets hacked or breached.\\
-The XC stands for Cross-Plattform. It is possible to use on it on Linux, Windows, macOS and Android.
 The program is customizable. It allows its user to customize literally everything to their needs.\\
-\includegraphics[scale=0.77]{./images/KeePass.png}\\
 
-\item \textbf{1Password}\cite{b7}\\
-This password safe is known for its easy to use interface and high security encryption.\\ 
+\item \textbf{1Password}\\
+This password save is known for its easy to use interface and high security encryption.\\
 1Password uses an uncommon encryption known as dual-key encryption.
@@
 The interface is very user-friendly. It even generates strong passwords for accounts that are newly created.
-\includegraphics[scale=0.077]{./images/1Password-Logo.wine.png}\\
+
 \end{itemize}
-\subsubsection{Benefits and drawbacks of password safe software}\cite{b8}
-\textbf{Benefits}:
+\subsubsection{Benefits and drawbacks}
+Benefits:
 \begin{itemize}
@@
 \end{itemize}
-\textbf{Drawbacks}:
+Drawbacks:
 \begin{itemize}
@@
 If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number).
-It only ever retrieves the data from storage then returns it.
-\vspace{0.5cm}
-\includegraphics[scale=0.32]{./images/haveibeenpwned.png}\\
+It only ever retrieves the data from storage then returns it.\\
 If you have already been pawned change that password. If not you can rest assured.
 \item Pick a password safe that fits best to your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password or better a master passphrase. Here is a easy way to do it: 
+\item Create a good master password. Here is a easy way to do it: 
 \begin{itemize}
-\item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
-\item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
+\item First make a sentence that easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
+\item The second step is to but the sentence in dialect or slang. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
 \item Thirdly you can change letters into numbers.
@@
 For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-\item The fifth step is to make all words start with a capital letter.
+\item The fourth step is to make all words start with a capital letter.
 This is only needed if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
 \end{itemize}
-In the end you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
+In the end you would end up with a password like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
 This is a really secure password because it has numbers, upper case and lower case letters, symbols and more that 8 characters.
@@
 
-\begin{thebibliography}{00}
-\bibitem{b1} Proof for Research, URL:https://tech.co/password-managers/how-many-passwords-average-person
-
-\bibitem{b2} Information for "The problem with passwords", URL:https://www.nomios.com/news-blog/password-problem/
-
-\bibitem{b3} List for most used passwords, URL:https://www.passwordmanager.com/most-common-passwords-latest-2022-statistics/
-
-\bibitem{b4} Information for "Current state of art", URL:\\ https://authy.com/what-is-2fa/
-
-\bibitem{b5} Information for "Future state of art", URL:\\https://fidoalliance.org/passkeys/\\ https://developer.apple.com/passkeys/\\ https://www.passkeys.io/\\ https://developers.google.com/identity/passkeys?hl=de
-
-\bibitem{b6} KeepassXC, Open source password manager, URL: https://keepassXC.org
-
-\bibitem{b7} 1Password, Paid password manager, URL: https://1password.com/de/
-
-\bibitem{b8} Information for "Benefits and drawbacks of password safe software", URL: \\
-https://expert.services/blog/managing-your-website/security/password-managers;\\ https://www.orangecountyscu.org/stories/pros-and-cons-of-using-a-password-manager/; \\ https://www.passwordboss.com/pros-and-cons-of-using-a-password-manager/; \\https://www.trustworthy.com/blog/pros-and-cons-password-manager/
-
-\bibitem{b9} HaveIbeenpawnd Website, Screenshot form Website, URL: https://haveibeenpwned.com/
-
-\bibitem{b10} 1Password Logo form Website, URL: https://www.logo.wine/logo/1Password
-
-\bibitem{b11} KeePass Logo form Website, URL: https://fosstorrents.com/softwares/keepassxc/
-
-\end{thebibliography}
-
 \end{document}

### 2024-03-02 2023_paper/Aktuelle_Version_Paper/passwords_cyber_security.tex
moved 2023 files to subfolder and prepared for 2024

--- 2023-03-21 Aktuelle_Version_Paper/passwords_cyber_security.tex
@@
 \begin{abstract}
-This document will focus on an important topic of cybersecurity: Passwords!
+This document will focus on an important topic of cybersecurity: Passwords.
 It will have a look on the indispensable role of passwords now and in the future.
-Additionally it will cover the role of passwords in robotics and will provide an best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
+Additionally it will cover the role of passwords in robotics and will provide a best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
 
@@
 \section{Introduction}
-This paper will concentrate on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
-It will focus on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems.
+The present paper concentrates on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
+It focuses on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems.
 \\
-It will cover state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
-Additionally it will try to highlight alternatives for effective password management and introduce a choice of selected password managing software. \\
+Further, it covers state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
+In addition the review highlights alternatives for effective password management and it introduces a choice of selected password managing software. \\
 Finally the paper includes a best practice guide to:
@@
 \section{Implications of weak passwords}
-In today's digital world many aspects of our daily live relies on remote services.
-Online banking, social media and smart home devices to name just a few of them.
-The majority of those services require the user to authenticate via username and password.
-A weak or re-used password could become a major security issue and if an account gets breached it can have devastating affects on the live of the person owning this account.\\
+In today's digital world many aspects of our daily lives rely on remote services, like online banking, social media and smart home devices to name just a few of them.
+The majority of those services require the user to authenticate him- or herself via username and password.
+A weak or re-used password could become a major security issue, if an account gets breached it can have devastating affects on the account owner's life.\\
 But not only humans can be negatively affected.
 Many robots in today's world are controlled or maintained remotely.
-To regulate access, users will need a high quality password.
-Especially in robotics using a weak one can cause major problems.
-Users don't want anybody to have access to their robots, because they could for example, keep private information about them and their house.
+To regulate access, users will need a high security password.
+Especially in robotics, using a weak one can cause major problems, because users don't want anybody to have access to their robots.
 
@@
 \begin{itemize}
-\item \textbf{Robot vacuum cleaner:} A harmless example is the case of Mister Greg.
-The computer scientist took part in a beta program where he was gifted a robot vacuum cleaner with an additional camera for orientation.
-What he didn't know was that a picture of him, sitting on the toilet, would find it's way onto Facebook.
+\item \textbf{Robot vacuum cleaner:} A harmless example to point out is the case of Mister Greg.
+The computer scientist took part in a program where he was gifted a robot vacuum cleaner with an additional camera for orientation.
+What he didn't know was that a picture of him, sitting on the toilet, would somehow find it's way onto Facebook.
 \cite{b10}\\
-These days many people already got themselves a robot vacuum cleaner with a camera.
+These days, many people already got themselves a robot vacuum cleaner with a camera.
 In most cases it communicates with the docking station or a device of its owner (e.g.: smart phone) remotely.
@@
 If not the user could undergo a similar experience as Mister Greg.
-\item \textbf{Military drones:} A extreme case of the danger of weak passwords is the military.
+\item \textbf{Military drones:} High risk or even danger of weak passwords lies in the area of military applications.
 Remote controlled spying and attack drones are already an important part there and will become even more significant in the future.
@@
 These signals are very vulnerable for hacking attacks.
-If somebody gains access to enemy robots, they can do a lot of damage.
-For example destroying military bases with the hacked drones.
+If somebody gains access to enemy robots, they can do a lot of damage, like destroying military bases.
 So even if it's not known for sure, there is a high probability that remote access to those drones will require a password at some point, and it should definitely be a strong and secure one.
 \item \textbf{In the future:} Boston Dynamics is a company which focuses on developing robots for the world of tomorrow.
-Robots are developing so fast that it is possible that in the near future everyone will have their own personal assistant robot.
+Robots are developing so fast that it is possible that in the near future many people will have an own personal assistant robot.
 \cite{b12}\\
-In a futuristic world it could become common to own a robotic dog that looks after a users house when he isn't home or asleep.
-Or an robotic nanny which helps in the house and becomes part of the family.
+In a futuristic world it could become common to own a robotic dog that looks after a user's house when they aren't home or asleep.
+Or a robotic nanny which helps in the house and becomes part of the family.
 If the password to such a robot gets hacked or leaked, it could do all sorts of damage.
@@
 Sensitive information or access to hardware must be protected in some way.
-Nowadays this is most likely to happen with a user name, which is normally easy to guess, and a password, which should be complex and secure.
-Also the data traffic should be encrypted, so nobody can read along and possibly gain access.\\
-For example users need to key in a user-name and password before they can access any network device over SSH.
+Nowadays this is likely to happen with a username, which is normally easy to guess, and a password, which should be complex and secure.
+Also the data traffic should be encrypted, so nobody can read along and possibly gain access to it.\\
+For example users need to key in a username and password before they can access any network device over SSH.
 This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely.
@@
 However users often are too lazy to create a strong password for every account they own.
-This is understandable, because as often cited study from NordPass shows, the average user has around a 100 accounts that require a unique password.
+This is understandable, because as often cited study from NordPass shows, the average user has around 100 accounts that require a unique password.
 Remembering a hundred unique passwords or more is really beyond a users capability.
@@
 For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots.
-If a user is using 150 different password, the hacker will only control one.\cite{b1} \cite{b2}\\
-Yes, the main problem is laziness but there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
-Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
+If a user is using 150 different password, a successful hacker will just be able to control one robot by one hack.\cite{b1} \cite{b2}\\
+Beside laziness as the main cause there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
+Every year long records get published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
 \begin{enumerate}
@@
 \end{enumerate}
-Other common password are locations, names, birthdays, etc.
+Other common passwords are locations, names, birthdays, etc.
 
@@
 The following graph shows that it is important to choose long passphrases with different types of characters (such as numbers, upper-case and lower-case letters and symbols).
-But of course users should remember that names, common words etc.
-are still easily hackable with rainbow list attacks (attackers try brute force attacks not with generated passwords, but with lists of often used passwords).
+The user should always consider that names, common words etc.
+are still easily hackable with a rainbow list attack (brute force attacks with lists of often used passwords).
 Users also shouldn't use personal information as a password.\\
@@
 
-\subsection{Current state of art}
+\subsection{Current state of the art}
 In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
 2-factor-authentication adds an extra layer of security to passwords.
-First the user has to key in the user name and password, which is the same as always.
+First the user has to fill in the username and the password like always.
 Then the user has to pass an additional login-method to verify it's really him or her.\\
-The second factor is usually divided into these different types of categories:
+The second factor is usually divided into one of the following categories:
 \begin{itemize}
-\item \textbf{Something the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
-\item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
-\item \textbf{Something unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print. 
+\item \textbf{Something only the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
+\item \textbf{Something only the user possesses.} This could be a small hardware token or smart-phone.
+\item \textbf{Something physical unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print. 
 \end{itemize}
-The most common type of the 2-factor-authentication is a hardware token.
-(Something the user has.) A leading company in this market is yubico.
+The most common type of the 2-factor-authentication is a smartphone.
+Another good - but less common - option is a hardware key.
+(Something only the user possesses.) A leading company in this market is yubico.
 They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC.
-Users that have a YubiKey should be aware that they have to carry it with them at all time and be careful not to lose it.\cite{b11}\\
-An other widely used method is the SMS Text-Message.
-(Something the user knows.) After entering the password and user-name the user receives a unique one time passcode (OTP) via text message.
-They need to key in the passcode during a limited time (for example one minute).\\
-After both the first factor and second factor are correct the user gains access.
+Users that have a YubiKey should be aware that they have to carry it with them at all times and be careful not to lose it.\cite{b11}\\
+Another widely used method is the SMS Text-Message.
+(Something only the user knows.) After entering the password and username the user receives a unique one time passcode (OTP) via text message.
+The user needs to key in the passcode during a limited time (for example one minute).\\
+After both the first factor and the second factor are correct the user gains access.
 
-\subsection{Future state of art}
-In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
+\subsection{Future state of the art}
+In the future passwords will likely be entirely replaced by passkeys because they are easier to use, securer, faster and still work on almost every device.\\
 A passkey is similar to the 2-factor-authentication but without the first factor.
-The second and only factor is the users device's security method such as a pin or a biometric sensor (fingerprint, face scan).
+The second and only factor is the user's device's security method such as a PIN or a biometric sensor (fingerprint, face scan).
 \\
-The passkey is hardware specific, which means the user always needs to carry the device (that the passkey is installed on) with them.
-If the user wants to use a passkey on their laptop they need to verify their identity with their smartphone.
+The passkey is hardware specific, which means the user always needs to carry a second device (that the passkey is installed on).
+If the user wants to use a passkey on the laptop he or she needs to verify the identity with an additional smartphone.
 This eliminates phishing, credential stuffing and other remote attacks.
@@
 \subsection{Solutions to password managing}
-If users don't want to or can't use passkeys there are different options how they could protect their sensitive information.
-Users can write their passwords with account info into a booklet.
-The passwords would be unique and strong.
+If users don't want to or are not able to use passkeys there are different options how they could protect their sensitive information.
+Users can write their passwords with their account information into a booklet.
+Thus, the passwords can be unique and strong.
 It's a really simple idea but also has a lot of drawbacks.
@@
 \begin{itemize}
-\item Users need to carry the booklet with them everywhere. If not they can't always access their accounts.
-\item Users can lose the booklet. If that happens they can't access any of their accounts any more.
+\item Users need to always carry the booklet with them. Without it they are not able to access their accounts.
+\item Users can lose the booklet. Depending on the security restrictions they can't gain access to some of their accounts anymore.
 \item It's not a really clear way of organizing passwords.
-Once something is written down user can't change the position in the booklet.
+Once something is written down it cannot change the position in the booklet.
 So users would have to search for each password for indefinite time before they find it.
@@
 \subsubsection{Explanation}
-A password safe allows you to save a list of your user-names and passwords for all your accounts.
+A password safe, also known as password manager, is a piece of sofware that allows saving a list of usernames and passwords for all accounts.
 It has one master password or passphrase that unlocks the safe.
@@
 \end{center}
-This password safe is known for its easy to use interface and high security encryption.\\ 
+This password safe is known for its easy to use interface and its high security encryption.\\ 
 1Password uses an uncommon encryption known as dual-key encryption.
-If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of its two keys.
+If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of the two required keys.
 The first key is the user's master-key.
@@
 The secret key is generated on every device the user logs into.
-It will only be saved on the user's devices and never saved together with the other pins.\\
-Depending on what account type users choose their data is stored differently.
-However in only one version users have the option to save their data locally.
+It will only be saved on the user's devices and will never be saved together with the second PINs.\\
+Depending on the account type the data is stored differently, but there is only one account type saving the users' data locally.
 In every other option the user's data gets saved in a cloud-based vault.\\
-The interface is very user-friendly. It even generates strong passwords for accounts that are newly created.
+The interface is known as user-friendly and generates strong passwords for accounts that are newly created.
 \cite{b7}
@@
 This makes it extremely difficult for hackers to steal the passwords and gain access to the user's online accounts.
-\item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without having to remember multiple passwords.
-\item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the user-name and password for the user, saving them time and reducing the risk of typing errors.
-\item \textbf{Multi-device access:} Many password safes allow user to access their passwords from multiple devices, including smartphones, tablets, and computers.
+\item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without remembering multiple passwords.
+\item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the username and the password of the user, saving time and reducing the risk of typing errors.
+\item \textbf{Multi-device access:} Many password safes allow the user to access their passwords from multiple devices, including smartphones, tablets, and computers.
 \item \textbf{Password strength:} Password safes often include a password generator feature, which creates strong passwords. This helps to ensure that the passwords are not easily guessable.
@@
 It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
-\item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if a user is not able to access their password safe for some reason.
-\item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to using them effectively.
-\item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if user are looking for a free solution.
+\item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if the user is not able to access their password safe for some reason.
+\item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to the application.
+\item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if the user is looking for a free solution.
 \item \textbf{Compatibility:} Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
@@
 \subsubsection{Scientific Studies regarding password managers}
-A very interesting scientific study by Peter Mayer et al.
+A thoughtful scientific study by Peter Mayer et al.
 shows, that even for highly educated people, which definitely know about the risk of bad or reused passwords, the adoption of password manager software depends largely on how easy it is to use.
-Mayer and his colleagues show, that browser based password managers are more widely used than stand alone password managers, mainly because they are easier to use.\cite{b13}
+Mayer and his colleagues show, that browser based password managers are more widely used than standalone password saves, mainly because they are easier to use.\cite{b13}
 
-\section{Concept and implementation for best practise}
-It can be very overwhelming to start with a secure option to manage passwords. So here is a best practise guide everyone can follow.
+\section{Concept and implementation for best practice}
+It can be very overwhelming to start with a secure option to manage passwords. So here is a best practice guide everyone can follow.
 \begin{enumerate}
-\item First you should check if your password has already been breached. There is a website that let's you easily check just that. It's called \textbf{"haveibeenpwned.com"}.\\
+\item First you should check if your password has already been breached. There is a website that makes it easy to check it out. It's called \textbf{"haveibeenpwned.com"}.\\
 Haveibeenpwned is open source and secure.
 If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number).
-It only ever retrieves the data from storage then returns it.
+It only ever retrieves the data from storage then returns the data.
 \vspace{0.5cm}
@@
 \end{center}
-If you have already been pawned, change that password. If not you can rest assured. 
+If you have already been pawned, change the present password. If not you can rest assured. 
 \item Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password or even better a master passphrase. Here is a easy way to do it: 
+\item Create a good master password or even better a master passphrase. Here is a easy way to to design it: 
 \begin{itemize}
@@
 \item The fifth step is to make all words start with a capital letter.
-This is only needed if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
+This is only useful if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
 \end{itemize}
-In the end you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-This is a really secure passphrase because it has numbers, upper case and lower case letters, symbols and more that 8 characters.
+Finally you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
+This is a really secure passphrase because it compromises numbers, upper case and lower case letters, symbols and more than 8 characters.
 This would make a strong master password that is nearly impossible to crack.
@@
 \section{Conclusion}
-Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable to attacks.
+Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable when attacked.
 While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
-As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and especially creative approaches to authentication and access control, that are both secure and user-friendly.
+As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and equally trustworthy approaches to authentication and access control, that are both secure and user-friendly.
 
@@
 
-\bibitem{b12} Resource for Future examples: https://www.bostondynamics.com/
+\bibitem{b12} Resource for "In the future": https://www.bostondynamics.com/
 

### 2024-03-02 2023_paper/Aktuelle_Version_Paper/passwords_cyber_security_markus.tex
moved 2023 files to subfolder and prepared for 2024

--- 2023-03-21 Aktuelle_Version_Paper/passwords_cyber_security.tex
@@
 
-\title{The role of passwords in cybersecurity and robotics}
+\title{The role of passwords in cybersecurity}
 \author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
@@
 \begin{center}
-\includegraphics[scale=0.37]{./images/PasswordMonster.png}
-\captionof{figure}[Screen-shot of PasswordMonster example  - The screenshot is taken from the homepage PasswordMonster (https://www.passwordmonster.com/)]{Example in PasswordMonster}
+\includegraphics[scale=0.38]{./images/PasswordMonster.png}
+\captionof{figure}[PasswordMonster - The figure is taken from the homepage of PasswordMonster (https://www.passwordmonster.com/) {A PasswordMonster example}
 \end{center}

### 2024-03-02 2023_paper/Verbesserungen_Signitzer/Aktuelle_Version_Paper/passwords_cyber_security.tex
moved 2023 files to subfolder and prepared for 2024

--- 2024-03-02 2023_paper/2te_Version_Paper/cyber_security_passwords.tex
@@
 \title{The role of passwords in cybersecurity}
-
-%\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
-%\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
-%\textit{HTL Anichstraße}\\
-%Innsbruck, Austria \\
-%lzobl@tsn.at}}
 \author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
@@
 email address or ORCID}
-
 }
@@
 \section{State of the art}
-Passwords are used to protect sensitive information or access to hardware.
+Passwords are used to protect sensitive information or access to hardware.\\
  For example users need to key in a user-name and password before they can access any network device over SSH.
@@
 
-\subsection{The problem with passwords}
-As mentioned above a password protects sensitive information or access to hardware like a robot.
+\subsection{The problem with passwords} \cite{b1} \cite{b2}
+As mentioned above a password protects sensitive information or access to hardware like a robot done.
 That's why it's so important to have a good, secure password.
 However users often are too lazy to create a strong password for every account they own.
-But that's not the only reason.
-A study from NordPass conducts that the average user has around a 100 accounts that require a unique password.
-\\
+This is understandable, because as a study from NordPass conducts, the average user has around a 100 accounts that require a unique password.Remembering a hundred unique Passwords or more is really beyond a users capability.\\
 Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
-Every year long record gets published by different providers reviling the most used passwords. Here are the top ten used passwords of last years:
-\vspace{0.5cm}
+Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last years: \cite{b3}
 \begin{enumerate}
@@
 \end{enumerate}
-
 Also common password are locations, names, birthdays, ect.
@@
 \subsection{Current state of art}
-In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
+\cite{b4}In recent years there was a massive increase in cybercrime. To prevent data-leaks companies started integrating 2-factor-authentication.\\
 2-factor-authentication adds an extra layer of security to your password.
@@
 \item \textbf{Something the user has.} This could be a small hardware token or smart-phone.
-\item \textbf{Something the user is.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
+\item \textbf{Something unique to the user.} This might include a  biometric pattern of a fingerprint, an iris scan, or a voice print. 
 \end{itemize}
@@
 \subsection{Future state of art}
-In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
-A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the users device's security method such as a pin or a biometric
-sensor (fingerprint, face scan). \\
+\cite{b5} In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
+A passkey is similar to the 2-factor-authentication but without the first factor.
+The second and only factor is the users device's security method such as a pin or a biometric sensor (fingerprint, face scan).
+\\
 The passkey is hardware specific, which means user always need to carry the device (that the passkey is installed on) with them.
@@
 \subsubsection{Explanation}
-A password safe allows you to save a list of your usenames linked to the associated password.
-It has one master password that locks the safe.
-Now user only have to recall one single password instead of a huge amount.
+A password safe allows you to save a list of your usernames and passwords for all your accounts.
+It has one master password or passphrase that unlocks the safe.
+Now users only have to recall one single master passphrase instead of a huge amount of passwords.
 \subsubsection{Examples}
 \begin{itemize}
-\item \textbf{KeepassXC}\\
+\item \textbf{KeepassXC}\cite{b6}\\ 
+\begin{center}
+\includegraphics[scale=0.6]{./images/KeePass.png}
+\captionof{figure}[KeePassXC - The figure is taken from the homepage of the KeePassXC Project (https://keepassxc.org/)]{KeePAssXC Logo}
+\end{center}
+
 This software is developed for users with extremely high demands of secure personal data management. \\
@@
 This prevents user's data from getting leaked when a server gets hacked or breached.\\
+The XC stands for Cross-Plattform. It is possible to use on it on Linux, Windows, macOS and Android.
 The program is customizable. It allows its user to customize literally everything to their needs.\\
 
-\item \textbf{1Password}\\
-This password save is known for its easy to use interface and high security encryption.\\
+\item \textbf{1Password}\cite{b7}\\
+\begin{center}
+\includegraphics[scale=0.06]{./images/1Password-Logo.wine.png}
+\captionof{figure}[1Password - The figure is taken from the homepage of the 1Password Project (https://1password.com/)]{1Password Logo}
+\end{center}
+This password safe is known for its easy to use interface and high security encryption.\\ 
 1Password uses an uncommon encryption known as dual-key encryption.
@@
 \end{itemize}
-\subsubsection{Benefits and drawbacks}
-Benefits:
+\subsubsection{Benefits and drawbacks of password safe software}\cite{b8}
+\textbf{Benefits}:
 \begin{itemize}
@@
 \end{itemize}
-Drawbacks:
+\textbf{Drawbacks}:
 \begin{itemize}
@@
 If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number).
-It only ever retrieves the data from storage then returns it.\\
+It only ever retrieves the data from storage then returns it.
+\vspace{0.5cm}
+\begin{center}
+\includegraphics[scale=0.29]{./images/haveibeenpwned.png}
+\captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{Have I been pwned? Homepage}
+\end{center}
 If you have already been pawned change that password. If not you can rest assured. 
 \item Pick a password safe that fits best to your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password. Here is a easy way to do it: 
+\item Create a good master password or better a master passphrase. Here is a easy way to do it: 
 \begin{itemize}
-\item First make a sentence that easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
-\item The second step is to but the sentence in dialect or slang. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
+\item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
+\item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
 \item Thirdly you can change letters into numbers.
@@
 For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-\item The fourth step is to make all words start with a capital letter.
+\item The fifth step is to make all words start with a capital letter.
 This is only needed if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
 \end{itemize}
-In the end you would end up with a password like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
+In the end you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
 This is a really secure password because it has numbers, upper case and lower case letters, symbols and more that 8 characters.
@@
 
+%Abbildungsverzeichnis erstellen
+\listoffigures
+
+\begin{thebibliography}{00}
+\bibitem{b1} Proof for Research, URL:https://tech.co/password-managers/how-many-passwords-average-person
+
+\bibitem{b2} Information for "The problem with passwords", URL:https://www.nomios.com/news-blog/password-problem/
+
+\bibitem{b3} List for most used passwords, URL:https://www.passwordmanager.com/most-common-passwords-latest-2022-statistics/
+
+\bibitem{b4} Information for "Current state of art", URL:\\ https://authy.com/what-is-2fa/
+
+\bibitem{b5} Information for "Future state of art", URL:\\https://fidoalliance.org/passkeys/\\ https://developer.apple.com/passkeys/\\ https://www.passkeys.io/\\ https://developers.google.com/identity/passkeys?hl=de
+
+\bibitem{b6} KeepassXC, Open source password manager, URL: https://keepassXC.org
+
+\bibitem{b7} 1Password, Paid password manager, URL: https://1password.com/de/
+
+\bibitem{b8} Information for "Benefits and drawbacks of password safe software", URL: \\
+https://expert.services/blog/managing-your-website/security/password-managers;\\ https://www.orangecountyscu.org/stories/pros-and-cons-of-using-a-password-manager/; \\ https://www.passwordboss.com/pros-and-cons-of-using-a-password-manager/; \\https://www.trustworthy.com/blog/pros-and-cons-password-manager/
+
+\bibitem{b9} HaveIbeenpawnd Website, Screenshot form Website, URL: https://haveibeenpwned.com/
+
+\bibitem{b10} 1Password Logo form Website, URL: https://www.logo.wine/logo/1Password
+
+\bibitem{b11} KeePass Logo form Website, URL: https://fosstorrents.com/softwares/keepassxc/
+
+\end{thebibliography}
+
 \end{document}

### 2024-03-02 2023_paper/Verbesserungen_Signitzer/aenderungen_in_fett.tex
moved 2023 files to subfolder and prepared for 2024

--- 2024-03-02 2023_paper/2te_Version_Paper/cyber_security_passwords.tex
@@
 
-%\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
-%\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
-%\textit{HTL Anichstraße}\\
-%Innsbruck, Austria \\
-%lzobl@tsn.at}}
 \author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
+\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße } \\
 \textit{HTL Anichstraße}\\
 Innsbruck, Austria \\
-lzobl@tsn.at}
-\and
-\IEEEauthorblockN{2\textsuperscript{nd} Matteo Reiter}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-mareiter@tsn.at}
-\and
-\IEEEauthorblockN{3\textsuperscript{rd} Niclas Prantl}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{4\textsuperscript{th} Lukas}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-email address or ORCID}
+lzobl@tsn.at}}
 
-}
 \maketitle
@@
 \end{abstract}
-\vspace{1cm}
+\vspace{2cm}
+\begin{IEEEkeywords}
+component, formatting, style, styling, insert
+\end{IEEEkeywords}
 
@@
 \section{State of the art}
-Passwords are used to protect sensitive information or access to hardware.
+Passwords are used to protect sensitive information.\\
+\textbf{Passwords are used to protect sensitive information or access to hardware. done}\\
  For example users need to key in a user-name and password before they can access any network device over SSH.
@@
 \subsection{The problem with passwords}
-As mentioned above a password protects sensitive information or access to hardware like a robot.
+As mentioned earlier/\textbf{above done} a password protects sensitive information \textbf{or access to hardware like a robot done}.
 That's why it's so important to have a good, secure password.
 However users often are too lazy to create a strong password for every account they own.
-But that's not the only reason.
-A study from NordPass conducts that the average user has around a 100 accounts that require a unique password.
-\\
+\textbf{This is understandable, because as a} study from NordPass conducts, the average user has around a 100 accounts that require a unique password.\textbf{Remembering a hundred unique Passwords or more is really beyond a users capability.}\\
 Yes, the main problem is laziness but there are many other reasons why users tend to create weak pins.\\
-Every year long record gets published by different providers reviling the most used passwords. Here are the top ten used passwords of last years:
+Every year long records gets published by different providers revealing the most used passwords. Here are the top ten used passwords of last years:
+\textbf{Lilo - Quelle angeben!!!)}
 \vspace{0.5cm}
@@
 In the future passwords will likely be replaced by passkeys because they are easier to use, securer, faster and work on almost every device.\\
-A passkey is similar to the 2-factor-authentication but without the first factor. The second and only factor is the users device's security method such as a pin or a biometric
-sensor (fingerprint, face scan). \\
+A passkey is similar to the 2-factor-authentication but without the first factor.
+The second and only factor is the users device's security method such as a pin or a biometric sensor (fingerprint, face scan).
+\\
 The passkey is hardware specific, which means user always need to carry the device (that the passkey is installed on) with them.
@@
 \subsubsection{Explanation}
-A password safe allows you to save a list of your usenames linked to the associated password.
-It has one master password that locks the safe.
-Now user only have to recall one single password instead of a huge amount.
+A password safe allows you to save a list of your usernames \textbf{and passwords for all your accounts.} It has one master password \textbf{or passphrase} that unlocks the safe.
+Now users only have to recall one single master passphrase instead of a huge amount of passwords.
 \subsubsection{Examples}
 \begin{itemize}
-\item \textbf{KeepassXC}\\
+\item \textbf{KeepassXC}\cite{b1}\\ \textbf{Lilo - Link angeben}\\
+URL: https://keepassxc.org/\\
 This software is developed for users with extremely high demands of secure personal data management. \\
@@
 This prevents user's data from getting leaked when a server gets hacked or breached.\\
+\textbf{The XC stands for Cross-Plattform. It is possible to use on it on Linux, Windows, macOS and Android.}
 The program is customizable. It allows its user to customize literally everything to their needs.\\
 
-\item \textbf{1Password}\\
-This password save is known for its easy to use interface and high security encryption.\\
+\item \textbf{1Password}\\ \textbf{Lilo - Link angeben}\\
+This password safe is known for its easy to use interface and high security encryption.\\ \textbf{Lili - erwähnen, das die Daten online gespeichert werden, im vergleich zu KeepassXC.}\\
 1Password uses an uncommon encryption known as dual-key encryption.
@@
 \end{itemize}
-\subsubsection{Benefits and drawbacks}
-Benefits:
+\subsubsection{Benefits and drawbacks of password safe software}
+\textbf{Benefits}:
 \begin{itemize}
@@
 \item Pick a password safe that fits best to your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password. Here is a easy way to do it: 
+\item Create a good master password \textbf{or better a master passphrase}. Here is a easy way to do it: 
 \begin{itemize}
-\item First make a sentence that easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
-\item The second step is to but the sentence in dialect or slang. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
+\item \textbf{First think of something which is easy to remember}. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
+\item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
 \item Thirdly you can change letters into numbers.
@@
 
+\begin{thebibliography}{00}
+\bibitem{b1} KeepassXC, Open source password manager,  URL: https://keepassXC.org
+\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
+\end{thebibliography}
+
 \end{document}

### 2024-03-02 2024_paper/paper/msglink_framework.tex
restructured and added info

--- 2024-03-02 2023_paper/Aktuelle_Version_Paper/passwords_cyber_security.tex
@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
 \begin{document}
 
-\title{The role of passwords in cybersecurity and robotics}
-\author{\IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
+\title{Programmer friendly communication framework}
+
+\author{
+    \IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
+    \IEEEauthorblockA{
+        \textit{Botball Team FrenchBakery} \\
         \textit{HTL Anichstraße}\\
         Innsbruck, Austria \\
-lzobl@tsn.at}
+        lzobl@tsn.at
+    }
     \and
-\IEEEauthorblockN{2\textsuperscript{th} Lukas Krahbichler}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
+    \IEEEauthorblockN{2\textsuperscript{nd} Matteo Reiter}
+    \IEEEauthorblockA{
+        \textit{Botball Team FrenchBakery} \\
         \textit{HTL Anichstraße}\\
         Innsbruck, Austria \\
-lkrahbichler@tsn.at}
-\and
-\IEEEauthorblockN{3\textsuperscript{nd} Matteo Reiter}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-mareiter@tsn.at}
-\and
-\IEEEauthorblockN{4\textsuperscript{rd} Niclas Prantl}
-\IEEEauthorblockA{\textit{Botball Team HTL Anichstraße} \\
-\textit{HTL Anichstraße}\\
-Innsbruck, Austria \\
-nprantl@tsn.at }
+        mareiter@tsn.at
     }
+}
+
 \maketitle
@@
 Additionally it will cover the role of passwords in robotics and will provide a best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
-
 \end{abstract}

### 2024-03-02 2024_paper/paper/msglink_framework.tex
added formatter info
cleaned up style of document

@@
 \begin{abstract}
-This document will focus on an important topic of cybersecurity: Passwords.
-It will have a look on the indispensable role of passwords now and in the future.
-Additionally it will cover the role of passwords in robotics and will provide a best practice guide for working with password safes and how to create a secure and easy to remember master passphrase.
+    This document will focus on an important topic in software development: Communication.
+It will provide an insight into why what communication is needed for and how communication between computers is typically realized.
+Additionally it will cover problems and annoyances with currently available communication frameworks for certain applications as well as introduce concepts and a possible solution for the before-mentioned.
 \end{abstract}
@@
 \section{Introduction}
+
 The present paper concentrates on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
@@
 \section{Implications of weak passwords}
+
 In today's digital world many aspects of our daily lives rely on remote services, like online banking, social media and smart home devices to name just a few of them.
@@
 \section{State of the art}
+
 Sensitive information or access to hardware must be protected in some way.
@@
 \subsection{The problem with passwords}
+
 As mentioned above, a password protects sensitive information or access to hardware like a robot.
@@
 \subsection{Current state of the art}
+
 In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
@@
 \subsection{Future state of the art}
+
 In the future passwords will likely be entirely replaced by passkeys because they are easier to use, securer, faster and still work on almost every device.\\
@@
 \subsection{Solutions to password managing}
+
 If users don't want to or are not able to use passkeys there are different options how they could protect their sensitive information.
@@
 \subsection{Password Safe}
+
 \subsubsection{Explanation}
+
 A password safe, also known as password manager, is a piece of sofware that allows saving a list of usernames and passwords for all accounts.
@@
 Now users only have to recall one single master passphrase instead of a huge amount of passwords.
+
 \subsubsection{Examples}
+
 \begin{itemize}
-\item \textbf{KeepassXC}\\ 
+    \item \textbf{KeepassXC} \\
         \begin{center}
@@
 
-\item \textbf{1Password}\\
+    \item \textbf{1Password} \\
         \begin{center}
@@
         \end{center}
-This password safe is known for its easy to use interface and its high security encryption.\\ 
+        This password safe is known for its easy to use interface and its high security encryption. \\
         1Password uses an uncommon encryption known as dual-key encryption.
@@
         \cite{b7}
+\end{itemize}
 
-\end{itemize}
 \subsubsection{Benefits and drawbacks of password safe software}
+
 \textbf{Benefits}:
@@
 \end{itemize}
+
 \textbf{Drawbacks}:
@@
 \end{itemize}
+
 Sources used to find the benefits and drawbacks: \cite{b8}
+
 \subsubsection{Scientific Studies regarding password managers}
+
 A thoughtful scientific study by Peter Mayer et al.
@@
 \section{Concept and implementation for best practice}
+
 It can be very overwhelming to start with a secure option to manage passwords. So here is a best practice guide everyone can follow.
+
 \begin{enumerate}
-\item First you should check if your password has already been breached. There is a website that makes it easy to check it out. It's called \textbf{"haveibeenpwned.com"}.\\
+    \item 
+        First you should check if your password has already been breached.
+There is a website that makes it easy to check it out.
+It's called \textbf{"haveibeenpwned.com"}.
 Haveibeenpwned is open source and secure.
@@
 It only ever retrieves the data from storage then returns the data.
-\vspace{0.5cm}
         \begin{center}
             \includegraphics[scale=0.29]{./images/haveibeenpwned.png}
-\captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{Have I been pwned? Homepage}
+            \captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{"Have I been pwned?" Homepage}
         \end{center}
         If you have already been pawned, change the present password. If not you can rest assured.
-\item Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
-\item Create a good master password or even better a master passphrase. Here is a easy way to to design it: 
+    \item 
+        Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
+    \item 
+        Create a good master password or even better a master passphrase. Here is a easy way to to design it:
         \begin{itemize}
@@
 \section{Conclusion}
+
 Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable when attacked.
@@
 
-%Abbildungsverzeichnis erstellen
+% Abbildungsverzeichnis
 \listoffigures

### 2024-03-03 2024_paper/paper/msglink_framework.tex
worked on structure
Co-authored-by: LiloZz <LiloZz@users.noreply.github.com>

@@
 \begin{abstract}
-    This document will focus on an important topic in software development: Communication.
-It will provide an insight into why what communication is needed for and how communication between computers is typically realized.
+    This document will focus on an important topic in software development: Network communication.
+It will provide an insight into why what network communication is needed for and how it is typically realized.
 Additionally it will cover problems and annoyances with currently available communication frameworks for certain applications as well as introduce concepts and a possible solution for the before-mentioned.
@@
 
-The present paper concentrates on the wide use of passwords or passphrases in IT-security, with a special consideration of robotics.
-It focuses on the security aspect of passwords, especially on weak passwords due to user (human) laziness, which in some cases can lead to major problems.
-\\
-Further, it covers state of the art password technologies (2 factor authentication with passkeys) and possible future technologies.\\
-In addition the review highlights alternatives for effective password management and it introduces a choice of selected password managing software. \\
-Finally the paper includes a best practice guide to:
+When it comes to computer 
+... Why do we need communication? What for?
+
+IP Networks erwähnen
+
+TCP/UDP Erwähnen und unterschied (Stream vs. Datagram).
+
+wir behandeln hier nur direkte punkt-zu-punkt verbindungen
+
+\subsection{Communication frameworks}
+
+... Why do we need communication frameworks?
+
+using TCP/UDP directly is tedious for many modern application, ...
+
+Job of cf/library?
+
 \begin{itemize}
-    \item check if user's passwords have already been breached
-    \item create an easy to remember but secure master password for a password managing software
+    \item abstrahieren
+    \item benutzerfreundlichkeit bzw programmiererfreundlichkeit
+    \item added features for specialized use cases
 \end{itemize}
 
-\section{Implications of weak passwords}
-
-In today's digital world many aspects of our daily lives rely on remote services, like online banking, social media and smart home devices to name just a few of them.
-The majority of those services require the user to authenticate him- or herself via username and password.
-A weak or re-used password could become a major security issue, if an account gets breached it can have devastating affects on the account owner's life.\\
-But not only humans can be negatively affected.
-Many robots in today's world are controlled or maintained remotely.
-To regulate access, users will need a high security password.
-Especially in robotics, using a weak one can cause major problems, because users don't want anybody to have access to their robots.
-
-\subsection{How weak passwords can cause problems in robotics}
+\section{State of the Art}
 
 \begin{itemize}
-    \item \textbf{Robot vacuum cleaner:} A harmless example to point out is the case of Mister Greg.
-The computer scientist took part in a program where he was gifted a robot vacuum cleaner with an additional camera for orientation.
-What he didn't know was that a picture of him, sitting on the toilet, would somehow find it's way onto Facebook.
-\cite{b10}\\
-          These days, many people already got themselves a robot vacuum cleaner with a camera.
-In most cases it communicates with the docking station or a device of its owner (e.g.: smart phone) remotely.
-Because it's accessed remotely the connections have to be secured with a strong password.
-If not the user could undergo a similar experience as Mister Greg.
-    \item \textbf{Military drones:} High risk or even danger of weak passwords lies in the area of military applications.
-Remote controlled spying and attack drones are already an important part there and will become even more significant in the future.
-The major problem is that these drones have to communicate with some kind of wireless radio signal.
-These signals are very vulnerable for hacking attacks.
-If somebody gains access to enemy robots, they can do a lot of damage, like destroying military bases.
-So even if it's not known for sure, there is a high probability that remote access to those drones will require a password at some point, and it should definitely be a strong and secure one.
-    \item \textbf{In the future:} Boston Dynamics is a company which focuses on developing robots for the world of tomorrow.
-Robots are developing so fast that it is possible that in the near future many people will have an own personal assistant robot.
-\cite{b12}\\
-          In a futuristic world it could become common to own a robotic dog that looks after a user's house when they aren't home or asleep.
-Or a robotic nanny which helps in the house and becomes part of the family.
-If the password to such a robot gets hacked or leaked, it could do all sorts of damage.
-For example spying on the family or robbing their home could get a lot easier, if one would have access to such a household assistant.
+    \item TCP \& UDP (mostly TCP) for low-level, very performant systems
+    \item WebSockets: Implements an additional protocol on top of TCP re-introducing the concept of messages. For Web applications already using HTTP(s)
+    \item SocketIO: Builds on top of WebSocket, allows sending javascript events over the network, adding event listeners/emitters and allow sending arbitrary data.
 \end{itemize}
 
-\section{State of the art}
+\section{Problems with current offer}
 
-Sensitive information or access to hardware must be protected in some way.
-Nowadays this is likely to happen with a username, which is normally easy to guess, and a password, which should be complex and secure.
-Also the data traffic should be encrypted, so nobody can read along and possibly gain access to it.\\
-For example users need to key in a username and password before they can access any network device over SSH.
-This is also the case if a client wants to access the Raspberry Pi inside the Wombat controller remotely.
-Over SSH the data traffic is encrypted and protected with a password.
+This section builds upon the last mentioned protocol SocketIO.
 
-\subsection{The problem with passwords}
+No reassurance of data validity.
+before using incoming data, one ideally has to check that the structure and type of the received data matches what is expected and handle the error if that isn't the case.
+This is tedious to do manually for every single event.
+Additionally since SocketIO allows sending arbitrary data, it is easy to accidentally send data in an incorrect structure, such as accidentally making a typo in an attribute name.
 
-As mentioned above, a password protects sensitive information or access to hardware like a robot.
-That's why it's so important to have a good, secure password.
-However users often are too lazy to create a strong password for every account they own.
-This is understandable, because as often cited study from NordPass shows, the average user has around 100 accounts that require a unique password.
-Remembering a hundred unique passwords or more is really beyond a users capability.
-In a matter of that, users often use the same password for all programs and devices.
-For example if a user owns 150 robots and all of them have the same password a hacker only needs to find out one and already has access to all of his or her robots.
-If a user is using 150 different password, a successful hacker will just be able to control one robot by one hack.\cite{b1} \cite{b2}\\
-Beside laziness as the main cause there are many other reasons why users tend to create weak passphrases such as lack of risk awareness or the fear of forgetting it.\\
-Every year long records get published by different providers revealing the most used passwords. Here are the top ten used passwords of last year (2022): \cite{b3}
-\begin{enumerate}
-    \item 123456
-    \item 123456789
-    \item qwerty
-    \item password
-    \item 12345
-    \item 12345678
-    \item 111111
-    \item 1234567
-    \item 123123
-    \item 1234567890
-\end{enumerate}
-Other common passwords are locations, names, birthdays, etc.
+\section{Solution and Design Overview}
 
-But how quickly can a password really be hacked?
-The following graph shows that it is important to choose long passphrases with different types of characters (such as numbers, upper-case and lower-case letters and symbols).
-The user should always consider that names, common words etc.
-are still easily hackable with a rainbow list attack (brute force attacks with lists of often used passwords).
-Users also shouldn't use personal information as a password.\\
-Figure 1 shows the time it takes to crack a password in a brute force attack depending on its length and the use of different character sets.
+\section{Implementation}
 
-\begin{center}
-    \includegraphics[scale=0.335]{./images/brute_force.png}
-    \captionof{figure}[Graph for secure passwords - The figure is taken from the homepage TechRepublic (https://www.techrepublic.com/article/how-an-8-character-password-could-be-cracked-in-less-than-an-hour)]{How secure is a password depending on its length?}
-\end{center}
+\section{Results and Applications}
 
-Now if a user is not really sure if his or her password is secure, he or she can use a website like PasswordMonster to check how long it would take a hacker to crack his or her password.\cite{b9}
-
-\begin{center}
-    \includegraphics[scale=0.37]{./images/PasswordMonster.png}
-    \captionof{figure}[Screen-shot of PasswordMonster example  - The screenshot is taken from the homepage PasswordMonster (https://www.passwordmonster.com/)]{Example in PasswordMonster}
-\end{center}
-
-\subsection{Current state of the art}
-
-In recent years there was a massive increase in cybercrime. To prevent data-leaks, companies started integrating 2-factor-authentication.\cite{b4}\\
-2-factor-authentication adds an extra layer of security to passwords.
-First the user has to fill in the username and the password like always.
-Then the user has to pass an additional login-method to verify it's really him or her.\\
-The second factor is usually divided into one of the following categories:
-\begin{itemize}
-    \item \textbf{Something only the user knows.}  This might be a personal question the user had to configure beforehand, a personal identification number (PIN), etc.
-    \item \textbf{Something only the user possesses.} This could be a small hardware token or smart-phone.
-    \item \textbf{Something physical unique to the user.} This might be a  biometric pattern like a fingerprint, an iris scan, or a voice print.
-\end{itemize}
-The most common type of the 2-factor-authentication is a smartphone.
-Another good - but less common - option is a hardware key.
-(Something only the user possesses.) A leading company in this market is yubico.
-They are providing a security key named YubiKey for devices with USB-A, USB-C, Lightning and NFC.
-Users that have a YubiKey should be aware that they have to carry it with them at all times and be careful not to lose it.\cite{b11}\\
-Another widely used method is the SMS Text-Message.
-(Something only the user knows.) After entering the password and username the user receives a unique one time passcode (OTP) via text message.
-The user needs to key in the passcode during a limited time (for example one minute).\\
-After both the first factor and the second factor are correct the user gains access.
-
-\subsection{Future state of the art}
-
-In the future passwords will likely be entirely replaced by passkeys because they are easier to use, securer, faster and still work on almost every device.\\
-A passkey is similar to the 2-factor-authentication but without the first factor.
-The second and only factor is the user's device's security method such as a PIN or a biometric sensor (fingerprint, face scan).
-\\
-The passkey is hardware specific, which means the user always needs to carry a second device (that the passkey is installed on).
-If the user wants to use a passkey on the laptop he or she needs to verify the identity with an additional smartphone.
-This eliminates phishing, credential stuffing and other remote attacks.
-\\
-In conclusion when a user is asked to sign-in to an app or website, he or she approves the sign-in with the same biometric or PIN that he or she uses to unlock the device (phone, computer or security key).
-The app or website can use this mechanism instead of the traditional (and insecure) username and password.
-\cite{b5}
-
-\subsection{Solutions to password managing}
-
-If users don't want to or are not able to use passkeys there are different options how they could protect their sensitive information.
-Users can write their passwords with their account information into a booklet.
-Thus, the passwords can be unique and strong.
-It's a really simple idea but also has a lot of drawbacks.
-For example:
-\begin{itemize}
-    \item Users need to always carry the booklet with them. Without it they are not able to access their accounts.
-    \item Users can lose the booklet. Depending on the security restrictions they can't gain access to some of their accounts anymore.
-    \item It's not a really clear way of organizing passwords.
-Once something is written down it cannot change the position in the booklet.
-So users would have to search for each password for indefinite time before they find it.
-\end{itemize}
-The other and better option is a password safe.
-
-\subsection{Password Safe}
-
-\subsubsection{Explanation}
-
-A password safe, also known as password manager, is a piece of sofware that allows saving a list of usernames and passwords for all accounts.
-It has one master password or passphrase that unlocks the safe.
-Now users only have to recall one single master passphrase instead of a huge amount of passwords.
-
-\subsubsection{Examples}
-
-\begin{itemize}
-    \item \textbf{KeepassXC} \\
-        \begin{center}
-            \includegraphics[scale=0.6]{./images/KeePass.png}
-            \captionof{figure}[KeePassXC - The figure is taken from the homepage of the KeePassXC Project (https://keepassxc.org/)]{KeePassXC Logo}
-        \end{center}
-
-        This software is developed for users with extremely high demands of secure personal data management. \\
-        KeePassXC uses Advanced Encryption Standard (AES) encryption algorithm with a 256-bit key to secure the password database.\\
-        The biggest difference to other password safes is that the data (password, account information and additional data such as URLs, attachments and notes) is stored in an offline, encrypted file that can be stored locally.
-This  prevents user's data from getting leaked when a server gets hacked or breached.\\
-        The XC stands for Cross-Plattform. It is possible to use it on Linux, Windows, macOS and Android.\\
-        The program is also customizable. It allows the user to customize literally everything to their needs.\cite{b6}\\
-
-    \item \textbf{1Password} \\
-        \begin{center}
-            \includegraphics[scale=0.06]{./images/1Password-Logo.wine.png}
-            \captionof{figure}[1Password - The figure is taken from the homepage of the 1Password Project (https://1password.com/)]{1Password Logo}
-        \end{center}
-        This password safe is known for its easy to use interface and its high security encryption. \\
-        1Password uses an uncommon encryption known as dual-key encryption.
-If the server gets breached it's impossible for the hacker to decrypt the user's sensitive information because of the two required keys.
-The first key is the user's master-key.
-The second key is a secret key, which is a 128-bit, machine-generated code.
-The secret key is generated on every device the user logs into.
-It will only be saved on the user's devices and will never be saved together with the second PINs.\\
-        Depending on the account type the data is stored differently, but there is only one account type saving the users' data locally.
-In every other option the user's data gets saved in a cloud-based vault.\\
-        The interface is known as user-friendly and generates strong passwords for accounts that are newly created.
-        \cite{b7}
-\end{itemize}
-
-\subsubsection{Benefits and drawbacks of password safe software}
-
-\textbf{Benefits}:
-\begin{itemize}
-    \item \textbf{Security:} Password safes use advanced encryption algorithms to protect the user's passwords.
-This makes it extremely difficult for hackers to steal the passwords and gain access to the user's online accounts.
-    \item \textbf{Convenience:} Password safes store all of the user's passwords in one place, which makes it easy for them to access their accounts without remembering multiple passwords.
-    \item \textbf{Auto-fill feature:} Many password safes have an auto-fill feature that automatically enters the username and the password of the user, saving time and reducing the risk of typing errors.
-    \item \textbf{Multi-device access:} Many password safes allow the user to access their passwords from multiple devices, including smartphones, tablets, and computers.
-    \item \textbf{Password strength:} Password safes often include a password generator feature, which creates strong passwords. This helps to ensure that the passwords are not easily guessable.
-\end{itemize}
-
-\textbf{Drawbacks}:
-\begin{itemize}
-    \item \textbf{Single point of failure:} If a user's password safe is hacked or compromised, all of their passwords are at risk.
-It is important to choose a password safe that has a strong encryption algorithm and to use a strong master password.
-    \item \textbf{Dependency:} Using a password safe can make a user dependent on it for remembering passwords, which may make it difficult to remember passwords if the user is not able to access their password safe for some reason.
-    \item \textbf{Learning curve:} Some password safes have a steep learning curve, and it can take some time to get used to the application.
-    \item \textbf{Cost:} Some password safes charge a fee for their services, which can be a disadvantage if the user is looking for a free solution.
-    \item \textbf{Compatibility:} Some password safes may not be compatible with all websites or devices, which can limit their usefulness.
-\end{itemize}
-
-Sources used to find the benefits and drawbacks: \cite{b8}
-
-\subsubsection{Scientific Studies regarding password managers}
-
-A thoughtful scientific study by Peter Mayer et al.
-shows, that even for highly educated people, which definitely know about the risk of bad or reused passwords, the adoption of password manager software depends largely on how easy it is to use.
-Mayer and his colleagues show, that browser based password managers are more widely used than standalone password saves, mainly because they are easier to use.\cite{b13}
-
-\section{Concept and implementation for best practice}
-
-It can be very overwhelming to start with a secure option to manage passwords. So here is a best practice guide everyone can follow.
-
-\begin{enumerate}
-    \item 
-        First you should check if your password has already been breached.
-There is a website that makes it easy to check it out.
-It's called \textbf{"haveibeenpwned.com"}.
-Haveibeenpwned is open source and secure.
-If someone searches for a breach it doesn't store the data that was typed in (an email address or phone number).
-It only ever retrieves the data from storage then returns the data.
-        \begin{center}
-            \includegraphics[scale=0.29]{./images/haveibeenpwned.png}
-            \captionof{figure}[Have I been pwned - The figure is taken from the homepage of the HaveIBeenPwned? Project (https://haveibeenpwned.com/)]{"Have I been pwned?" Homepage}
-        \end{center}
-        If you have already been pawned, change the present password. If not you can rest assured.
-    \item 
-        Pick a password safe that fits best for your needs. You can choose one of the ones mentioned above or do your own research.
-    \item 
-        Create a good master password or even better a master passphrase. Here is a easy way to to design it:
-        \begin{itemize}
-            \item First think of something which is easy to remember. For example "I love apples and bananas" or "Ich komme aus der Stadt Wien" if you want to use your native language.
-            \item The second step is to transform the sentence in dialect or slang form. For example "I luv appls and bananas" or "I komm aus da Stadt Wien".
-            \item Third, you can change letters into numbers and remove spaces.
-A "s" gets changes into a "5", an "e" gets changed into a "3" and an "o" gets changed into a "0".
-For example "Iluvappl5andbanana5" or "Ik0mmau5da5tadtWi3n".
-            \item The fourth step is to add symbols to your password.
-Change characters that look like a symbol into one such as "a" to "@" or "s" to "?".
-For example "Iluv@ppl5@ndb@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-            \item The fifth step is to make all words start with a capital letter.
-This is only useful if the language you choose does not have a grammar rule for that (It's needed in English but not in German for example).
-        \end{itemize}
-        Finally you would end up with a passphrase like "ILuv@ppl5@ndB@n@n@5" or "Ikomm@u5d@5tadtWi3n".
-This is a really secure passphrase because it compromises numbers, upper case and lower case letters, symbols and more than 8 characters.
-This would make a strong master password that is nearly impossible to crack.
-\end{enumerate}
+\section{Limitations}
 
@@
 
-Passwords remain a widely used and convenient method of authentication, but they have significant weaknesses that make them vulnerable when attacked.
-While there are emerging alternatives to passwords, such as biometric authentication and passkeys, there are also challenges associated with their adoption, such as cost, compatibility, and user experience.\\
-As the cybersecurity landscape continues to evolve, it is important to explore new, innovative and equally trustworthy approaches to authentication and access control, that are both secure and user-friendly.
-
-% Abbildungsverzeichnis
 \listoffigures
@@
 \begin{thebibliography}{00}
+
     \bibitem{b1} Proof for NordPass study, URL:https://tech.co/password-managers/how-many-passwords-average-person
-
-    \bibitem{b2} Information for "The problem with passwords", URL:https://www.nomios.com/news-blog/password-problem/
-
-    \bibitem{b3} List for most used passwords, URL:https://www.passwordmanager.com/most-common-passwords-latest-2022-statistics/
-
-    \bibitem{b4} Information for "Current state of art", URL:\\ https://authy.com/what-is-2fa/
-
-    \bibitem{b5} Information for "Future state of art", URL:\\https://fidoalliance.org/passkeys/\\ https://developer.apple.com/passkeys/\\ https://www.passkeys.io/\\ https://developers.google.com/identity/passkeys?hl=de
-
-    \bibitem{b6} KeepassXC, Open source password manager, URL: https://keepassXC.org
-
-    \bibitem{b7} 1Password, Paid password manager, URL: https://1password.com/de/
-
-    \bibitem{b8} Information for "Benefits and drawbacks of password safe software", URL: \\
-    https://expert.services/blog/managing-your-website/security/password-managers;\\ https://www.orangecountyscu.org/stories/pros-and-cons-of-using-a-password-manager/; \\ https://www.passwordboss.com/pros-and-cons-of-using-a-password-manager/; \\https://www.trustworthy.com/blog/pros-and-cons-password-manager/
-
-    \bibitem{b9} Link to PasswordMonster: https://www.passwordmonster.com/
-
-    \bibitem{b10} Resource for harmless example: https://www.heise.de/hintergrund/Roomba-Wie-private-Fotos-eines-Staubsauger-Roboters-auf-Facebook-landen-koennen-7457283.html
-
-    \bibitem{b11} Information for YubiKey: https://www.yubico.com/der-yubikey/?lang=de
-
-    \bibitem{b12} Resource for "In the future": https://www.bostondynamics.com/
-
-    \bibitem{b13} Why Users (Don’t) Use Password Managers at a Large Educational Institution: https://www.usenix.org/system/files/sec22-mayer.pdf
 

### 2024-03-07 2024_paper/paper/msglink_framework.tex
finished introduction
Co-authored-by: LiloZz <LiloZz@users.noreply.github.com>

@@
 It will provide an insight into why what network communication is needed for and how it is typically realized.
-Additionally it will cover problems and annoyances with currently available communication frameworks for certain applications as well as introduce concepts and a possible solution for the before-mentioned.
+Additionally it will cover problems and annoyances with currently available communication frameworks for user-level applications as well as introduce concepts and a possible solution for the before-mentioned.
 \end{abstract}
@@
 
-When it comes to computer 
-... Why do we need communication? What for?
+For network to function, there needs to be at least two network capable devices.
 
-IP Networks erwähnen
+When it comes to computer science and robotics, sooner or later, multiple computers need to communicate with each other.
+In many cases this communication is achieved over a computer network.
+Nowadays this is typically implemented using the internet protocol (IP).
 
-TCP/UDP Erwähnen und unterschied (Stream vs. Datagram).
+\subsection{IP Networks}
 
-wir behandeln hier nur direkte punkt-zu-punkt verbindungen
+Without going into detail of how the Internet Protocol or any of it's underlying technologies work (out of the scope of this paper), this section provides a quick overview of IP.
+
+In an IP network, every participating communication party (so called "host",such as a computer) is assigned an IP Address (either manually by the Administrator or using automated systems such as DHCP which will not be explained further).
+The IP address is the lowest level of addressing most user-level applications have to deal with.
+It is a 32-Bit number typically expressed by four octets (bytes) in dot notation: \textbf{192.168.1.1}
+
+The IP address is split into two parts: the network address bits (some number of bits on the left of the address) and the host address bits (the rest of the bits on the right side).
+How many of the bits are part of the network and host area is variable and defined by a so-called netmask, which will not be discussed in more detail.
+All host with the same network-address-bits are considered to be on the same network and can therefore communicate with each other.
+Each host is uniquely identified in the network by the host-address-bits.
+
+The Internet Protocol provides the foundations for transmitting data between hosts on this network (or possibly multiple networks): IP network packets.
+These contain Among other details, the sender and receiver addressing information and a block of data to be transmitted between the host.
+The operating system's network stack and other networking hardware then switch and route the packet from the sender to the desired destination host.
+
+When  developing applications however, the programmer almost never has to worry about the working of these underlying systems.
+Instead, they use higher level protocols build on top of IP, which abstract abstract IP packets and allow programmers to more conveniently achieve the desired functionality.
+
+\subsection{TCP and UDP protocols}
+
+There are two main protocols on top of IP for user applications - TCP and UDP.
+
+TCP stands for \textbf{Transmission Control Protocol} and is a connection based protocol, meaning there will be a standing connection between two hosts.
+Before any data can be sent over TCP, a connection has to be established during an initial handshake between both communication parties.
+This is done automatically in the background.
+Once established, a TCP connection provides a pipe-like environment for transmitting a continuous stream of data bytes.
+The stream is of undefined length and data sent by one host is guaranteed to be received in the same order as sent by the other (without missing bytes in between).
+However, since TCP "streams" bytes, it does guarantee that any block of bytes send is received in one piece.
+
+UDP stands for \textbf{User Datagram Protocol} and is .
+The UDP protocol finds the destination device and sends the data inform of a datagram.
+There is nothing set in place to ensure that the datagram reaches the destination.
+There is also no process set in place that tells the receiver end in which order the packets were send.
+A datagram has a size restriction (65.535 bytes) which the packets gets split into.
+\textcolor{red}{(Achtung!!!!!!!!!!!!!!!!!!!!!!
+wenn man mehrere geräte hat verwendet man oft eigentlich UDP da das schneller geht)}
+
+UDP is less reliable than TCP.
+Before choosing a respective protocol one has to decide if they need speed or reliability more.
+A common uses for UDP are videos whereas a common use for TCP are websites.
+
+From here on, this paper will focus on point-to-point communication between two hosts, although many of the concepts described later are also applicable to \textcolor{red}{multicast communication}.
+For this reason, it will focus higher-level protocols and frameworks based on TCP from here-on out.
 
@@
 
-... Why do we need communication frameworks?
+While TCP communication is the basis of many point-to-point networking applications requiring data integrity, it is still rather tedious for a programmer to implement in high-level applications.
+The programmer still has to know how to use the OS APIs to create and manage TCP sockets and since only bytes can be sent over the socket, the programmer needs to manually serialize and deserialize internal program structures and variables to transmit them.
+Besides this, raw TCP is only really suited for continuous data streams, while many applications require event-based bidirectional communication where messages (blocks of data with clear boundaries) are sent back and fourth.
 
-using TCP/UDP directly is tedious for many modern application, ...
+Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks or library.
+A communication framework does not only implement a higher level protocol to support the additional functionality, but also provides programming language functions to integrate the functionality as seamlessly as possible with language features.
 
-Job of cf/library?
+In summary, they may provide the following services to the user:
 
 \begin{itemize}
-    \item abstrahieren
-    \item benutzerfreundlichkeit bzw programmiererfreundlichkeit
-    \item added features for specialized use cases
+    \item Abstraction of low-level socket APIs, exposing the most important functionality and automatically handle the rest in the background (according to common use-cases)
+    \item Automatic management of connection state, handling disconnects and reconnects
+    \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
+    \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network (with the least code possible)
+    \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
+    \item Provide additional, commonly used communication schemes that further simplify common use cases of network communication for the developer such as Remote Procedure Calls (RPCs)
 \end{itemize}

### 2024-03-08 2024_paper/paper/msglink_framework.tex
some changes

@@
 
-UDP stands for \textbf{User Datagram Protocol} and is .
-The UDP protocol finds the destination device and sends the data inform of a datagram.
-There is nothing set in place to ensure that the datagram reaches the destination.
-There is also no process set in place that tells the receiver end in which order the packets were send.
-A datagram has a size restriction (65.535 bytes) which the packets gets split into.
-\textcolor{red}{(Achtung!!!!!!!!!!!!!!!!!!!!!!
+UDP stands for \textbf{User Datagram Protocol}.
+Unlike TCP, UDP does not require a connection to be established.
+The UDP merely sends a block of data (called the user datagram) all at once to a destination host in a single IP packet.
+There is nothing set in place to ensure that the packet reaches the destination.
+Without further protocols, UDP packets do not have any relation to each other, therefore there is no insurance of the order packets are received.
+Since the entire datagram is sent in one IP packet, it's maximum is restricted to the limits imposed by the latter.\textcolor{red}{(65.535 bytes)} \textcolor{red}{(Achtung!!!!!!!!!!!!!!!!!!!!!!
 wenn man mehrere geräte hat verwendet man oft eigentlich UDP da das schneller geht)}
@@
 
-Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks or library.
+Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks or libraries.
 A communication framework does not only implement a higher level protocol to support the additional functionality, but also provides programming language functions to integrate the functionality as seamlessly as possible with language features.
@@
 \begin{itemize}
-    \item Abstraction of low-level socket APIs, exposing the most important functionality and automatically handle the rest in the background (according to common use-cases)
+    \item Abstraction of low-level socket APIs, exposing the most important functionality and automatically handle the rest in the background (according to common use-cases) including but not limited to:
+    \begin{itemize}
         \item Automatic management of connection state, handling disconnects and reconnects
+    \end{itemize}
     \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
@@
     \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
-    \item Provide additional, commonly used communication schemes that further simplify common use cases of network communication for the developer such as Remote Procedure Calls (RPCs)
+    \item Provide additional, commonly used communication schemes such as Remote Procedure Calls (RPCs) that further simplify common use cases of network communication for the developer
 \end{itemize}

### 2024-03-09 2024_paper/paper/msglink_framework.tex
state of the art done

@@
 \usepackage{caption}
+% next 2 needed for code blocks
+\usepackage{listings}
+%\usepackage{etoolbox}  % needed for number detection, but didn't work reliably
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+% colors for code blocks
+\definecolor{code-variable}{HTML}{E06C75}
+\definecolor{code-keyword}{HTML}{A626A4}
+\definecolor{code-string}{HTML}{60A14F}
+\definecolor{code-number}{HTML}{986801}
+\definecolor{code-comment}{HTML}{A0A1A7}
+
+% definition to detect numbers
+
+%\newtoggle{InString}{}% Keep track of if we are within a string
+%\togglefalse{InString}% Assume not initally in string
+%
+%\newcommand*{\ColorIfNotInString}[1]{\iftoggle{InString}{#1}{\color{red}#1}}%
+%\newcommand*{\ProcessQuote}[1]{#1\iftoggle{InString}{\global\togglefalse{InString}}{\global\toggletrue{InString}}}%
+
+% Code blocks
+\lstdefinelanguage{JavaScript}{
+    keywords={typeof, new, true, false, catch, function, return, null, catch, switch, const, let, var, if, in, for, while, do, else, case, break},
+    keywordstyle=\color{blue}\bfseries,
+    ndkeywords={class, export, boolean, throw, implements, import, this},
+    ndkeywordstyle=\color{darkgray}\bfseries,
+    identifierstyle=\color{black},
+    sensitive=false,
+    comment=[l]{//},
+    morecomment=[s]{/*}{*/},
+    morestring=[b]',
+    morestring=[b]" %,
+    %literate=%
+    %    {"}{{{\ProcessQuote{"}}}}1% Disable coloring within double quotes
+    %    {'}{{{\ProcessQuote{'}}}}1% Disable coloring within single quote
+    %    {`}{{{\ProcessQuote{`}}}}1% Disable coloring backtick quote
+    %    {0}{{{\ColorIfNotInString{0}}}}1
+    %    {1}{{{\ColorIfNotInString{1}}}}1
+    %    {2}{{{\ColorIfNotInString{2}}}}1
+    %    {3}{{{\ColorIfNotInString{3}}}}1
+    %    {4}{{{\ColorIfNotInString{4}}}}1
+    %    {5}{{{\ColorIfNotInString{5}}}}1
+    %    {6}{{{\ColorIfNotInString{6}}}}1
+    %    {7}{{{\ColorIfNotInString{7}}}}1
+    %    {8}{{{\ColorIfNotInString{8}}}}1
+    %    {9}{{{\ColorIfNotInString{9}}}}1
+}
+
+\lstnewenvironment{code-js}{
+    \lstset{
+        language=javascript,
+        frame=none,
+        aboveskip=3mm,
+        belowskip=3mm,
+        showstringspaces=false,
+        columns=flexible,
+        basicstyle={\small\ttfamily},
+        numbers=none,
+        numberstyle=\color{code-number},    % line numbers? we want code numbers
+        keywordstyle=\color{code-keyword},
+        commentstyle=\color{code-comment},
+        stringstyle=\color{code-string},
+        breaklines=true,
+        breakatwhitespace=true,
+        tabsize=3
+    }
+}
+{}
+
+\lstnewenvironment{code-cpp}{
+    \lstset{
+        language=C++,
+        frame=none,
+        aboveskip=3mm,
+        belowskip=3mm,
+        showstringspaces=false,
+        columns=flexible,
+        basicstyle={\small\ttfamily},
+        numbers=none,
+        numberstyle=\color{code-number},    % line numbers? we want code numbers
+        keywordstyle=\color{code-keyword},
+        commentstyle=\color{code-comment},
+        stringstyle=\color{code-string},
+        breaklines=true,
+        breakatwhitespace=true,
+        tabsize=3
+    }
+}
+{}
 
@@
 \section{Introduction}
-
-For network to function, there needs to be at least two network capable devices.
 
@@
 
-Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks or libraries.
-A communication framework does not only implement a higher level protocol to support the additional functionality, but also provides programming language functions to integrate the functionality as seamlessly as possible with language features.
+Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks.
+A communication framework does not only implement a higher level protocol to support the additional functionality, but also provides libraries of programming language functions to integrate the functionality as seamlessly as possible with language features.
 
@@
 
-\begin{itemize}
+\begin{enumerate}
     \item Abstraction of low-level socket APIs, exposing the most important functionality and automatically handle the rest in the background (according to common use-cases) including but not limited to:
@@
     \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
-    \item Provide additional, commonly used communication schemes such as Remote Procedure Calls (RPCs) that further simplify common use cases of network communication for the developer
-\end{itemize}
+    \item Provide additional, commonly used communication schemes such as metadata exchange or Remote Procedure Calls (RPCs) that further simplify common use cases of network communication for the developer
+\end{enumerate}
 
@@
 
-\begin{itemize}
-    \item TCP \& UDP (mostly TCP) for low-level, very performant systems
-    \item WebSockets: Implements an additional protocol on top of TCP re-introducing the concept of messages. For Web applications already using HTTP(s)
-    \item SocketIO: Builds on top of WebSocket, allows sending javascript events over the network, adding event listeners/emitters and allow sending arbitrary data.
-\end{itemize}
+There are many different network protocols and accompanying frameworks building on top of the above described, which solve different problems depending on the application requirements.
+Not all of them will be described in this paper.
+
+The following section explains a typical protocol stack used for web apps, since this is one of the original primary use cases for the work introduced in this paper.
+Since web browsers offer only limited to no access to system socket APIs, which already defines a large part of the protocol stack used.
+
+That being said, the following is the protocol stack used as a starting point for the improvements described later:
+
+\subsection{Transmission Control Protocol (TCP)}
+
+\textbf{TCP} as the underlying transportation layer. In an end-user application, this would typically only be used directly if a very high performance custom application protocol is key.
+
+\subsection{Hyper Text Transfer Protocol (HTTP)}
+
+\textbf{HTTP} builds on top of TCP.
+This protocol is used to transmit files and other chunks of data (such as JSON documents) between servers or between servers and a browser.
+One limitation is that HTTP is (mostly) unidirectional communication.
+HTTP works by the client connecting to a server using TCP and then sending an \textbf{HTTP Request}.
+The request consists of at least one line of human readable text denoting the requested resource (like an HTML file) and some optional parameters, called headers.
+The end of the request is identified by a CRLF sequence (an empty line of text).
+Depending on the request details, the server may then answers with some status code, response headers and finally the requested data.
+
+There are libraries for most programming languages, allowing the programmer to send and listen for HTTP request with very little code, varying depending on language.
+
+\begin{code-js}
+// JavaScript
+let resp = await fetch("http://10.5.5.5/");
+\end{code-js}
+
+By automatically handling TCP connection establishment and implementing a way to transfer connection metadata and status codes, HTTP already implements parts of point 1) and 5) of the before mentioned goals for a communication framework.
+
+\subsection{WebSockets}
+
+\textbf{WebSockets} are a bidirectional communication protocol building on top of and somewhat besides HTTP.
+A WebSocket works by first establishing a TCP connection and sending an HTTP request to a server.
+However instead of requesting or transmitting a resource directly, the client requests the server to change the protocol on top of the already open TCP socket to the WebSocket protocol using HTTP protocol upgrade headers.
+Since browsers don't allow web pages and web applications to create TCP sockets directly, this is the only way to create a long-term bidirectional communication channel between the browser and a server application.
+
+The WebSocket protocol however doesn't expose the raw stream socket to the programmer, instead it introduced the concept of messages, hereby implementing point 2) of the goals for a communication framework.
+A message is a self contained chunk of data (typically a string, raw binary is also possible) the transmission of which over the socket is transparent to the user.
+When the programmer sends messages, they guaranteed to arrive at the other communication party in one piece and in order in a message handler.
+
+Like with HTTP, many languages provide simple APIs to create and use WebSocket connections.
+
+\begin{code-js}
+// JavaScript
+// connect
+let socket = new WebSocket("ws://10.5.5.5/");
+// send message
+socket.send("hello");
+// receive message
+socket.onmessage = (e) => {
+    console.log(e.data);
+}
+\end{code-js}
+
+\subsection{Socket.IO}
+
+\textbf{Socket.IO} again builds on top of WebSockets and is the last protocol in the here-described stack.
+Socket.IO works on the principle of sending events over the network.
+An event can be imagined like a message consisting of an \textbf{event name} and some \textbf{event data}.
+Events are identified by their name and there can be multiple different event names in an active connection.
+As soon as the connection between the server and client is established, both sides can \textbf{emit} and \textbf{listen} for events identified by their name.
+Listening works by registering callback (handler) functions, to be called when a specific event is received (i.e.
+emitted by the other communication party).
+When emitting an event, one can also send some additional data with it, which will be passed to the listener functions.
+
+Thanks to support libraries for multiple programming languages, Socket.IO also provides the functionality to serialize and deserialize the data to language-native structures (at least for some languages) and therefore partially satisfies goal 3).
+All this functionality can be achieved with very little code.
+
+\begin{code-js}
+// JavaScript client connect
+const socket = io("ws://10.5.5.5");
+// emit event
+socket.emit("myevent", "some", [47, "data"]);
+// listen for event
+socket.on("event2", (arg1, arg2, ...) => {
+    console.log("event2:", arg1, arg2);
+})
+\end{code-js}
+
+Socket.IO also further implements the abstractions outlined by goal 1) of a communication framework, by implementing automatic connection state management and reconnects.
+Where with WebSockets, the programmer explicitly needs to write code to \textbf{connect to a server},  \textbf{detect disconnects}, and repeat the procedure, with Socket.IO, the programmer is merely required to tell the computer to \textbf{"be connected to a specified server"}.
+The library will then connect, detect failures and reconnect automatically to satisfy this request.
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
problems done

@@
 It will provide an insight into why what network communication is needed for and how it is typically realized.
-Additionally it will cover problems and annoyances with currently available communication frameworks for user-level applications as well as introduce concepts and a possible solution for the before-mentioned.
+Additionally it will cover problems and annoyances with currently available communication frameworks for a specific user-level applications as well as introduce concepts and a possible solution for the before-mentioned.
 \end{abstract}
@@
 From here on, this paper will focus on point-to-point communication between two hosts, although many of the concepts described later are also applicable to \textcolor{red}{multicast communication}.
-For this reason, it will focus higher-level protocols and frameworks based on TCP from here-on out.
+For this reason, it will focus on higher-level protocols and frameworks based on TCP from here-on out.
 
@@
 
-In summary, they may provide the following services to the user:
+The work described in this paper aims to create a communication framework to solve problems for a rather specific set of requirements.:
 
 \begin{enumerate}
-    \item Abstraction of low-level socket APIs, exposing the most important functionality and automatically handle the rest in the background (according to common use-cases) including but not limited to:
+    \item Abstraction of low-level socket APIs, exposing the most important functionality with little code and automatically handle the rest in the background (according to common use-cases) including but not limited to:
     \begin{itemize}
@@
     \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
-    \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network (with the least code possible)
+    \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network with the least code possible
     \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
@@
 
-The following section explains a typical protocol stack used for web apps, since this is one of the original primary use cases for the work introduced in this paper.
-Since web browsers offer only limited to no access to system socket APIs, which already defines a large part of the protocol stack used.
+\textcolor{red}{(possible section "Motivation")}
+The following section explains a typical protocol stack used for web apps, since one of the original primary use-cases of the work described in this paper is the communication between a single-page web-app and a high performance, low level server written in C++.
+This used to implement (among other uses) to implement the touchscreen UI to control FrenchBakery robots for the Botball competition.
+Since web browsers offer only limited to no access to system socket APIs, a large part of the protocol stack used is already defined.
 
-That being said, the following is the protocol stack used as a starting point for the improvements described later:
+That being said, the following is the protocol stack used as a starting point for the improvements described later.
 
@@
 
-This section builds upon the last mentioned protocol SocketIO.
+Although this protocol stack satisfies the most important goals listed, it still has some problems to be solved, namely the following:
 
-No reassurance of data validity.
-before using incoming data, one ideally has to check that the structure and type of the received data matches what is expected and handle the error if that isn't the case.
-This is tedious to do manually for every single event.
-Additionally since SocketIO allows sending arbitrary data, it is easy to accidentally send data in an incorrect structure, such as accidentally making a typo in an attribute name.
+\begin{enumerate}
+    \item The Events system provided by Socket.IO is useful for many applications, sometimes responses to events are required, which is often times named a \textbf{Remote Procedure Call}.
+This needs to be implemented manually for every event:
+    \begin{itemize}
+        \item define and keep track of a separate request and response event
+        \item add an identifying element to request and response data to identify which response belongs to which request and prevent mixing up responses if multiple requests are sent out in short succession
+        \item keep a list of open requests to handle responses once received.
+    \end{itemize}
+    Since the implementation is the same every time and is potentially needed many times in an application, it should be abstracted by the framework.
+    \item While Socket.IO provides a very clean and feature-complete API for some programming languages (like Python and JavaScript), it doesn't to so for others, like C++ which is only supported for clients.
+While there are libraries to create C++ WebSocket servers and clients (like websocket++ which is used in this paper), they require a lot of code (to some part due to the nature of the language) and tedious state management when used directly.
+Additionally, the event system would need to be implemented manually for every event every time which is tedious.
+    \item The data serialization and deserialization of Socket.IO is fully dynamic, meaning that before an event is actually processed, the code cannot possibly know the structure of the event data.
+This presents multiple problems:
+    \begin{itemize}
+        \item Deserialization of event data to language-native data structures is impossible for statically typed languages like C++, which \textcolor{red}{nullifies (bad wording?)} the benefits of static typing.
+It requires the use of dynamic containers like maps to represent objects with strings as attribute names.
+Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type.
+        \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed, it is still necessary or at least helpful to perform the same validation to provide adequate error handling and avoid unexpected behaviors.
+There are also additional systems (TypeScript for JS and type hints for Python) that provide the benefits of static typing even though the underlying language is dynamic.
+An example is rich IntelliSense editor support to avoid typos in attribute names.
+    \end{itemize}
+    Since validation should always be performed (expect when explicitly specifying a type to be unknown), it should be abstracted by the communication framework.
+Instead of the programmer having to write the validation code, the programmer will define the structure of the expected value in a language native format as far as possible (structures in C++, Zod schemas in JS/TS).
+The framework is then responsible for performing the validation, so data reaching the user's code is guaranteed to have the expected structure.
+\end{enumerate}
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
started working on solution and design overview

@@
 
+To solve the before mentioned problems, and implement the last missing goals, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol which has been named \textbf{msglink}.
+
+msglink will build on top of WebSockets, implementing all features of Socket.IO that are relevant to achieve the listed goals.
+In addition it will add direct framework support for more commonly used communication schemes.
+
+This section will cover the core principles of the msglink protocol and how it solves the problems and implements the goals listed above.
+
 \section{Implementation}

### 2024-03-09 2024_paper/paper/msglink_framework.tex
worked on solution

--- 2024-03-09 2024_paper/paper/msglink_framework.tex
@@
 
+To solve the before mentioned problems, and implement the last missing goals, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol with language support libraries which has been named \textbf{msglink}.
+
+msglink will build on top of WebSockets, implementing all features of Socket.IO that are relevant to achieve the listed goals.
+In addition it will add direct framework support for more commonly used communication schemes.
+
+This section will cover the core principles of the msglink framework and how it solves the problems and implements the goals listed above.
+
+\subsection{Simple API}
+
+Similar to the protocols and accompanying libraries listed above (especially Socket.IO), msglink tries to achieve it's core functionality with the least code possible using a declarative API.
+Rather than writing code explicitly stating how to perform the serialization, validation or communication step-by-step (imperative programming), the developer writes code specifying (declaring) what result is expected (declarative programming).
+
+Like Socket.IO, to connect an msglink client to a server, one only has to specify
+
 \section{Implementation}

### 2024-03-09 2024_paper/paper/msglink_framework.tex
Design overview and implementation more or less done

@@
     \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
-    \item Provide additional, commonly used communication schemes such as metadata exchange or Remote Procedure Calls (RPCs) that further simplify common use cases of network communication for the developer
+    \item Provide additional, commonly used communication schemes and primitives such as metadata exchange or Remote Procedure Calls (RPCs).
+These further simplify common practices within network communication for the developer
 \end{enumerate}
@@
 
-To solve the before mentioned problems, and implement the last missing goals, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol with language support libraries which has been named \textbf{msglink}.
+To solve the before mentioned problems, and implement the last missing goals of data validation and more, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol with language support libraries which has been named \textbf{msglink}.
 
@@
 
-This section will cover the core principles of the msglink framework and how it solves the problems and implements the goals listed above.
+\textcolor{red}{remove this paragraph?}
+It should be noted here, that msglink is a point-to-point communication protocol, like WebSocket and Socket.IO.
+Socket.IO provides the ability to send events from a server to multiple connected clients, which is possible to implement using msglink but is not a primary goal and will not be covered here.
+
+These are the core principles of the msglink framework:
+
+\begin{itemize}
+    \item Simple API
+    \item Automatic state management
+    \item Platform independence
+    \item Data validation
+    \item Strict type-safety
+    \item Communication party equality
+    \item Provision of common communication primitives
+    \item Bandwidth efficiency
+\end{itemize}
+
+\section{Implementation}
+
+This section will cover how the the core principles of the msglink framework work and how they solve the problems and implement the goals listed above.
 
@@
 
-Like Socket.IO, to connect an msglink client to a server, one only has to specify
+Like Socket.IO, to connect an msglink client to a server, one only has to specify what server the client should \textbf{be connected to} and the library will automatically handle connection and reconnection in case of connection loss.
 
-\section{Implementation}
+\subsection{Automatic state management}
+
+Like Socket.IO, msglink uses a heartbeat (ping-pong) system to determine when the connection drops.
+This happens internally and requires no code from the user.
+It works by sending a small message back and fourth every few seconds.
+In msglink, the server is responsible for sending this so called "ping" message to the client.
+If the client doesn't respond with the corresponding "pong" message withing a certain time, the connection is dropped.
+Similarly, if the client doesn't receive a ping in a certain amount of time, it drops the connection on it's side and attempts to reconnect automatically.
+
+\subsection{Platform independence}
+
+msglink should be able to be used with different programming languages and on any operating system. All implementations should be able to communicate with each other.
+
+\subsection{Data validation and strict type-safety}
+
+A key feature of msglink is the above discussed integrated data validation, along with other type-safety features (as far as the specific programming language allows).
+
+Even though msglink is the top-level protocol in the newly defined protocol stack, the end-user application's communication itself also happens according to some protocol, defined by the application developer.
+The job of the communication framework (msglink) is, to make the implementation and definition of this protocol as simple as possible.
+In msglink, this user-level protocol is called the \textbf{link}.
+
+Unlike Socket.IO, msglink requires this link to be fully defined in a static manner, meaning the programmer has to explicitly state the existence and associated data structure of every communication primitive (e.g.
+an event, but msglink has others as well) before it can be used for communication.
+In case of an event, that would mean explicitly stating that an event with a given name exists, wether it can be emitted or just be received and providing the schema for the data structure transmitted with the event.
+All of these definitions have to happen statically, so that (for applicable languages) the entire structure of the link (user protocol) is known at compile time.
+This gives the library the ability to automatically perform validation of the communication and serialization/deserialization of the data from/into language-native formats, even for statically typed languages like C++, without the developer having to manually perform any of those steps.
+
+In msglink, both communication parties have to define the link from their perspective of the communication.
+This means for example, that an incoming event on one side has to have a matching outgoing event on the other side.
+When connecting, msglink performs a link-compatibility check, to make sure both parties have defined a compatible protocol.
+If the two links are not compatible, the connection is aborted before trying to sending any user data.
+
+One of the goals for msglink is, to keep the definition of this link entirely native to the language, so no extra build step is required to generate code from the definition.
+This has the slight disadvantage of possibly having to write the same schemas with two different syntaxes.
+
+\subsection{Party equality}
+
+In msglink, both communication parties (client and server) are treated equally.
+Apart from a few internal details, after a msglink connection is established, both parties can perform the exact same actions, in the bounds of what is defined by the link.
+
+\subsection{Common communication primitives}
+
+msglink defines three commonly used communication primitives that a programmer can use to create the user-protocol.
+
+\begin{itemize}
+    \item \textbf{Events} work in the same way as they do with Socket.IO, with the exception of having to define them and their data structure statically.
+    \item \textbf{Remote Function Calls} can be viewed as two events with opposite directions.
+One "call" event and one "result" event.
+RFCs work as the name implies.
+One the called (receiving) side of the communication defines a function taking some input data (statically defined by schema) and returning some result data (also defined by a schema).
+Since a traditional function cannot yield multiple results, there can only be one handler function.
+The other side of the communication can remotely call the function over the network, efficiently using asynchronous abilities of a language if available so the remote function call can be used almost as seamlessly as a local one.
+The framework automatically handles the matching of results to pending calls so data is not mixed up.
+Other framework typically call these Remote Procedure Calls (RPCs), though "function" was chosen for msglink as it clarifies the fact that the call always yields some result data.
+    \item \textbf{Data Sources} can be imagined as functions which can be called with some initial input data, that can follow up with more results over time until canceled.
+This might seem confusing at first, but it is useful for implementing dynamic subscription models.
+This is best explained using an example: \textit{A web dashboard manages 100 servo actuators and needs to display live updates of their positions.
+Always transmitting the position of each results is too much traffic.
+Since only 5 can be displayed on screen at a time, a protocol needs to be implemented to activate or deactivate the updates for each when shown or hidden.} This could be implemented manually using a "sub", "unsub" and "update" event, providing the actuator ID in the "update" event.
+msglink abstracts and standardizes this however.
+In the UI component requiring the data of a specific actuator, a data source listener could be registered with the actuator ID as an input argument.
+This can be imagined as the function being called with the actuator ID as a parameter, and the function then following up with a new result event every time the actuator position updates.
+\end{itemize}
+
+\subsection{Bandwidth efficiency}
+
+At the time of writing, the msglink protocol uses JSON to encode it's internal messages as well as user data.
+This is not very bandwidth efficient, especially when long attribute names are used, which is helpful for development purposes but may be improved in the future.
+
+Never the less, msglink tries to perform the communication as efficiently as possible by only transmitting data over the network that is actually required.
+For example, the msglink framework internally tracks whether an event currently has any listeners and informs the other communication party if there are any.
+If there are no listeners, the event doesn't need to be transmitted over the network, even if one is emitted.
+A similar concept applies to data sources.
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
Addes citations

@@
 
-Like with HTTP, many languages provide simple APIs to create and use WebSocket connections.
+Like with HTTP, many languages provide simple APIs to create and use WebSocket connections. \cite{b3}
 
@@
 The library will then connect, detect failures and reconnect automatically to satisfy this request.
+\cite{b1}
 
@@
 
-    \bibitem{b1} Proof for NordPass study, URL:https://tech.co/password-managers/how-many-passwords-average-person
+    \bibitem{b1} Information for Socket.io, URL:https://socket.io/
+    \bibitem{b2} Information for Zod, URL:https://zod.dev/
+    \bibitem{b3} Information for WebSocket++, URL: https://docs.websocketpp.org/ 
+    https://github.com/zaphoyd/websocketpp/
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
finished base structure

@@
 \author{
-    \IEEEauthorblockN{1\textsuperscript{st} Lilo Zobl}
+    \IEEEauthorblockN{1\textsuperscript{st} Matteo Reiter}
+    \IEEEauthorblockA{
+        \textit{Botball Team FrenchBakery} \\
+        \textit{HTL Anichstraße}\\
+        Innsbruck, Austria \\
+        mareiter@tsn.at
+    }
+    \and
+    \IEEEauthorblockN{2\textsuperscript{nd} Lilo Zobl}
     \IEEEauthorblockA{
@@
         lzobl@tsn.at
-    }
-    \and
-    \IEEEauthorblockN{2\textsuperscript{nd} Matteo Reiter}
-    \IEEEauthorblockA{
-        \textit{Botball Team FrenchBakery} \\
-        \textit{HTL Anichstraße}\\
-        Innsbruck, Austria \\
-        mareiter@tsn.at
     }
@@
 
-One of the goals for msglink is, to keep the definition of this link entirely native to the language, so no extra build step is required to generate code from the definition.
+One of the goals for msglink is to keep the definition of this link entirely native to the respective language, so no extra build step is required to generate code from the definition.
 This has the slight disadvantage of possibly having to write the same schemas with two different syntaxes.
 
-\subsection{Party equality}
+\subsection{Communication party equality}
 
@@
 This can be imagined as the function being called with the actuator ID as a parameter, and the function then following up with a new result event every time the actuator position updates.
+Although the function can be used as an illustration, it is important to understand that the initial "function call" when adding a subscriber with specific input parameters is not intended to execute code on the other parties system which is call-time dependant.
+It is merely a way to "enable" this data source.
 \end{itemize}
@@
 
+Data sources can be optimized further.
+If there are multiple data source "subscribers" with the same request parameters, the request doesn't need to be sent to the server multiple times, and there is no need to send value updates multiple times.
+The framework can detect when there is already an active data source with the same input parameters and simply use it's update messages until there are no subscribers anymore, at which point the source can be disabled.
+
 \section{Results and Applications}
+
+msglink is the result of solving very specific problems for a very specific use case.
+Nevertheless, since it uses widely supported web technologies as it's backbone, it can be easily integrated wherever and expanded with whatever uses these technologies.
+Examples would be a HTTP reverse proxy to add SSL encryption.
+
+Thanks to the rather general API and structure, msglink is useful for potentially many scenarios ranging from server-client communication in UI applications (web and local) to games to data exchange between servers to data exchange between robots in Botball.
 
@@
 
+Since msglink uses WebSockets as an underlying transport layer, a rather inefficient JSON encoding for serialization and is an internally rather complex protocol to manage, speed is not one of it's strengths.
+As mentioned in the protocol stackup, a more lower-level protocol like TCP or UDP directly is required for high-performance applications, trading speed for convenience.
+
+A further limitation is the point-to-point nature of the framework.
+There always needs to be a msglink client connected to a msglink server, although a server can simultaneously have connections to multiple clients and can therefore distribute messages between them.
+At this time, it is not planned to directly implement support for broadcast or multicast communication into msglink, even on the server side, since this was not part of the original goal.
+For applications where many hosts need to communicate with each other directly in a mesh-like topology, there are other protocols better suited for the purpose that don't have the limitation of the browser.
+An example is the DDS which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
+
 \section{Conclusion}
+
+msglink satisfies all the goals defined in this paper for it's specific use case and allows building new, innovative applications quicker without worrying about communication.
+
+msglink is an open source project which can currently be viewed and contributed to as part of the \textbf{melektron/el-std} library on GitHub.
+At the time of writing, msglink is still in the development phase and has not been fully implemented in multiple languages yet.
+The C++ library is the most developed at this time, with TypeScript and Python following very soon.
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
changed authors
state before shortening
Co-authored-by: LiloZz <LiloZz@users.noreply.github.com>
state with just about 5 pages

@@
     This document will focus on an important topic in software development: Network communication.
-It will provide an insight into why what network communication is needed for and how it is typically realized.
-Additionally it will cover problems and annoyances with currently available communication frameworks for a specific user-level applications as well as introduce concepts and a possible solution for the before-mentioned.
+It will provide an insight into what network communication is needed for and how it is typically realized.
+Additionally it will cover problems and annoyances with currently available communication frameworks for specific user-level applications as well as introduce concepts and a possible solution for the before-mentioned problems.
 \end{abstract}
@@
 
-In an IP network, every participating communication party (so called "host",such as a computer) is assigned an IP Address (either manually by the Administrator or using automated systems such as DHCP which will not be explained further).
+In an IP network, every participating communication party (so called "host",such as a computer) is assigned an IP Address.
 The IP address is the lowest level of addressing most user-level applications have to deal with.
@@
 
-The IP address is split into two parts: the network address bits (some number of bits on the left of the address) and the host address bits (the rest of the bits on the right side).
-How many of the bits are part of the network and host area is variable and defined by a so-called netmask, which will not be discussed in more detail.
-All host with the same network-address-bits are considered to be on the same network and can therefore communicate with each other.
-Each host is uniquely identified in the network by the host-address-bits.
-
-The Internet Protocol provides the foundations for transmitting data between hosts on this network (or possibly multiple networks): IP network packets.
-These contain Among other details, the sender and receiver addressing information and a block of data to be transmitted between the host.
+The Internet Protocol provides the foundations for transmitting data between hosts on a network (or possibly multiple networks) in terms of IP network packets.
+These contain among other details, the sender and receiver addressing information and a block of data to be transmitted between the hosts.
 The operating system's network stack and other networking hardware then switch and route the packet from the sender to the desired destination host.
@@
 Once established, a TCP connection provides a pipe-like environment for transmitting a continuous stream of data bytes.
-The stream is of undefined length and data sent by one host is guaranteed to be received in the same order as sent by the other (without missing bytes in between).
-However, since TCP "streams" bytes, it does guarantee that any block of bytes send is received in one piece.
+The data sent is guaranteed to be received in the same order (without missing bytes in between).
+However, since TCP "streams" bytes, it does guarantee that any block of bytes sent at once is received in one piece.
 
@@
 Unlike TCP, UDP does not require a connection to be established.
-The UDP merely sends a block of data (called the user datagram) all at once to a destination host in a single IP packet.
-There is nothing set in place to ensure that the packet reaches the destination.
-Without further protocols, UDP packets do not have any relation to each other, therefore there is no insurance of the order packets are received.
-Since the entire datagram is sent in one IP packet, it's maximum is restricted to the limits imposed by the latter.\textcolor{red}{(65.535 bytes)} \textcolor{red}{(Achtung!!!!!!!!!!!!!!!!!!!!!!
-wenn man mehrere geräte hat verwendet man oft eigentlich UDP da das schneller geht)}
+UDP merely sends a block of data (user datagrams) all at once to a destination host within a single IP packet.
+There is no guarantee that the packet reaches the destination.
 
-UDP is less reliable than TCP.
-Before choosing a respective protocol one has to decide if they need speed or reliability more.
-A common uses for UDP are videos whereas a common use for TCP are websites.
-
-From here on, this paper will focus on point-to-point communication between two hosts, although many of the concepts described later are also applicable to \textcolor{red}{multicast communication}.
-For this reason, it will focus on higher-level protocols and frameworks based on TCP from here-on out.
+From here on, this paper will focus on reliable point-to-point communication between two hosts (although many of the concepts described later are also applicable to multicast communication) and on higher-level protocols and frameworks based on TCP.
 
@@
 
-Providing solutions for these \textcolor{red}{annoyances/problems} is the job of communication frameworks.
+Providing solutions for these annoyances is the job of communication frameworks.
 A communication framework does not only implement a higher level protocol to support the additional functionality, but also provides libraries of programming language functions to integrate the functionality as seamlessly as possible with language features.
 
-The work described in this paper aims to create a communication framework to solve problems for a rather specific set of requirements.:
+The work described in this paper aims to create a communication framework to solve problems for a rather specific set of requirements:
 
 \begin{enumerate}
-    \item Abstraction of low-level socket APIs, exposing the most important functionality with little code and automatically handle the rest in the background (according to common use-cases) including but not limited to:
-    \begin{itemize}
-        \item Automatic management of connection state, handling disconnects and reconnects
-    \end{itemize}
+    \item Abstraction of low-level socket APIs, exposing the most important functionality with little code and automatically handle the rest in the background (according to common use-cases) including but not limited to automatic management of connection state, handling disconnects and reconnects
     \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
@@
 
-There are many different network protocols and accompanying frameworks building on top of the above described, which solve different problems depending on the application requirements.
+There are many different network protocols and accompanying frameworks building on top of TCP, which solve different problems depending on the application requirements.
 Not all of them will be described in this paper.
 
-\textcolor{red}{(possible section "Motivation")}
 The following section explains a typical protocol stack used for web apps, since one of the original primary use-cases of the work described in this paper is the communication between a single-page web-app and a high performance, low level server written in C++.
-This used to implement (among other uses) to implement the touchscreen UI to control FrenchBakery robots for the Botball competition.
-Since web browsers offer only limited to no access to system socket APIs, a large part of the protocol stack used is already defined.
-
-That being said, the following is the protocol stack used as a starting point for the improvements described later.
+This is used to implement (among other uses) the touchscreen UI to control FrenchBakery robots for the Botball competition.
+Since web browsers offer limited to no access to system socket APIs, a large part of the protocol stack is already defined.
 
@@
 
-Like with HTTP, many languages provide simple APIs to create and use WebSocket connections. \cite{b3}
+Like with HTTP, many languages provide simple APIs to create and use WebSocket connections.
 
@@
 \end{code-js}
+\cite{b3}
 
@@
 Additionally, the event system would need to be implemented manually for every event every time which is tedious.
+\cite{b4}
     \item The data serialization and deserialization of Socket.IO is fully dynamic, meaning that before an event is actually processed, the code cannot possibly know the structure of the event data.
@@
 The framework is then responsible for performing the validation, so data reaching the user's code is guaranteed to have the expected structure.
+\cite{b2}
 \end{enumerate}
@@
     \bibitem{b2} Information for Zod, URL:https://zod.dev/
-    \bibitem{b3} Information for WebSocket++, URL: https://docs.websocketpp.org/ 
-    https://github.com/zaphoyd/websocketpp/
+    \bibitem{b3} Websocket standart, URL:  https://www.rfc-editor.org/rfc/rfc6455.html
+    \bibitem{b4} Information for WebSocket++, URL: https://github.com/zaphoyd/websocketpp/
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
rework complete, compacted

@@
 %\usepackage{etoolbox}  % needed for number detection, but didn't work reliably
+% for URLs
+\usepackage{url}
 
@@
 When it comes to computer science and robotics, sooner or later, multiple computers need to communicate with each other.
-In many cases this communication is achieved over a computer network.
+In many cases this is achieved over a computer network.
 Nowadays this is typically implemented using the internet protocol (IP).
-
-\subsection{IP Networks}
-
-Without going into detail of how the Internet Protocol or any of it's underlying technologies work (out of the scope of this paper), this section provides a quick overview of IP.
-
-In an IP network, every participating communication party (so called "host",such as a computer) is assigned an IP Address.
-The IP address is the lowest level of addressing most user-level applications have to deal with.
-It is a 32-Bit number typically expressed by four octets (bytes) in dot notation: \textbf{192.168.1.1}
-
-The Internet Protocol provides the foundations for transmitting data between hosts on a network (or possibly multiple networks) in terms of IP network packets.
-These contain among other details, the sender and receiver addressing information and a block of data to be transmitted between the hosts.
-The operating system's network stack and other networking hardware then switch and route the packet from the sender to the desired destination host.
-
-When  developing applications however, the programmer almost never has to worry about the working of these underlying systems.
-Instead, they use higher level protocols build on top of IP, which abstract abstract IP packets and allow programmers to more conveniently achieve the desired functionality.
 
@@
 This protocol is used to transmit files and other chunks of data (such as JSON documents) between servers or between servers and a browser.
-One limitation is that HTTP is (mostly) unidirectional communication.
+One limitation is that HTTP is (mostly) unidirectional.
 HTTP works by the client connecting to a server using TCP and then sending an \textbf{HTTP Request}.
-The request consists of at least one line of human readable text denoting the requested resource (like an HTML file) and some optional parameters, called headers.
 The end of the request is identified by a CRLF sequence (an empty line of text).
 Depending on the request details, the server may then answers with some status code, response headers and finally the requested data.
+\cite{http-mdn}
 
@@
 \begin{code-js}
-// JavaScript
+// JavaScript example request
 let resp = await fetch("http://10.5.5.5/");
@@
 By automatically handling TCP connection establishment and implementing a way to transfer connection metadata and status codes, HTTP already implements parts of point 1) and 5) of the before mentioned goals for a communication framework.
+\cite{http-mdn}
 
@@
 Since browsers don't allow web pages and web applications to create TCP sockets directly, this is the only way to create a long-term bidirectional communication channel between the browser and a server application.
+\cite{b3}
 
@@
 A message is a self contained chunk of data (typically a string, raw binary is also possible) the transmission of which over the socket is transparent to the user.
-When the programmer sends messages, they guaranteed to arrive at the other communication party in one piece and in order in a message handler.
+When the programmer sends messages, they are guaranteed to arrive at the other communication party in one piece and in order.
+\cite{b3}
 
@@
 \begin{code-js}
-// JavaScript
+// JavaScript WS client example snippets
 // connect
@@
 \end{code-js}
-\cite{b3}
+\cite{ws-client-mdn}
 
@@
 As soon as the connection between the server and client is established, both sides can \textbf{emit} and \textbf{listen} for events identified by their name.
-Listening works by registering callback (handler) functions, to be called when a specific event is received (i.e.
+Listening works by registering callback (handler) functions to be called when a specific event is received (i.e.
 emitted by the other communication party).
 When emitting an event, one can also send some additional data with it, which will be passed to the listener functions.
+\cite{b1}
 
@@
 Socket.IO also further implements the abstractions outlined by goal 1) of a communication framework, by implementing automatic connection state management and reconnects.
-Where with WebSockets, the programmer explicitly needs to write code to \textbf{connect to a server},  \textbf{detect disconnects}, and repeat the procedure, with Socket.IO, the programmer is merely required to tell the computer to \textbf{"be connected to a specified server"}.
+Where with WebSockets the programmer explicitly needs to write code to \textbf{connect to a server},  \textbf{detect disconnects}, and repeat the procedure, with Socket.IO the programmer is merely required to tell the computer to \textbf{"be connected to a specified server"}.
 The library will then connect, detect failures and reconnect automatically to satisfy this request.
-\cite{b1}
 
@@
 \begin{enumerate}
-    \item The Events system provided by Socket.IO is useful for many applications, sometimes responses to events are required, which is often times named a \textbf{Remote Procedure Call}.
-This needs to be implemented manually for every event:
+    \item While the event system provided by Socket.IO is useful for many applications, sometimes responses to events are required.
+This is often times named a \textbf{Remote Procedure Call} and needs to be implemented manually for every event:
     \begin{itemize}
@@
     \end{itemize}
-    Since the implementation is the same every time and is potentially needed many times in an application, it should be abstracted by the framework.
-    \item While Socket.IO provides a very clean and feature-complete API for some programming languages (like Python and JavaScript), it doesn't to so for others, like C++ which is only supported for clients.
+    \item While Socket.IO provides a very clean and feature-complete API for some programming languages (like Python and JavaScript), it doesn't do so for others like C++ which is only supported for clients.
 While there are libraries to create C++ WebSocket servers and clients (like websocket++ which is used in this paper), they require a lot of code (to some part due to the nature of the language) and tedious state management when used directly.
-Additionally, the event system would need to be implemented manually for every event every time which is tedious.
+Additionally, the event system would need to be implemented manually for every event every time, which is tedious.
 \cite{b4}
@@
     \begin{itemize}
-        \item Deserialization of event data to language-native data structures is impossible for statically typed languages like C++, which \textcolor{red}{nullifies (bad wording?)} the benefits of static typing.
+        \item Deserialization of event data to language-native data structures is impossible for statically typed languages like C++, which negates the benefits of static typing.
 It requires the use of dynamic containers like maps to represent objects with strings as attribute names.
-Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type.
-        \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed, it is still necessary or at least helpful to perform the same validation to provide adequate error handling and avoid unexpected behaviors.
-There are also additional systems (TypeScript for JS and type hints for Python) that provide the benefits of static typing even though the underlying language is dynamic.
-An example is rich IntelliSense editor support to avoid typos in attribute names.
+Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type, which is tedious.
+        \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed it is still helpful to perform the same validation in order to provide adequate error handling and avoid unexpected behaviors.
+There are also additional systems (TypeScript for JS and type hints for Python) that provide the benefits of static typing (such as rich IntelliSense) even though the underlying language is dynamic.
     \end{itemize}
-    Since validation should always be performed (expect when explicitly specifying a type to be unknown), it should be abstracted by the communication framework.
-Instead of the programmer having to write the validation code, the programmer will define the structure of the expected value in a language native format as far as possible (structures in C++, Zod schemas in JS/TS).
-The framework is then responsible for performing the validation, so data reaching the user's code is guaranteed to have the expected structure.
-\cite{b2}
 \end{enumerate}
@@
 
-To solve the before mentioned problems, and implement the last missing goals of data validation and more, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol with language support libraries which has been named \textbf{msglink}.
+To solve the aforementioned problems, and implement the last missing goals of data validation and more, the top-level protocol in the stack (Socket.IO) will be replaced by a new protocol with language support libraries which has been named \textbf{msglink}.
 
 msglink will build on top of WebSockets, implementing all features of Socket.IO that are relevant to achieve the listed goals.
-In addition it will add direct framework support for more commonly used communication schemes.
+In addition it will add direct framework support for more commonly used communication schemes/primitives.
 
-\textcolor{red}{remove this paragraph?}
 It should be noted here, that msglink is a point-to-point communication protocol, like WebSocket and Socket.IO.
@@
 Rather than writing code explicitly stating how to perform the serialization, validation or communication step-by-step (imperative programming), the developer writes code specifying (declaring) what result is expected (declarative programming).
+\cite{imperative-vs-declarative}
 
-Like Socket.IO, to connect an msglink client to a server, one only has to specify what server the client should \textbf{be connected to} and the library will automatically handle connection and reconnection in case of connection loss.
+Like Socket.IO, to connect a msglink client to a server, one only has to specify what server the client should \textbf{be connected to} and the library will automatically handle connection and reconnection in case of connection loss.
+
+\begin{code-cpp}
+// example msglink client in C++
+auto client = new msglink::client<my_link>(
+    "ws://10.5.5.5/"
+);
+\end{code-cpp}
 
@@
 
-Like Socket.IO, msglink uses a heartbeat (ping-pong) system to determine when the connection drops.
+Like Socket.IO, msglink uses an additional heartbeat (ping-pong) system on top of the one provided by WebSockets to determine when the connection drops.
 This happens internally and requires no code from the user.
@@
 
-msglink should be able to be used with different programming languages and on any operating system. All implementations should be able to communicate with each other.
+msglink should be usable with different programming languages and on any operating system. All implementations should be able to communicate with each other.
 
@@
 
-A key feature of msglink is the above discussed integrated data validation, along with other type-safety features (as far as the specific programming language allows).
+A key feature of msglink is integrated data validation, along with other type-safety features (as far as the specific programming language allows).
 
 Even though msglink is the top-level protocol in the newly defined protocol stack, the end-user application's communication itself also happens according to some protocol, defined by the application developer.
-The job of the communication framework (msglink) is, to make the implementation and definition of this protocol as simple as possible.
+The job of the communication framework (msglink) is to make the implementation and definition of this protocol as simple as possible.
 In msglink, this user-level protocol is called the \textbf{link}.
@@
 an event, but msglink has others as well) before it can be used for communication.
-In case of an event, that would mean explicitly stating that an event with a given name exists, wether it can be emitted or just be received and providing the schema for the data structure transmitted with the event.
+
+In case of an event, that would mean explicitly stating that an event with a given name exists, whether it can be emitted or received and providing the schema for the data structure transmitted with the event (using language native format as far as possible such as structures in C++ or Zod schemas in JS/TS).
+
 All of these definitions have to happen statically, so that (for applicable languages) the entire structure of the link (user protocol) is known at compile time.
-This gives the library the ability to automatically perform validation of the communication and serialization/deserialization of the data from/into language-native formats, even for statically typed languages like C++, without the developer having to manually perform any of those steps.
+This gives the library the ability to automatically perform validation of the communication and serialization/deserialization of the data from/into language-native formats (even for statically typed languages like C++) without the developer having to manually perform any of those steps.
 
@@
 In msglink, both communication parties (client and server) are treated equally.
-Apart from a few internal details, after a msglink connection is established, both parties can perform the exact same actions, in the bounds of what is defined by the link.
+Apart from a few internal details, after a msglink connection is established, both parties can perform the exact same actions, in the bounds of what is defined by their link.
 
@@
     \item \textbf{Events} work in the same way as they do with Socket.IO, with the exception of having to define them and their data structure statically.
-    \item \textbf{Remote Function Calls} can be viewed as two events with opposite directions.
+    \item \textbf{Remote Function Calls (RFC)} can be viewed as two events with opposite directions.
 One "call" event and one "result" event.
 RFCs work as the name implies.
-One the called (receiving) side of the communication defines a function taking some input data (statically defined by schema) and returning some result data (also defined by a schema).
-Since a traditional function cannot yield multiple results, there can only be one handler function.
-The other side of the communication can remotely call the function over the network, efficiently using asynchronous abilities of a language if available so the remote function call can be used almost as seamlessly as a local one.
+The called (receiving) side of the communication defines a function taking some input data (statically defined by schema) and returning some result data (also defined by a schema).
+Since a traditional function cannot yield multiple results, there can only be one such handler function.
+The other side of the communication can remotely call the function over the network, using asynchronous abilities of a language if available.
+This way, the remote function call can be used almost as seamlessly as a local one.
 The framework automatically handles the matching of results to pending calls so data is not mixed up.
-Other framework typically call these Remote Procedure Calls (RPCs), though "function" was chosen for msglink as it clarifies the fact that the call always yields some result data.
+
+    Other frameworks typically call this scheme "Remote Procedure Call" (RPCs), though "function" was chosen for msglink as it clarifies the fact that the call always yields some result data.
     \item \textbf{Data Sources} can be imagined as functions which can be called with some initial input data, that can follow up with more results over time until canceled.
-This might seem confusing at first, but it is useful for implementing dynamic subscription models.
+This might seem confusing at first, but it is useful for implementing \textbf{dynamic subscription models}.
 This is best explained using an example: \textit{A web dashboard manages 100 servo actuators and needs to display live updates of their positions.
@@
 Since only 5 can be displayed on screen at a time, a protocol needs to be implemented to activate or deactivate the updates for each when shown or hidden.} This could be implemented manually using a "sub", "unsub" and "update" event, providing the actuator ID in the "update" event.
-msglink abstracts and standardizes this however.
+
+    msglink abstracts and standardizes this further.
 In the UI component requiring the data of a specific actuator, a data source listener could be registered with the actuator ID as an input argument.
-This can be imagined as the function being called with the actuator ID as a parameter, and the function then following up with a new result event every time the actuator position updates.
-Although the function can be used as an illustration, it is important to understand that the initial "function call" when adding a subscriber with specific input parameters is not intended to execute code on the other parties system which is call-time dependant.
-It is merely a way to "enable" this data source.
+This can be imagined as the function being called with the actuator ID as a parameter, and the function then following up with a new result event every time this specific actuator's position updates.
+There can also be multiple of those "functions" (possibly with different input values) active at the same time.
+Although the function analogy can be used as an illustration, it is important to understand that the initial "function call" when adding a subscriber with specific input parameters is merely a way to "enable" this data source and not an actual remote function call.
 \end{itemize}
@@
 
-Never the less, msglink tries to perform the communication as efficiently as possible by only transmitting data over the network that is actually required.
+Nevertheless, msglink tries to perform the communication as efficiently as possible by only transmitting data over the network that is actually required.
 For example, the msglink framework internally tracks whether an event currently has any listeners and informs the other communication party if there are any.
@@
 If there are multiple data source "subscribers" with the same request parameters, the request doesn't need to be sent to the server multiple times, and there is no need to send value updates multiple times.
-The framework can detect when there is already an active data source with the same input parameters and simply use it's update messages until there are no subscribers anymore, at which point the source can be disabled.
+The framework can detect when there is already an active data source with the same input parameters and simply use it's update messages until there are no subscribers anymore, at which point the source is disabled.
 
@@
 msglink is the result of solving very specific problems for a very specific use case.
-Nevertheless, since it uses widely supported web technologies as it's backbone, it can be easily integrated wherever and expanded with whatever uses these technologies.
-Examples would be a HTTP reverse proxy to add SSL encryption.
+Nevertheless, since it uses widely supported web technologies as it's backbone, it can be easily integrated and expanded with whatever uses these technologies.
+Examples would be an HTTP reverse proxy to add SSL encryption.
 
@@
 Since msglink uses WebSockets as an underlying transport layer, a rather inefficient JSON encoding for serialization and is an internally rather complex protocol to manage, speed is not one of it's strengths.
-As mentioned in the protocol stackup, a more lower-level protocol like TCP or UDP directly is required for high-performance applications, trading speed for convenience.
+As mentioned in the protocol stackup, directly using a more lower-level protocol like TCP or UDP is required for high-performance applications, trading convenience for speed.
 
@@
 There always needs to be a msglink client connected to a msglink server, although a server can simultaneously have connections to multiple clients and can therefore distribute messages between them.
-At this time, it is not planned to directly implement support for broadcast or multicast communication into msglink, even on the server side, since this was not part of the original goal.
+At this time, it is not planned to directly implement support for broadcast or multicast communication into msglink, even on the server side (which Socket.IO does support), since this was not part of the original goal.
 For applications where many hosts need to communicate with each other directly in a mesh-like topology, there are other protocols better suited for the purpose that don't have the limitation of the browser.
-An example is the DDS which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
+An example is DDS which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
 
@@
 
-msglink is an open source project which can currently be viewed and contributed to as part of the \textbf{melektron/el-std} library on GitHub.
+msglink is an open source project which can currently be viewed and contributed to as part of the \textbf{melektron/el-std} library on GitHub (branch "msglink\_dev").
 At the time of writing, msglink is still in the development phase and has not been fully implemented in multiple languages yet.
 The C++ library is the most developed at this time, with TypeScript and Python following very soon.
-
-\listoffigures
+\cite{el-std-gh}
 
@@
 
-    \bibitem{b1} Information for Socket.io, URL:https://socket.io/
-    \bibitem{b2} Information for Zod, URL:https://zod.dev/
-    \bibitem{b3} Websocket standart, URL:  https://www.rfc-editor.org/rfc/rfc6455.html
-    \bibitem{b4} Information for WebSocket++, URL: https://github.com/zaphoyd/websocketpp/
+    \bibitem{http-mdn} Information for HTTP, URL: \url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Overview}
+    \bibitem{b1} Information for Socket.io, URL: \url{https://socket.io/}
+    \bibitem{b2} Information for Zod, URL: \url{https://zod.dev/}
+    \bibitem{b3} Websocket standart, URL: \url{https://www.rfc-editor.org/rfc/rfc6455.html}
+    \bibitem{ws-client-mdn} Writing a websocket client, URL: \url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API/Writing_WebSocket_client_applications}
+    \bibitem{b4} Information for WebSocket++, URL: \url{https://github.com/zaphoyd/websocketpp/}
+    \bibitem{imperative-vs-declarative} Imperative vs. declarative, URL: \url{https://programiz.pro/resources/imperative-vs-declarative-programming/}
+    \bibitem{el-std-gh} el-std library for C++, URL: \url{https://github.com/melektron/el-std}
 

### 2024-03-09 2024_paper/paper/msglink_framework.tex
added code example, finished paper for now

@@
 
+The following is a compressed code example of how one might define a msglink link in C++. The explanation of the code example is beyond the scope of this paper.
+
+\begin{code-cpp}
+// structure name is event name
+struct my_event 
+    : public msglink::incoming_event {
+    int a;
+    std::string b;
+    EL_MSGLINK_DEFINE_INCOMING_EVENT(
+        my_event, a, b
+    )
+};
+class my_link : public msglink::link {
+    using link::link;
+    EL_MSGLINK_LINK_VERSION(1);
+    msglink::event_sub_hdl_ptr my_sub;
+    void handle_my_event(in_event &_evt) {
+        std::cout << _evt.a << _evt.b;
+    }
+public:
+    virtual void define() noexcept override {
+        my_sub = define_event<my_event>(
+            &my_link::handle_my_event
+        );
+    }
+};
+\end{code-cpp}
+
 Thanks to the rather general API and structure, msglink is useful for potentially many scenarios ranging from server-client communication in UI applications (web and local) to games to data exchange between servers to data exchange between robots in Botball.

### 2024-03-10 2024_paper/paper/msglink_framework.tex
improved title
implemented changes from markus

@@
 \usepackage{url}
+%used to to get the \hl{text} command for highlighting text - used by markus for feedback
+\usepackage{soul}
 
@@
 
-\title{Programmer friendly communication framework}
+\title{A programmer friendly communication framework}
 
@@
 The data sent is guaranteed to be received in the same order (without missing bytes in between).
-However, since TCP "streams" bytes, it does guarantee that any block of bytes sent at once is received in one piece.
+However, since TCP "streams" bytes, it does guarantee that any segment (group of bytes) sent with one system call is received in one piece.
 
@@
 Unlike TCP, UDP does not require a connection to be established.
-UDP merely sends a block of data (user datagrams) all at once to a destination host within a single IP packet.
+UDP merely sends a segment of data (user datagrams) to a destination host within a single IP packet.
 There is no guarantee that the packet reaches the destination.
 
-From here on, this paper will focus on reliable point-to-point communication between two hosts (although many of the concepts described later are also applicable to multicast communication) and on higher-level protocols and frameworks based on TCP.
+From here on, this paper will focus on reliable point-to-point communication between two hosts and on higher-level protocols and frameworks based on TCP (although many of the concepts described later are also applicable to multicast communication).
 
@@
 
-The work described in this paper aims to create a communication framework to solve problems for a rather specific set of requirements:
+The work described in this paper aims to create a communication framework with the following set of goals:
 
@@
     \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network with the least code possible
-    \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
+    \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust, that received data will be in the expected format.
     \item Provide additional, commonly used communication schemes and primitives such as metadata exchange or Remote Procedure Calls (RPCs).
@@
 This is used to implement (among other uses) the touchscreen UI to control FrenchBakery robots for the Botball competition.
-Since web browsers offer limited to no access to system socket APIs, a large part of the protocol stack is already defined.
+Since web browsers offer limited to no access to system socket APIs, a large part of the selected protocol stack is already defined by what browser APIs offer.
 
@@
 
-\textbf{TCP} as the underlying transportation layer. In an end-user application, this would typically only be used directly if a very high performance custom application protocol is key.
+\textbf{TCP} is selected as the underlying transportation layer. In an end-user application, this would typically only be used directly if a very high performance custom application protocol is key.
 
@@
 The end of the request is identified by a CRLF sequence (an empty line of text).
-Depending on the request details, the server may then answers with some status code, response headers and finally the requested data.
+Depending on the request details, the server may then answer with some status code, response headers and finally the requested data.
 \cite{http-mdn}
@@
 
-The WebSocket protocol however doesn't expose the raw stream socket to the programmer, instead it introduced the concept of messages, hereby implementing point 2) of the goals for a communication framework.
+The WebSocket protocol however doesn't expose the raw stream socket to the programmer, instead it introduces the concept of messages, hereby implementing point 2) of the goals for a communication framework.
 A message is a self contained chunk of data (typically a string, raw binary is also possible) the transmission of which over the socket is transparent to the user.
@@
 
-\section{Problems with current offer}
+\section{Problems with current stack}
 
-Although this protocol stack satisfies the most important goals listed, it still has some problems to be solved, namely the following:
+Although this protocol stack satisfies the most important goals listed, it still has the following problems:
 
@@
     \item While the event system provided by Socket.IO is useful for many applications, sometimes responses to events are required.
-This is often times named a \textbf{Remote Procedure Call} and needs to be implemented manually for every event:
+This is typically named a \textbf{Remote Procedure Call} and needs to be implemented manually for every event:
     \begin{itemize}
@@
     \item \textbf{Events} work in the same way as they do with Socket.IO, with the exception of having to define them and their data structure statically.
-    \item \textbf{Remote Function Calls (RFC)} can be viewed as two events with opposite directions.
+    \item \textbf{Remote Function Calls (RFCs)} can be viewed as two events with opposite directions.
 One "call" event and one "result" event.
@@
 
-The following is a compressed code example of how one might define a msglink link in C++. The explanation of the code example is beyond the scope of this paper.
+The following is a compressed code example of how one might define a msglink link in C++, which is significantly shorter than implementing a comparable solution manually using WebSocket++.
+The explanation of the code example is beyond the scope of this paper.
 
@@
 For applications where many hosts need to communicate with each other directly in a mesh-like topology, there are other protocols better suited for the purpose that don't have the limitation of the browser.
-An example is DDS which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
+An example is the \textbf{Data Distribution Serivce (DDS)} which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
 

### 2024-03-10 2024_paper/paper/msglink_framework.tex

@@
 It requires the use of dynamic containers like maps to represent objects with strings as attribute names.
-Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type, which is tedious.
+Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its type and finally convert it to the representative language-native type, which is tedious.
         \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed it is still helpful to perform the same validation in order to provide adequate error handling and avoid unexpected behaviors.
@@
 
-Similar to the protocols and accompanying libraries listed above (especially Socket.IO), msglink tries to achieve it's core functionality with the least code possible using a declarative API.
+Similar to the protocols and accompanying libraries listed above (especially Socket.IO), msglink tries to achieve its core functionality with the least code possible using a declarative API.
 Rather than writing code explicitly stating how to perform the serialization, validation or communication step-by-step (imperative programming), the developer writes code specifying (declaring) what result is expected (declarative programming).
@@
 
-Since msglink uses WebSockets as an underlying transport layer, a rather inefficient JSON encoding for serialization and is an internally rather complex protocol to manage, speed is not one of it's strengths.
+Since msglink uses WebSockets as an underlying transport layer, a rather inefficient JSON encoding for serialization and an internally rather complex protocol to manage, speed is not one of its strengths.
 As mentioned in the protocol stackup, directly using a more lower-level protocol like TCP or UDP is required for high-performance applications, trading convenience for speed.
@@
 
-msglink satisfies all the goals defined in this paper for it's specific use case and allows building new, innovative applications quicker without worrying about communication.
+msglink satisfies all the goals defined in this paper for its specific use case and allows building new, innovative applications quicker without worrying about communication.
 
